# Data model {#sec-erds}

```{r}
library(dplyr)
library(tibble)
library(gt)
library(gtExtras)
```

The following diagrams demonstrate one way in which the data model defined in this manual can be structured. The diagrams only deal with the data themselves - security, user management, change tracking and other helper functionalities are left out for clarity. The model is optimised to highlight the relationships between the various data points and contains some structures that support the controlled vocabulary lists, but is not optimised for performance.

## Helper tables

### Measurement type

Some numeric parameters require a separate specification for types of missing data (e.g. aspect at the site, @sec-pr-aspect). These codes are built in to most categorical code lists in this standard (where it makes sense to include them). For numeric data, a companion text column is required. It is generally expected that it will be populated in a semi-automated manner - for each parameter, it should default to [NR]{.ceg}, and be automatically updated to [NM]{.ceg} if a value is supplied. Users may also update to [ND]{.ceg} as needed.

```{r}
#| label: tbl-missing
#| tbl-cap: "Data record type"

dat_ss_eventt <-
  tribble(~Code, ~Name,
            'NM', 'Parameter was recorded normally',
            'ND', 'Parameter could not be reliably determined or does not apply',
            'NR', 'Parameter was not recorded')

tbl_ss_eventt <- gt(dat_ss_eventt) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red', 
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10))

tbl_ss_eventt
```


## Soil Setting {#sec-erd-setting}

```{mermaid}
%%| fig-cap: "Overview - main tables only"

%%{ init: { 'er': { 'layoutDirection': 'LR' } } }%%
erDiagram
  Site ||--|| Location : "has one"
  Site ||--|| Landform : "is in one"
  Site ||--o{ "Recent events" : "may have some"
  Site ||--|| "Landscape" : "is in one"
  Landscape ||--|{ "Landform" : "contains at least one"
  

```

```{mermaid}
%%| fig-cap: "Overview - including support tables"

%%{ init: { 'er': { 'layoutDirection': 'LR' } } }%%
erDiagram
  site ||--|| location : "has one"
  location }|--|{ elevation_measurement_methods : "code comes from"
  location }|--|{ gnss_types : "code comes from"
  site ||--|| landscape : "has one associated"
  landscape }|--|{ provinces : "code comes from"
  landscape }|--|{ landscapes : "code comes from"
  landscape ||--|{ landform : "contains at least one"
  landform }|--|{ landform_types : "code comes from"
  landform }|--|{ measurement_types : "code comes from"
  landform }|--|{ landform_elements : "code comes from"
  landform }|--|{ landform_element_vmods : "code comes from"
  landform }|--|{ landform_element_hmods : "code comes from"

```

```{mermaid}
%%| fig-cap: "Details"

%%{ init: { 'er': { 'layoutDirection': 'LR' } } }%%
erDiagram
  site ||--|| location : "has one"
  location }|--|{ elevation_measurement_methods : "code comes from"
  location }|--|{ gnss_types : "code comes from"
  site ||--|| landscape : "has one associated"
  landscape }|--|{ provinces : "code comes from"
  landscape }|--|{ landscapes : "code comes from"
  landscape ||--|{ landform : "contains at least one"
  landform }|--|{ landform_types : "code comes from"
  landform }|--|{ measurement_types : "code comes from"
  landform }|--|{ landform_elements : "code comes from"
  landform }|--|{ landform_element_vmods : "code comes from"
  landform }|--|{ landform_element_hmods : "code comes from"
  
  site {
    id           id_site                   "Section 10.2.1"
    txt          usr_author                "Section 10.2.2"
    dat          dt_site_visit             "Section 10.2.4"
    cat          catu_site_purpose         "Section 10.2.5"
    cat          catu_site_selection       "Section 10.2.6"
  }
  
  location {
    id           id_site                   
    geo          point_location            "Section 3.1.1"
    num[-5-3724] amt_elevation_abs_m       "Section 3.1.1"
    num[0-Inf]   amt_error_location_m      "Section 3.1.2"
    num[0-Inf]   amt_error_elevation_m     "Section 3.1.2"
    cat          catu_elevation_method     "Section 3.1.3"
    cat          catu_gnss                 "Section 3.1.4"
    num[-5-3724] amt_elevation_rel_m       "Section 3.2.1"
    txt          txt_notes_triangulation   "Section 3.2.2"
    txt          txt_location_desc         "Section 3.2.3"
    %% not defining code tables for these just now %%
    cat          catu_admin_rc             "Section 3.2.4"
    cat          catu_admin_dc             "Section 3.2.4"
  }
  
  elevation_measurement_methods {
    id           id_elevation_meas
    cat          catu_elevation_meas       "Table 3.1"
    txt          txt_elevation_meas_name   "Table 3.1"
    txt          txt_elevation_meas_desc   "Table 3.1"
  }

  gnss_types {
    id           id_gnss
    cat          catu_gnss_meas            "Table 3.2"
    txt          txt_gnss_meas_name        "Table 3.2"
    txt          txt_gnss_meas_desc        "Table 3.2"
  }  
  
  landscape {
    id id_site
    id id_landscape
    cat          catu_province             "Section 4.1"
    cat          catu_landscape            "Section 4.2.1.4"
    %% landscape relief %%
    num[0-Inf]   amt_lscape_relief_med_m   "Section 4.2.1.1"
    %% landscape slope %%
    num[0-90]    amt_lscape_slope_med_d    "Section 4.2.1.2"
    num[0-Inf]   amt_lscape_slope_min_d    "Section 4.2.1.2"
    num[0-Inf]   amt_lscape_slope_max_d    "Section 4.2.1.2"
    %% not implementing stream channel stuff here %%
  }

  provinces {
    id           id_province
    cat          catu_province             "Table 4.1"
    txt          txt_province_name         "Table 4.1"
    txt          txt_province_desc         "Note 4.1"
  }
  
  landscapes {
    id           id_landscape
    cat          catu_landscape            "Table 4.2"
    txt          txt_landscape_name        "Table 4.2"
    txt          txt_landscape_desc        "Table 4.2"
  }

  landform {
    id           id_site
    id           id_landscape
    id           id_landform
    cat          catu_landform              "Section 4.3.1"
    amt[0-100]   amt_landform_p             "Section 4.1.2.5"
    %% can only be true for one landform per site/landscape: %%
    bool         ind_landform_contains_site  
    %% (other way is separate tables for site-landform and    %%
    %% landscape-landforms)                                   %%
    
    %% detailed optional stuff: %%
    num[0-Inf]   amt_landform_relief_m      "Section 4.3.2.1"
    num[0-90]    amt_landform_slope_med_d   "Section 4.3.2.2"
    num[0-Inf]   amt_landform_slope_min_d   "Section 4.3.2.2"
    num[0-Inf]   amt_landform_slope_max_d   "Section 4.3.2.2"
    num[0-359]   val_landform_aspect_med_d  "Section 4.3.2.3"
    num[0-359]   val_landform_aspect_min_d  "Section 4.3.2.3"
    num[0-359]   val_landform_aspect_max_d  "Section 4.3.2.3"
    cat          catu_landform_aspect_rec   "Table D.1.1"
    cat          cat_landform_element       "Section 4.3.2.4"
    cat          cat_landform_element_vmod  "Section 4.3.2.4"
    cat          cat_landform_element_hmod  "Section 4.3.2.4"
    %% extremely optional %%
    num[0-Inf]   amt_slope_length_m         "Section 4.4.1"
    num[-90-90]  amt_mesoscale_topo_index   "Section 4.4.2"
    num[-90-90]  amt_terrain_shape_index    "Section 4.4.3"
    
  }
  
  landform_types {
    id           id_landform
    cat          catu_landform              "Table 4.3"
    txt          txt_landform_name          "Table 4.3"
    txt          txt_landform_desc          "Table 4.3"
  }

  measurement_types {
    id          id_mtype
    cat         catu_mtype
    txt         txt_mtype_name              "Table D.1.1"
    txt         txt_mtype_desc              "Table D.1.1"
  }
  
  landform_elements {
    id          id_landform_element
    cat         catu_landform_element       "Table 4.4"
    txt         txt_landform_element_name   "Table 4.4"
    txt         txt_landform_element_desc   "Table 4.4"
  }
  
  landform_element_vmods {
    id          id_landform_element_vmod
    cat         catu_landform_element_vmod       "Table 4.5"
    txt         txt_landform_element_vmod_name   "Table 4.5"
    txt         txt_landform_element_vmod_desc   "Table 4.5"
  } 

  landform_element_hmods {
    id          id_landform_element_hmod
    cat         catu_landform_element_hmod       "Table 4.6"
    txt         txt_landform_element_hmod_name   "Table 4.6"
    txt         txt_landform_element_hmod_desc   "Table 4.6"
  } 

```

## Profile data {#sec-erd-profile}

```{mermaid}
%%| fig-cap: "Overview - main tables only"

%%{ init: { 'er': { 'layoutDirection': 'LR' } } }%%
erDiagram
  site ||--o{ disturbances : "may have some"
  

```


```{mermaid}
%%| fig-cap: "Profile overview - including support tables"

%%{ init: { 'er': { 'layoutDirection': 'LR' } } }%%
erDiagram
  site ||--|{ disturbances : "has one or more"
  disturbances }|--|{ disturbance_types : "code comes from"
  disturbances }|--|{ disturbance_ages : "code comes from"
  site }|--|{ site_purposes : "code comes from"
  site }|--|{ site_selection_methods : "code comes from"
  site }|--|{ site_exposure_types : "code comes from"
  site }|--|{ site_early_stop_types : "code comes from"
  

```

```{mermaid}
%%| fig-cap: "Profile - Details"

%%{ init: { 'er': { 'layoutDirection': 'LR' } } }%%
erDiagram
  site ||--|{ disturbances : "has one or more"
  disturbances }|--|{ disturbance_types : "code comes from"
  disturbances }|--|{ disturbance_ages : "code comes from"
  site }|--|{ site_purposes : "code comes from"
  site }|--|{ site_selection_methods : "code comes from"
  site }|--|{ site_exposure_types : "code comes from"
  site }|--|{ site_early_stop_types : "code comes from"

  site {
    id          id_site                  "10.2.1"
    usr         usr_author               "10.2.2"
    dt          dt_site                  "10.2.4"
    cat         catu_site_purpose        "10.2.5"
    cat         catu_site_selection      "10.2.6"
    cat         catu_site_exposure       "10.2.7"
    num[0-1000] amt_site_depth_max_cm    "10.2.8"
    txt         catu_early_stop          "10.2.9"
    %% weather at visit %%
    cat         catu_weather             "Section 6.3.1"
    num[-50-50] val_temperature_c        "Section 6.3.1"
    cat         cato_recent_rain         "Section 6.3.2"
    %% veg %%
    cat        cat_vegetation_luc        "Section 7.1"
    
  }
  
  site_purposes {
    id          id_site_purpose
    cat         catu_site_purpose        "Table 10.1"
    txt         txt_site_purpose         "Table 10.1"
  }
  
  site_selection_methods {
    id          id_site_selection
    cat         catu_site_selection      "Table 10.2"
    txt         txt_site_selection       "Table 10.2"
  }
  
  site_exposure_types {
    id          id_site_exposure
    cat         catu_site_exposure       "Table 10.3"
    txt         txt_site_exposure_name   "Table 10.3"
  }
  
  site_early_stop_types {
    id          id_site_stop
    cat         catu_site_stop           "Table 10.4"
    txt         txt_site_stop            "Table 10.4"
}
  
  disturbances {
    id          id_site
    id          id_disturbance
    txt         catu_disturbance         "Table 10.5"
    txt         catu_disturbance_age     "Table 10.6"
    num[0-Inf]  amt_disturbance_age_days "10.3"
  }

  disturbance_types {
    id          id_disturbance
    cat         catu_disturbance         "Table 10.5"
    txt         txt_disturbance_name     "Table 10.5"
    txt         txt_disturbance_desc     "Table 10.5"
  }
  
  disturbance_ages {
    id          id_disturbance_age
    cat         catu_disturbance_age     "Table 10.6"
    txt         txt_disturbance_age_name "Table 10.6"
    txt         txt_disturbance_age_desc "Table 10.6"
  }
  
```


## Soil Surface {#sec-erd-surface}

```{mermaid}
%%| fig-cap: "Surface - main tables only"

erDiagram
  site ||--|{ surface_cover : "has"
  site ||--o{ erosion : "may have"
  site ||--o{ deposition : "may have"

```

```{mermaid}
%%| fig-cap: "Surface - including support tables"

%%{ init: { 'er': { 'layoutDirection': 'LR' } } }%%
erDiagram
  site }|--|{ slope_methods : "code comes from"
  site }|--|{ measurement_types : "code comes from"
  site }|--|{ aspect_methods : "code comes from"
  site ||--|{ surface_cover : "has"
  surface_cover }|--|{ surface_cover_types : "code comes from"
  site }|--|{ surface_water_types : "code comes from"
  site }|--|{ surface_water_persts : "code comes from"
  site }|--|{ surface_crack_types : "code comes from"
  site }|--|{ surface_crack_shapes : "code comes from"
  site ||--o{ site_erosion : "may have"
  site }|--|{ site_microrelief_nat : "code comes from"
  site }|--|{ site_microrelief_ant : "code comes from"
  site_erosion }|--|{ erosion_types : "code comes from"
  site_erosion }|--|{ erosion_severity : "code comes from"
  site_erosion }|--|{ erosion_activity : "code comes from"
  site ||--o{ site_deposition : "may have"
  site_deposition }|--|{ deposition_types : "code comes from"
  site_deposition }|--|{ deposition_severity : "code comes from"
  site_deposition }|--|{ deposition_activity : "code comes from"
  site }|--|{ water_depths : "code comes from"
  

```

```{mermaid}
%%| fig-cap: "Surface - Details"

%%{ init: { 'er': { 'layoutDirection': 'LR' } } }%%
erDiagram
  site }|--|{ slope_methods : "code comes from"
  site }|--|{ measurement_types : "code comes from"
  site }|--|{ aspect_methods : "code comes from"
  site ||--|{ surface_cover : "has at least one"
  surface_cover }|--|{ surface_cover_types : "code comes from"
  site }|--|{ surface_water_types : "code comes from"
  site }|--|{ surface_water_persts : "code comes from"
  site }|--|{ surface_crack_types : "code comes from"
  site }|--|{ surface_crack_shapes : "code comes from"
  site ||--o{ site_erosion : "may have some"
  site }|--|{ site_microrelief_nat : "code comes from"
  site }|--|{ site_microrelief_ant : "code comes from"
  site_erosion }|--|{ erosion_types : "code comes from"
  site_erosion }|--|{ erosion_severity : "code comes from"
  site_erosion }|--|{ erosion_activity : "code comes from"
  site ||--o{ site_deposition : "may have some"
  site_deposition }|--|{ deposition_types : "code comes from"
  site_deposition }|--|{ deposition_severity : "code comes from"
  site_deposition }|--|{ deposition_activity : "code comes from"
  site }|--|{ water_depths : "code comes from"

  site {
    id          id_site                  "Section 10.2.1"
    num[0-90]   amt_slope_d              "Section 11.1"
    cat         catu_slope_rec           "Table D.1.1"
    cat         catu_slope_method        "Table 11.1"
    num[0-359]  amt_aspect_d             "Section 11.2"
    cat         catu_aspect_rec          "Table D.1.1"
    cat         catu_aspect_method       "Table 11.2"
    cat         catu_surface_water       "Section 11.4"
    cat         cato_surface_water_perst "Section 11.4"
    cat         catu_surface_cracks      "Section 11.5"
    cat         catu_surface_crack_shp   "Section 11.5"
    cat         catu_microrelief_nat     "Section 11.6.1"
    cat         catu_microrelief_ant     "Section 11.6.2"
    amt[0-Inf] amt_depth_water           "Section 12.7"
    cat        catu_depth_water          "Section 12.7"
  }
  
  slope_methods {
    id          id_slope_method
    cat         catu_slope_method        "Table 11.1"
    txt         txt_slope_method_name    "Table 11.1"
    txt         txt_slope_method_desc    "Table 11.1"
  }
  
  aspect_methods {
    id          id_aspect_method
    cat         catu_aspect_method        "Table 11.2"
    txt         txt_aspect_method_name    "Table 11.2"
    txt         txt_aspect_method_desc    "Table 11.2"
  }
  
  measurement_types {
    id          id_mtype
    cat         catu_mtype
    txt         txt_mtype_name           "Table D.1.1"
    txt         txt_mtype_desc           "Table D.1.1"
  }

  surface_cover {
    id          id_site
    cat         catu_surface_cover       "11.3"
    num[0-100]  amt_surface_cover        "11.3"
  }
  
  surface_cover_types {
    id          id_surfcov
    cat         catu_surface_cover       "Table 11.3"
    txt         txt_surface_cover_name   "Table 11.3"
    txt         txt_surface_cover_desc   "Table 11.3"
  }
  
  surface_water_types {
    id          id_surface_water
    cat         catu_surface_water      "Table 11.4"
    txt         txt_surface_water_name  "Table 11.4"
    txt         txt_surface_water_desc  "Table 11.4"
  }

  surface_water_persts {
    id          id_surface_water_perst
    cat         catu_surface_water_perst       "Table 11.5"
    txt         txt_surface_water_perst_name   "Table 11.5"
    txt         txt_surface_water_perst_desc   "Table 11.5"
  }

  surface_crack_types {
    id          id_surface_crack
    cat         catu_surface_crack             "Table 11.6"
    txt         txt_surface_crack_name         "Table 11.6"
    txt         txt_surface_crack_desc         "Table 11.6"
  }
  
  surface_crack_shapes {
    id          id_surface_crack_shp
    cat         catu_surface_crack_shp         "Table 11.7"
    txt         txt_surface_crack_shp_name     "Table 11.7"
    txt         txt_surface_crack_shp_desc     "Table 11.7"
  }
  
  site_microrelief_nat {
    id          id_microrelief_nat
    cat         catu_microrelief_nat         "Table 11.8"
    txt         txt_microrelief_nat_name     "Table 11.8"
    txt         txt_microrelief_nat_desc     "Table 11.8"
  }
  
  site_microrelief_ant {
    id          id_microrelief_ant
    cat         catu_microrelief_ant         "Table 11.9"
    txt         txt_microrelief_ant_name     "Table 11.9"
    txt         txt_microrelief_ant_desc     "Table 11.9"
  }
  
  site_erosion {
    id          id_site
    id          id_erosion                   
    cat         catu_erosion                  
    cat         cato_erosion_severity
    cat         cato_erosion_activity
  }
  
  erosion_types {
    id          id_erosion
    cat         catu_erosion                   "Table 11.10"
    txt         txt_erosion_name               "Table 11.10"
    txt         txt_erosion_desc               "Table 11.10"
  }
  
  erosion_severity {
    id          id_erosion_severity
    cat         catu_erosion_severity          "Table 11.11"
    txt         txt_erosion_severity_name      "Table 11.11"
    txt         txt_erosion_severity_desc      "Table 11.11"
  }  
  
  erosion_activity {
    id          id_erosion_activity
    cat         catu_erosion_activity          "Table 11.12"
    txt         txt_erosion_activity_name      "Table 11.12"
    txt         txt_erosion_activity_desc      "Table 11.12"
  }  
  
 site_deposition {
    id          id_site
    id          id_deposition               
    cat         catu_deposition                   
    cat         cato_deposition_severity
    cat         cato_deposition_activity
  }
  
  deposition_types {
    id          id_deposition
    cat         catu_deposition                   "Table 11.13"
    txt         txt_deposition_name               "Table 11.13"
    txt         txt_deposition_desc               "Table 11.13"
  }  
  
  deposition_severity {
    id          id_deposition_severity
    cat         catu_deposition_severity          "Table 11.14"
    txt         txt_deposition_severity_name      "Table 11.14"
    txt         txt_deposition_severity_desc      "Table 11.14"
  }  
  
  deposition_activity {
    id          id_deposition_activity
    cat         catu_deposition_activity          "Table 11.15"
    txt         txt_deposition_activity_name      "Table 11.15"
    txt         txt_deposition_activity_desc      "Table 11.15"
  }  
  
  water_depths {
    id          id_water_depths
    cat         catu_water_depths           "Table 12.4"
    txt         txt_water_depths_name       "Table 12.4"
    txt         txt_water_depths_desc       "Table 12.4"  
  }
  
```

## Horizon overview data {#sec-erd-horizono}

```{mermaid}
%%| fig-cap: "Horizon - main tables only"

%%{ init: { 'er': { 'layoutDirection': 'LR' } } }%%
erDiagram
  Site ||--|{ Horizon : "has at least one"

```

```{mermaid}
%%| fig-cap: "Horizon - including support tables"

%%{ init: { 'er': { 'layoutDirection': 'LR' } } }%%
erDiagram
  site ||--|{ horizon : "has at least one"
  horizon }|--|{ boundary_shapes : "code comes from"
  horizon }|--|{ boundary_widths : "code comes from"
  horizon }|--|{ soil_moistures : "code comes from"

```

```{mermaid}
%%| fig-cap: "Horizon - Details"

%%{ init: { 'er': { 'layoutDirection': 'LR' } } }%%
erDiagram
  site ||--|{ horizon : "has at least one"
  horizon }|--|{ boundary_shapes : "code comes from"
  horizon }|--|{ boundary_widths : "code comes from"
  horizon }|--|{ soil_moistures : "code comes from"
  
  horizon {
  id         id_site
  id         id_horizon
  amt[0-Inf] amt_upper_depth_cm      "12.2"
  amt[0-Inf] amt_upper_depth_med_cm  "12.2"
  amt[0-Inf] amt_upper_depth_min_cm  "12.2"
  amt[0-Inf] amt_upper_depth_max_cm  "12.2"
  amt[0-Inf] amt_upper_depth_cm      "12.2"
  amt[0-Inf] amt_lower_depth_med_cm  "12.2"
  amt[0-Inf] amt_upper_depth_min_cm  "12.2"
  amt[0-Inf] amt_upper_depth_max_cm  "12.2"
  cat        catu_boundary_shape     "12.4"
  amt[0-Inf] amt_boundary_cm         "12.5"
  cat        catu_boundary_width     "12.5" 
  cat        cato_soil_moisture      "12.6"
  }
  
  boundary_shapes {
    id          id_boundary_shape
    cat         catu_boundary_shape          "Table 12.1"
    txt         txt_boundary_shape_name      "Table 12.1"
    txt         txt_boundary_shape_desc      "Table 12.1"
  }
 
  boundary_widths {
    id          id_boundary_width
    cat         catu_boundary_width          "Table 12.2"
    txt         txt_boundary_width_name      "Table 12.2"
    txt         txt_boundary_width_desc      "Table 12.2"
  } 
  
  soil_moistures {
    id          id_soil_moisture
    cat         catu_soil_moisture          "Table 12.3"
    txt         txt_soil_moisture_name      "Table 12.3"
  }
  
```

## Horizon architecture {#sec-erd-horizona}

```{mermaid}
%%| fig-cap: "Horizon architecture - simple model"

%%{ init: { 'er': { 'layoutDirection': 'LR' } } }%%
erDiagram 
  Horizon ||--|{ "Peds" : "may contain"
  Horizon ||--|{ "Rock Fragments" : "may contain"
  Horizon ||--|{ "Other Fragments" : "may contain"
  Horizon ||--|{ "Artefacts" : "may contain"
  Horizon ||--|{ "Roots" : "may contain"
  Horizon ||--|{ "Voids" : "may contain"
  Horizon ||--|{ "Pores" : "may contain"

```

```{mermaid}
%%| fig-cap: "Horizon architecture - table arrangement"

%%{ init: { 'er': { 'layoutDirection': 'LR' } } }%%
erDiagram 
  horizon ||--|{ horizon_peds : "may have"
  horizon }|--|{ parent_material_origin : "code comes from"
  horizon }|--|{ parent_material_origin_mod : "code comes from"
  horizon }|--|{ pedality_grades : "code comes from"
  horizon }|--|{ ped_sizes_rapid : "code comes from"
  horizon_peds }|--|{ pedality_grades : "code comes from"
  horizon_peds }|--|{ ped_shapes : "code comes from"
  horizon ||--|{ horizon_rock_fragments : "may have"
  horizon_rock_fragments }|--|{ fragment_sizes_simple : "code comes from"
  horizon_rock_fragments }|--|{ rock_fragment_shapes : "code comes from"
  horizon_rock_fragments }|--|{ nzsc_lithologies : "code comes from"
  horizon ||--|{ horizon_other_fragments : "may have"
  horizon_other_fragments }|--|{ fragment_sizes_simple : "code comes from"
  horizon_other_fragments }|--|{ other_fragment_types : "code comes from"
  horizon ||--|{ horizon_artefacts : "may have"
  horizon_artefacts }|--|{ fragment_sizes_simple : "code comes from"
  horizon_artefacts }|--|{ artefact_types: "code comes from"
  horizon ||--|{ horizon_roots : "may have"
  horizon }|--|{ root_positions: "code comes from"
  horizon_roots }|--|{ root_positions : "code comes from"
  horizon_roots }|--|{ root_sizes_simple : "code comes from"
  horizon ||--|{ horizon_voids : "may have"
  horizon_voids }|--|{ void_types : "code comes from"
  horizon_voids }|--|{ void_conns : "code comes from"

```

```{mermaid}
%%| fig-cap: "Horizon architecture - Details"

%%{ init: { 'er': { 'layoutDirection': 'LR' } } }%%
erDiagram
  horizon ||--|{ horizon_peds : "may have"
  horizon }|--|{ parent_material_origin : "code comes from"
  horizon }|--|{ parent_material_origin_mod : "code comes from"
  horizon }|--|{ pedality_grades : "code comes from"
  horizon }|--|{ ped_sizes_rapid : "code comes from"
  horizon_peds }|--|{ pedality_grades : "code comes from"
  horizon_peds }|--|{ ped_shapes : "code comes from"
  horizon ||--|{ horizon_rock_fragments : "may have"
  horizon_rock_fragments }|--|{ fragment_sizes_simple : "code comes from"
  horizon_rock_fragments }|--|{ rock_fragment_shapes : "code comes from"
  horizon_rock_fragments }|--|{ nzsc_lithologies : "code comes from"
  horizon ||--|{ horizon_other_fragments : "may have"
  horizon_other_fragments }|--|{ fragment_sizes_simple : "code comes from"
  horizon_other_fragments }|--|{ other_fragment_types : "code comes from"
  horizon ||--|{ horizon_artefacts : "may have"
  horizon_artefacts }|--|{ fragment_sizes_simple : "code comes from"
  horizon_artefacts }|--|{ artefact_types: "code comes from"
  horizon ||--|{ horizon_roots : "may have"
  horizon }|--|{ root_positions: "code comes from"
  horizon_roots }|--|{ root_positions: "code comes from"
  horizon_roots }|--|{ root_sizes_simple: "code comes from"
  horizon ||--|{ horizon_voids : "may have"
  horizon_voids }|--|{ void_types : "code comes from"
  horizon_voids }|--|{ void_conns : "code comes from"
  
  horizon {
    id         id_site
    id         id_horizon
    %% pmo %%
    cat        catu_parent_material_origin    "13.7.1"
    cat        catu_anthropic_origin_modifier "13.7.1"
    %% pedality rapid (detailed in subtable) %%
    cat        cato_pedality_grade            "14.1.1 (rapid)"
    cat        cato_ped_shape                 "14.1.2/3 (rapid)"
    cat        cato_ped_size                  "14.1.3 (rapid)"
    %% all frags super rapid (record this OR the next three) %%
    num[0-100] amt_total_frags_p               "14.2 (v. rapid)"
    %% rock frags rapid (detailed in subtable) %%
    num[0-100] amt_rockfrags_p                "14.2 (rapid)"
    %% notrock frags rapid %%
    num[0-100] amt_otherfrags_p               "14.2.1.1 (rapid)"
    %% notrock frags rapid %%
    num[0-100] amt_artefacts_p                "14.2.1.2 (rapid)"
    %% roots rapid %%
    num[0-100] amt_total_roots_p               "14.3.1 (rapid)"
    cat        catu_total_roots_pos            "14.3.2 (rapid)"
    %% void rapid %%
    num[0-100] amt_total_voids_p               "14.4 (rapid)"
    %% pores rapid %%
    num[0-10]  amt_total_pores_p               "14.5 (rapid)"
    %% pores detailed (could also spin out to subtable) %%
    num[0-10]  amt_macropores_p                "14.5"
    num[0-10]  amt_micropores_p                "14.5"
                  
  }
 
  parent_material_origin {
    id          id_pmo
    cat         catu_pmo                "Table 24.5"
    txt         txt_pmo_name            "Table 24.5"
    txt         txt_pmo_desc            "Table 24.5"  
  }
  
  parent_material_origin_mod {
    id          id_pmo_mod
    cat         catu_pmo_mod            "Table 13.1"
    txt         txt_pmo_mod_name        "Table 13.1"
    txt         txt_pmo_mod_desc        "Table 13.1"  
  }

  pedality_grades {
    id          id_pedality_grade      
    cat         catu_pedality_grade         "Table 14.1"
    txt         txt_pedality_grade_name     "Table 14.1"
    txt         txt_pedality_grade_desc     "Table 14.1"
  }

  ped_sizes_rapid {
    id          id_ped_size_rapid       "14.1.3"
    cat         catu_ped_size_rapid     "C, F"
    txt         txt_ped_size_rapid_name "Coarse, Fine"
    txt         txt_ped_size_rapid_desc "> 40 mm, <= 40 mm"
  }
  
  horizon_peds {
    id id_site
    id id_horizon
    id id_ped
    cat        cato_pedality_grade            "14.1.1"
    cat        cato_ped_shape                 "14.1.2/3"
    amt[0-Inf] amt_ped_size_med_mm            "14.1.3" 
    amt[0-Inf] amt_ped_size_min_mm            "14.1.3" 
    amt[0-Inf] amt_ped_size_max_mm            "14.1.3"
  }
  

  ped_shapes {
    id          id_ped_shape     
    cat         catu_ped_shape           "Table 14.2/3"
    txt         txt_ped_shape_name       "Table 14.2/3"
    txt         txt_ped_shape_desc       "Table 14.2/3"
  }
  
  horizon_rock_fragments {
    id         id_site
    id         id_horizon
    id         id_rock_fragment
    cat        cato_rock_fragment_size        "Table 14.4"
    cat        catu_rock_fragment_shape       "Table 14.5"
    num[0-100] num_rock_fragment_p         
    cat        catu_rock_fragment_lithology   "Table 22.1"
  }
  
  fragment_sizes_simple {
    id          id_rockfrag_size     
    cat         catu_rockfrag_size            "Table 14.4"
    txt         txt_rockfrag_size_name        "Table 14.4"
    txt         txt_rockfrag_size_desc        "Table 14.4"
  }
  
  rock_fragment_shapes {
    id          id_rockfrag_shape     
    cat         catu_rockfrag_shape           "Table 14.5"
    txt         txt_rockfrag_shape_name       "Table 14.5"
    txt         txt_rockfrag_shape_desc       "Table 14.5"
  }
  
  nzsc_lithologies {
    id          id_nzsc_lithology     
    cat         catu_nzsc_lithology           "Table 24.2"
    txt         txt_nzsc_lithology_name       "Table 24.2"
  }
  
  %% %%
  horizon_other_fragments {
    id         id_site
    id         id_horizon
    id         id_other_fragment
    cat        cato_other_fragment_size       "Table 14.4"
    cat        cato_other_fragment            "Table 14.6"
    num[0-100] num_rock_fragment_p         
  }  
  
  other_fragment_types {
    id          id_other_fragment
    cat         cato_other_fragment           "Table 14.6"
    txt         txt_other_fragment_name       "Table 14.6"
    txt         txt_other_fragment_desc       "Table 14.6"
  }

  %% %%
  horizon_artefacts {
    id         id_site
    id         id_horizon
    id         id_artefact
    cat        cato_artefact_size           "Table 14.4"
    cat        cato_artefact                "Table 14.7"
    num[0-100] num_artefact_p         
  } 
  
  artefact_types {
    id          id_artefact
    cat         catu_artefact                "Table 14.7"
    txt         txt_artefact_type_name       "Table 14.7"
    txt         txt_artefact_type_desc       "Table 14.7"
  }
  
  %% %%
  horizon_roots {
    id         id_site
    id         id_horizon
    id         id_root
    cat        cato_root_size             "Table 14.8"
    cat        cato_root_pos              "Table 14.9"
    num[0-100] num_root_p         
  } 

  root_sizes_simple {
    id          id_root_size     
    cat         catu_root_size            "Table 14.8"
    txt         txt_root_size_name        "Table 14.8"
    txt         txt_root_size_desc        "Table 14.8"
  }
  
  root_positions {
    id          id_root_pos     
    cat         catu_root_pos            "Table 14.9"
    txt         txt_root_pos_name        "Table 14.9"
    txt         txt_root_pos_desc        "Table 14.9"
  }

  %% %%
  horizon_voids {
    id         id_site
    id         id_horizon
    id         id_void
    cat        catu_void                  "Table 14.10"
    cat        catu_void_conn             "Table 14.11"
    num[0-100] num_void_p         
  } 
  
  void_types {
    id          id_void  
    cat         catu_void                 "Table 14.10"
    txt         txt_void_name             "Table 14.10"
    txt         txt_void_desc             "Table 14.10"
  }  

  void_conns {
    id          id_void_conn  
    cat         catu_void_conn            "Table 14.10"
    txt         txt_void_conn_name        "Table 14.10"
    txt         txt_void_conn_desc        "Table 14.10"
  }  
  
```

## Horizon colour {#sec-erd-colour}

```{mermaid}

%%| fig-cap: "Horizon colour - simple model"

%%{ init: { 'er': { 'layoutDirection': 'LR' } } }%%
erDiagram 
  Horizon ||--|{ "Matrix colour" : "has one or more"
  Horizon ||--o{ "Colour patterns" : "may have"

```

```{mermaid}
%%| fig-cap: "Horizon colour - table arrangement"

%%{ init: { 'er': { 'layoutDirection': 'LR' } } }%%
erDiagram
  horizon ||--|{ horizon_matrix_colours : "has one or more"
  horizon ||--|{ horizon_colour_patterns : "has one or more"
  horizon_matrix_colours }|--|{ munsell_colours : "code comes from"
  horizon_colour_patterns }|--|{ munsell_colours : "code comes from"
  horizon_colour_patterns }|--|{ short_colours : "code comes from"
  horizon_colour_patterns }|--|{ pattern_types : "code comes from"
  horizon_colour_patterns }|--|{ pattern_shapes : "code comes from"
  horizon_colour_patterns }|--|{ pattern_contrasts : "code comes from"
  horizon_colour_patterns }|--|{ pattern_edges : "code comes from"
```

```{mermaid}
%%| fig-cap: "Horizon colour - Details"

%%{ init: { 'er': { 'layoutDirection': 'LR' } } }%%
erDiagram
  horizon ||--|{ horizon_matrix_colours : "has one or more"
  horizon ||--|{ horizon_colour_patterns : "has one or more"
  horizon_matrix_colours }|--|{ munsell_colours : "code comes from"
  horizon_colour_patterns }|--|{ munsell_colours : "code comes from"
  horizon_colour_patterns }|--|{ short_colours : "code comes from"
  horizon_colour_patterns }|--|{ pattern_types : "code comes from"
  horizon_colour_patterns }|--|{ pattern_shapes : "code comes from"
  horizon_colour_patterns }|--|{ pattern_contrasts : "code comes from"
  horizon_colour_patterns }|--|{ pattern_edges : "code comes from"
  
  horizon_matrix_colours {
    id          id_site
    id          id_horizon
    id          id_colour_matrix           "Section 15.2"
    %% can also split into hue value chroma %%
    cat         catu_matrix_munsell        "Section 15.2"
    %% no simple colours on the matrix!! %%
    cat         catu_colour_moisture       "Section 15.2"
  }
  
  horizon_colour_patterns {
    id          id_site
    id          id_horizon
    id          id_colour_pattern
    %% either %%
    cat         catu_pattern_munsell    "Section 15.2"
    %% or, %%
    catu        catu_pattern_short_colour   "Section 15.1.1"
    cat         catu_pattern_type           "Section 15.3.1"
    amt[0-100]  amt_pattern_p               "Section 15.3.2"
    cat         catu_pattern_shape          "Section 15.3.3"
    amt[0-Inf]  amt_pattern_size_mm         "Section 15.3.4"
    cat         cato_pattern_contrast       "Section 15.3.5"
    cat         cato_pattern_edge           "Section 15.3.6"
   }
 
  munsell_colours {
    id          id_munsell
    cat         catu_munsell              "Table 15.2"
    %% optional but handy: %%
    txt         txt_colour_hex
  }  
   
  short_colours {
    id          id_colour
    cat         catu_colour               "Table 15.2"
    txt         txt_colour_name           "Table 15.2"
    txt         txt_colour_desc           "Table 15.2"
  } 

  pattern_types {
    id          id_pattern
    cat         catu_pattern              "Table 15.3"
    txt         txt_pattern_name          "Table 15.3"
    txt         txt_pattern_desc          "Table 15.3"
  } 

  pattern_shapes {
    id          id_pattern_shape
    cat         catu_pattern_shape        "Table 15.4"
    txt         txt_pattern_shape_name    "Table 15.4"
    txt         txt_pattern_shape_desc    "Table 15.4"
  }

  pattern_contrasts {
    id          id_pattern_contrast
    cat         catu_pattern_contrast     "Table 15.5"
    txt         txt_pattern_contrast_name "Table 15.5"
    txt         txt_pattern_contrast_desc "Table 15.5"
  }
  
  pattern_edges {
    id          id_pattern_edge
    cat         catu_pattern_edge         "Table 15.6"
    txt         txt_pattern_edge_name     "Table 15.6"
    txt         txt_pattern_edge_desc     "Table 15.6"
  }
```










## Sample data {#sec-erd-samples}

## Interpretation data {#sec-erd-interp}

