---
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse) 
library(gt)
library(gtExtras)
library(ggtern)
library(extrafont)
loadfonts(quiet = TRUE)
```

# Soil texture {#sec-tex}

Soil texture refers to the behaviour of the fine-earth fraction when thoroughly reworked under moist conditions. Texture provides a field estimate of particle size distribution, which is determined in the laboratory. The two measurements are analogous but not identical.

Soil texture is a rapid method that can suggest information about clay mineral type, angularity or roundedness in the sand and silt fractions, organic matter content, and the influence of secondary minerals, as well as the mineral particle size distribution itself. However, the method has precision limits and requires careful user calibration.

Laboratory particle size measurement can involve measurements of particle shape and mineralogy but these methods are labour-intensive and generally confined to a research context. High-throughput analyses confine themselves to assessment of particle size alone, making assumptions about shape and mineralogy (see @loveland2001, for a more detailed discussion.)

Particle size analysis usually involves removal of organic materials and soluble minerals as well as more aggressive aggregate dispersion using shaking and chemical solutions. Following dispersion, measurements of 3-8 specific size fractions are recorded. These data are then usually summarised into a three part sand/silt/clay composition and classified against a texture triangle for efficient communication.

The boundaries of these three size fractions are somewhat unsettled. The silt-clay boundary aligns with some major changes in particle properties; chief among them the emergence of electrostatic behaviour in the clay fraction. The sand/silt boundary can range from 0.02-0.063um. See @tbl-hr-texcompar for the fractions used in NZ and other jurisdictions.

```{r}
#| label: tbl-hr-texcompar
#| tbl-cap: "Size fraction limits: New Zealand and elsewhere" 
#| classes : no-stripe

# can also build out to coarse/med/fine sand and fine clay etc

dat_hr_texcompar <-
  tribble(~Jurisdiction, ~Sand, ~Silt, ~Clay,
          'New Zealand, UK', '< 2.0 mm',  '< 0.06 mm', '< 0.002 mm',
          'Australia  ',     '< 2.0 mm',  '< 0.02 mm', '< 0.002 mm',
          'USA, Canada',     '< 2.0 mm',  '< 0.05 mm', '< 0.002 mm',
          'Global',          '< 2.0 mm', '< 0.063 mm', '< 0.002 mm')

tbl_hr_texcompar <- gt(dat_hr_texcompar) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Jurisdiction ~pct(40))
  
tbl_hr_texcompar


```

## Texture classes {#sec-hr-tex}

Sand/silt/clay compositional data can be displayed on a triangle plot (see @fig-texture-tri), and domains within the plot, or 'texture classes', are labelled for ease of communication. It is important to remember that these classes do not have sharp changes in hand-feel texture at their boundaries. A very large number of soil texture triangles have been defined in different jurisdictions.

```{r}
#| label: fig-texture-tri
#| fig-cap: "New Zealand Soil Texture Triangle"

nztt_dat <- readRDS(file.path('data', 'NZTT.RDS')) 

nztt_labels <- readRDS(file.path('data', 'NZTT_LABELS.RDS'))

nz_tri <-
  ggtern(data = nztt_dat, aes(x = SAND, y = CLAY, z = SILT)) +
  geom_polygon(aes(group = NAME, fill = GROUP),
               colour = 'grey50', alpha = 0.7) +
  geom_text(data = nztt_labels, aes(label = cat_texture_code), 
            colour = 'grey30') +
  scale_fill_manual(values = c('Sandy' = '#ffffb3', 'Silty' = '#bebada',
                               'Loamy' = '#8dd3c7', 'Clayey' = '#fb8072')) +
  theme_showarrows() +
  theme_nomask() +
  labs(x = 'Sand %', y = 'Clay %', z = 'Silt %', fill = 'Texture Group') +
  theme(panel.border = element_rect(colour = 'grey40'),
        panel.background = element_rect(fill = NA),
        tern.panel.grid.ontop = FALSE,
        panel.grid = element_line(colour = 'grey90'),
        text = element_text(family = "Ebrima"),
        legend.key = element_blank(),
        tern.axis.title.show = FALSE,
        legend.position = 'right')

nz_tri
```

```{r}
#| label: tbl-texture-tri
#| tbl-cap: "New Zealand Texture Triangle class names and compositional limits"
#| classes : no-stripe

tbl_hr_textri <- nztt_labels |>
  dplyr::select(matches('cat_'), matches('txt_'), -cat_texture_group) |>
  dplyr::select(cat_texture_code, cat_texture_name, everything()) |>
  gt() |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(cat_texture_code ~pct(10), cat_texture_name ~ pct(20)) |>
  cols_label(cat_texture_name = 'Name', cat_texture_code = 'Code',
             txt_clay_range = 'Clay range',
             txt_silt_range = 'Silt range',
             txt_sand_range = 'Sand range')
  
tbl_hr_textri

```

### Generalised texture classes

Generalised texture classes for rapid assessment and simple communication are supplied below, including codes for gravelly and organic textures.

```{r}
#| label: tbl-hr-simpletex
#| tbl-cap: "Generalised soil texture classes"
#| classes : no-stripe

dat_hr_sandquals <-
  tribble(~Code, ~Name, ~Description,
         'C', 'Clayey', 'Clay content ≥ 35%, texture class C, LC or ZC',
         'Z', 'Silty', 'Clay content < 35%, silt content ≥ 40%; texture class Z, LZ, ZL', 
         'L', 'Loamy', 'Clay content 8 - 35%, silt content < 40%; texture class SL, SC, CL',
         'S', 'Sandy', 'Clay content < 8, silt content, < 40%, texture class S, LS',
         'R', 'Rocky', 'Fine earth fraction < 15%, dominated by rock fragments > 2 mm',
         'P', 'Peaty', 'Organic matter fraction > 30%')

tbl_hr_sandquals <- gt(dat_hr_sandquals) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_hr_sandquals

```

## Recording soil texture

Soil texture is preferably assessed in the field but can also be assessed from representative bagged samples. Follow the procedure in \@fig-texturing. Record one texture code from @tbl-texture-tri per horizon. If desired, add modifiers as discussed below.

[\@fig-texturing goes here]{.diags}

#### Sand modifiers

Hand-texture classes can be accompanied by additional information about the sand fraction. These codes can be appended to a texture class as a prefix, e.g. [(K)LS]{.ceg} for coarse loamy sand.

Note that appreciable coarse sand in a sample leads to underestimates of clay content in hand texturing.

```{r}
#| label: tbl-hr-sandquals
#| tbl-cap: "Sand fraction qualifiers"
#| classes : no-stripe

dat_hr_sandquals <-
  tribble(~Code, ~Name, ~Description,
         'K', 'Coarse', 'Sand is dominantly 2.0 - 0.6 mm',
         'M', 'Medium', 'Sand is dominantly 0.6 - 0.2 mm',
         'F', 'Fine',   'Sand is dominantly 0.2 - 0.06 mm')

tbl_hr_sandquals <- gt(dat_hr_sandquals) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_hr_sandquals

```

#### Organics modifiers

Hand-texture classes can also be modified to express the presence of elevated soil organic matter and its degree of decomposition (where it remains below the cutoff for organic soil materials). These codes can be appended to a texture class as a suffix, e.g. [SL(T)]{.ceg} for slightly organic sandy loam, or [SL(TF)]{.ceg} for slightly organic, fibric sandy loam.

Either of the organic matter abundance qualifiers can be used to signify the 'peaty topsoil' NZSC diagnostic.

```{r}
#| label: tbl-hr-orgmquals
#| tbl-cap: "Organic matter qualifiers - amount"
#| classes : no-stripe

dat_hr_orgmquals <-
  tribble(~Code, ~Name, ~Description,
         'T', 'Slightly organic', '17-30% organic matter',
         'P', 'Moderately organic', '30-50% organic matter')

tbl_hr_orgmquals <- gt(dat_hr_orgmquals) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_hr_orgmquals

```

```{r}
#| label: tbl-hr-orgmdec
#| tbl-cap: "Organic matter qualifiers - decomposition"
#| classes : no-stripe

dat_hr_orgmdec <-
  tribble(~Code, ~Name, ~Description,
         'F', 'Fibric', 'Organic matter weakly decomposed and dominated by visible plant remains',
         'M', 'Mesic', 'Organic matter moderately decomposed, some visible plant structures but mostly amorphous',
         'H', 'Humic', 'Organic matter no longer identifiable, minimal fibre content')

tbl_hr_orgmdec <- gt(dat_hr_orgmdec) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_hr_orgmdec
```

## Particle size estimate

Estimating particle size in the field involves compensating for the influences described above and arriving at estimates of sand and clay content (silt can be determined by difference). Record estimated clay and sand to the nearest whole percent. Optionally, add a separate error estimate (+/- x%) as an expression of confidence.

Example: [(K)LS, 70 ± 5% sand, 3% clay]{.ceg} for a coarse loamy sand.
