# Horizon architecture {#sec-hr-arch}

```{r}
library(dplyr)
library(tibble)
library(gt)
```

Each soil horizon may include any or all of the following major components:

-   Solid fraction
    -   Fine earth
    -   Rock Fragments
    -   Artefacts
-   Biological fraction
    -   Fine organic matter
    -   Coarse organic matter
    -   Living plant roots
-   Liquid fraction
    -   Soil solution
-   Void space fraction
    -   Voids
    -   Pores
        -   Macropores
        -   Micropores

These components can be communicated as compositional percentage by volume, adding to 100%. Volumetric water content is best determined in the laboratory, and so is not described further in this handbook.

[complicated exploding pie chart or similar goes here]{.diags}

Soil horizon architecture involves describing the nature and the physical arrangement of the above components.

## Structure {#sec-hr-stru}

Structure refers to the physical arrangement of the fine-earth fraction (mineral materials \< 2 mm). The type of structure present in a horizon and the extent of its development is determined by the fine-earth texture (@sec-hr-eatxt), mineralogy, chemistry, the action of soil biota, weathering, and land use history. Well-developed structural units are commonly referred to as 'peds' and can be identified by observing any or all of the following:

-   repeating, regular units of similar shape and size
-   small gaps or planes of physical weakness between units
-   outer surfaces that have a distinct appearance when compared to the ped interior.

Peds can range in size from \<1 mm to \>1 m and sometimes 'nest', where larger units break down into smaller units. Given their potential size range, the method of profile exposure determines whether structures can be reliably observed and what maximum unit size will be detectable.

### Structure origin {#sec-str-origin}

Structural development can be controlled by different processes depending on vertical position within the profile and land use history [@anderson2024]. Record the dominant mode of origin for the structural units observed in each horizon using the options in @tbl-hr-form.

```{r}
#| label: tbl-hr-form
#| tbl-cap: "Type of structural development (adapted from @anderson2024)"

dat_hr_form <-
  tribble(~Code, ~Name,  ~Description,
          'P', 'Pedogenic', 'Structural units meet the definition of natural soil peds and occur largely within the solum',
          'G', 'Geogenic', 'Structural units are inherited from parent materials that have undergone limited weathering e.g. bedding planes, and occur largely below the solum',
          'A', 'Anthropogenic', 'Structural units have been induced by human activity, usually within the solum and most commonly in surface horizons',
          'U', 'Unknown', 'Type of structural development could not be determined with confidence')

tbl_hr_form <- gt(dat_hr_form) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_hr_form

```

### Structure development {#sec-hr-stru-pedality}

Degree of structural development (also frequently referred to as, 'grade', 'pedality' or 'aggregation') refers to how distinctive ped appearance is and how much of the soil within a horizon is organised into peds. Developed ped structure can suggest information about mineralogy, biological activity, and soil age. Structural development also provides information about permeability, in concert with texture and firmness.

For rapid assessment, record degree of overall structural development once per horizon using @tbl-hr-pedality. For routine or detailed assessment, record degree of development for each type of structural unit observed (if more than one is seen; see also @sec-hr-stru-nest).

```{r}
#| label: tbl-hr-pedality
#| tbl-cap: "Degree of structural development"

dat_hr_pedality <-
  tribble(~p, ~Code, ~Name,  ~Description,
          'Apedal',  'G', 'Single grain', 'Material is structureless or has <15% identifiable structural units, is uncemented and falls apart into single grains when disturbed',
          'Apedal', 'V', 'Massive', 'Material is structureless or has <15% identifiable structural units, is uncemented and holds together until substantially disturbed, and has no fissures spaced closer than 200 mm',
          'Pedal', 'W', 'Weak', 'Material where 15–25% of the material comprises coherent structural units',
          'Pedal', 'M', 'Moderate', 'Material where 25–75% of the material comprises coherent structural units',
          'Pedal', 'S', 'Strong', 'Material where >75% of the material comprises coherent structural units')

tbl_hr_pedality <- gt(group_by(dat_hr_pedality, p)) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%',
    row_group.as_column = TRUE
  ) |>
  tab_style(style = list(cell_text(weight = 'bold',
                                   v_align = 'middle')),
            locations = cells_row_groups()) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 2)) |>
  cols_width(p ~pct(15), Code ~pct(10), Name ~ pct(20))
  
tbl_hr_pedality

```

### Structure shape {#sec-hr-stru-shp}

Structural units take on a range of distinctive shapes depending on their mineralogy, profile position, organic matter content, and the hydrological and climatic regime. Briefly, four major classes can be recognised, each with a few subtypes:

-   **Blocky:** Blocky peds are roughly cubic (having six faces) and pack relatively closely together, but still have defined interped spaces. Blocky peds are most common in the upper subsoil but can be found in any part of the solum.
    -   **Angular blocky:** Angular blocky peds have sharp corners and flat faces, so can pack closely together.
    -   **Subangular blocky:** Subangular blocky peds have slightly rounded corners and more irregular faces, while still being recognisably rectangular.
-   **Rounded:** Rounded peds have more than six faces and sometimes pack relatively loosely, creating a large proportion of irregular macrovoid space. Rounded peds are most commonly observed in surface horizons.
    -   **Polyhedral:** Polyhedral peds have a rounded shape with more than six faces and angular edges. Their surfaces are irregular, with small divots that align with protrusions in neighbouring peds. As such, they fit together relatively closely, a characteristic sometimes described as 'accommodation to surrounding peds'. Polyhedral peds can be observed throughout the solum.
    -   **Granular:** Granular peds are rounded but lack sharp edges or the divots that allow close packing, and can be very porous. They are most common in near-surface horizons with high organic matter, and can sometimes be held in a net of fine roots.
    -   **Cast:** Cast peds are created by soil fauna, comprising a mix of excreted soil and biological material like mucous or saliva. They are smooth and rounded or ovate and can be observed scattered through surface horizons or down channels extending into subsoil layers. Most are made by earthworms, but beetles and other insects can also create casts.
-   **Platelike:** Platelike peds are wider than they are tall and develop roughly horizontally or in parallel to the soil surface. Layers of platelike structures overlap each other, creating a convoluted path for water or roots moving downwards.
    -   **Platy:** Platy structures are flat-faced and sharp-edged, like angular blocky structures, but are usually very thin. Platy structures can be pedogenic, emerging from freeze-thaw processes, geogenic, reflecting the structure left by rock weathering in place, or anthropogenic, reflecting compaction from heavy machinery use.
    -   **Lenticular:** Lenticular structures have thicker middles and curved surfaces that come to a point along their outer edge. Lenticular structures emerge from shrink-swell clay activity and often have striated faces (@sec-hr-stress).
    -   **Wedgelike:** Wedge structures are related to lenticular structures and appear like a lenticular shape that has been cut in half vertically through its thickest part.
-   **Prismlike:** Prismlike peds are taller than they are wide and have relatively flat faces that fit together closely. They are confined to subsoils and would only be observable near the surface if erosion has recently removed the topsoil.
    -   **Prismatic:** Prismatic peds have a flat top surface with sharp edges.
    -   **Columnar:** Columnar peds have a rounded, or domed, top surface, creating an irregular horizon boundary with the overlying material.

Where distinct structural units can be identified in a horizon (development types [W]{.ceg}, [M]{.ceg}, [S]{.ceg}), record their dominant shape using @tbl-hr-pedshp.

```{r}
#| label: tbl-hr-pedshp
#| tbl-cap: "Structural shapes (adapted from @milne1995, @iussworkinggroupwrb2022, and @ncst2024)"

dat_hr_pedshp <-
  tribble(~Group,     ~Code, ~Name,              ~Diagram,
          'Blocky',    'AB', 'Angular blocky',       '*to be added*',
          'Blocky',    'SB', 'Subangular blocky',    '*to be added*',
          'Rounded',   'PH', 'Polyhedral',           '*to be added*',
          'Rounded',   'GR', 'Granular',             '*to be added*',
          'Rounded',   'CA', 'Cast',                 '*to be added*',
          'Platelike', 'PL', 'Platy',                '*to be added*',
          'Platelike', 'LE', 'Lenticular',           '*to be added*',
          'Platelike', 'WE', 'Wedgelike',            '*to be added*',
          'Prismlike', 'PR', 'Prismatic',            '*to be added*',
          'Prismlike', 'CO', 'Columnar',             '*to be added*') 

tbl_hr_pedshp <- gt(group_by(dat_hr_pedshp, Group)) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%',
    row_group.as_column = TRUE
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 2)) |>
  tab_style(style = list(cell_text(weight = 'bold',
                                   align  = 'center',
                                   v_align = 'middle')),
            locations = cells_row_groups()) |>
  fmt_markdown(columns = 4) |>
  cols_width(Group ~pct(20), Code ~pct(10), Name ~ pct(30))
  
tbl_hr_pedshp
```

::: {#nte-nut .callout-note collapse="true" title="Older terms for structural units" appearance="minimal"}
Some structural terms that don't seem to fit into @tbl-hr-pedshp can be found in older soils literature. These terms should no longer be used because they conflate size and shape in a way that modern schemas seek to avoid.

-   *Crumb:* a very fine structure seen in many silty to loamy horizons. The appearance is as of breadcrumbs, with 0.5–2 mm rounded peds loosely to firmly packed, depending on depth and drainage. Larger weak subangular blocky structures in well-drained allophanic soils almost always break down into moderately or strongly developed crumb structure. Crumb should be described as type [P]{.ceg} with [GR]{.ceg} shape.
-   *Nut:* a larger version of crumb, usually 5–20 mm in diameter, often encased in a network of grass roots. Nut structures are more common in topsoils but can also occur in well-drained loamy subsoils. Nut should be described as type [P]{.ceg} with [GR]{.ceg} or [PR]{.ceg} shape.
-   *Clod:* a tillage-induced structure \> 100 mm in length. Clods should be described as type [A]{.ceg} structures with [SB]{.ceg} or [PH]{.ceg} shape.
-   *Fragment:* a tillage-induced structure \< 100 mm in length. Fragments should be described as type [A]{.ceg} structures with [SB]{.ceg} or [PH]{.ceg} shape.

Ped strength (@sec-cn-stra) can assist in identifying tillage-induced structures, as they can be compacted and firmer than pedogenic structural units.
:::

### Structure size {#sec-hr-stru-size}

Record the median length of the structural units observed, along their shortest axis. Record in millimetres with a range estimate, e.g. [25 ± 2 mm]{.ceg}. Options for classifying the recorded measurements may be found in @sec-in-solsz.

For rapid assessment, rather than recording a numeric value, the dominant size may be noted as [C]{.ceg} for Coarse when over 20 mm in size, and [F]{.ceg} for Fine when \<20 mm.

#### Associated structures {#sec-hr-stru-assc}

Many horizons only have one major structure, but multiple structural units may also coexist. For instance, casts may be interspersed among polyhedral, granular or blocky peds in a surface horizon, or a saprolith layer (@sec-pr-wdeg) might have remnant geogenic bedding among more heavily weathered, apedal-massive material. In such cases, record the structures separately. Where the horizon is partly apedal, estimate the volume of apedal material.

Example: [G M PL 20 ± 3]{.ceg} with [V 35%]{.ceg}

#### Nested structures {#sec-hr-stru-nest}

Well-developed peds may be nested, with large structures breaking down to smaller ones that may also be a different type and shape. For highly detailed descriptions, multiple sets of nesting peds may be described in terms of their origin, development, shape and size.

For example, [P S PR 80 ± 10]{.ceg} breaking to [P M AB 20 ± 2]{.ceg}.

### Method: Assessing structure in the field

-   To assess the structure in an [E]{.ceg} or [P2]{.ceg} type profile (@tbl-pr-gear), insert a spade horizontally near the base of the horizon and push down on the handle, levering out a section of soil. Drop the spade blade with the soil, approximately 0.3 m onto a hard surface, e.g., the blade of another spade. Remove major roots that bind the mass together and again note the attributes of the newly produced solids. Note that this procedure will not necessarily disaggregate very moist or wet soils.
-   Structure may also be observed *in situ* by using a small hand tool to gently pick back the profile face. See @sec-dig. Structure of hard layers may need to be assessed in place, or by careful removal of individual units, and will usually require an [E]{.ceg} or [P2]{.ceg} type profile.
-   For smaller exposures ([P1]{.ceg}, [C1]{.ceg}, [C2]{.ceg}), remove a handful of soil from the extracted horizon and carefully use your fingers to disaggregate the soil mass to segregate all peds. Do not destroy peds to create fragments, or disaggregate existing apedal material. When coring, also use a spade to extract a block of adjacent surface material so that at least the topmost horizon can be assessed in detail.
-   Assessing structure is generally impossible for [A]{.ceg} augered profiles, as the extraction method destroys it. The surface horizon should still be extracted with a spade and assessed if possible.

Direct measurement of peds will help estimate size. Weighing, or filling volumetric containers, may help estimate abundance.

::: {#nte-stro .callout-note collapse="true" title="Structure in organic horizons" appearance="minimal"}
Organic soil materials cannot be described in terms of the structural parameters in this section. Their architecture is flexible and determined largely by water content, species of origin and degree of decomposition, and are effectively all apedal-massive. The physical nature of organic soil materials is otherwise largely inferred from their textural characteristics (@sec-hr-ortxt). However, drained peats and peaty mineral soils may have some recognisable structural development.
:::

## Fragments {#sec-frag}

Fragments are mineral or non‑mineral components \>2 mm within a soil horizon. They provide information about soil development, particularly mode of emplacement (@sec-pr-orig). They also reduce the soil’s water‑holding capacity and, when large and abundant enough, can restrict land use and management practices.

For very rapid assessment, estimate total fragment abundance (rocks and non-rock materials as well as any artefacts) as an integer percentage of the whole horizon volume. Where fragments are absent, always record [0%]{.ceg} as confirmation.

### Rock fragments {#sec-rockfrag}

Rock fragments are soil mineral particles \>2 mm. For rapid assessment, use the texture class code modifiers provided for rock fragments in @sec-hr-rftex, and confirm absence by recording an abundance of [0%]{.ceg}.

Example: [SL(G)]{.ceg} for a sandy loam soil with 5–35% gravels (size range 2–20 mm).

For routine assessment, append an estimate of rock fragment abundance as integer percentage of the whole horizon volume, as well as dominant lithology class using @tbl-ss-pm and shape using @tbl-hr-rockfrshp.

Example: [SL(G) 15% Hs R]{.ceg} for a sandy loam soil with 15% rounded greywacke gravels.

For detailed assessment, one may optionally continue to add detail by recording volumetric abundance, lithology and shape for each of the three rock fragment size classes given in @tbl-hr-rocksz. The three additional gravel size subdivisions in @tbl-hr-gvquals may also be used. Picking and sieving from a known volume of soil will likely be required for sufficient accuracy when making detailed assessments.

```{r}
#| label: tbl-hr-rockfrshp
#| tbl-cap: "Coarse fragment shape classes"

dat_hr_rockfrshp <-
  tribble(~Code, ~Name, ~Description,
         'R', 'Rounded',   'Rock fragments considerably smoothed by transport processes',
         'A', 'Angular',   'Rock fragments with little smoothing by transport processes',
         'M',   'Mixed',   'Both rounded and angular fragments mixed together') 

tbl_hr_rockfrshp <- gt(dat_hr_rockfrshp) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_hr_rockfrshp
```

### Natural fragments {#sec-hr-nonrock}

Biologically-derived coarse materials can alter the soil's water holding capacity and other parameters just as rock fragments can, although in general they can themselves hold more water. They will also break down faster.

Where present in significant amounts (\> 5% by volume), record each type present and estimate volume and median fragment size in centimetres.

```{r}
#| label: tbl-hr-nonrock
#| tbl-cap: "Non-rock fragment types"

dat_hr_nonrock <-
  tribble(~Code, ~Name, ~Description,
         'Sh', 'Shell',    'Seashells or chitinous material',
         'Bo', 'Bone',     'Skeletal fragments',
         'Ti', 'Timber',   'Dead wood from trees and shrubs, coarse roots',
         'Ch', 'Charcoal', 'Burnt organic materials',
         'Gu', 'Gum',      'Kauri gum')

tbl_hr_nonrock <- gt(dat_hr_nonrock) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_hr_nonrock

```

### Artefacts {#sec-hr-artef}

Artefacts can both function like rock fragments and signify strong human influence on the soil profile.

Where present in significant amounts, for rapid assessment simply record total volumetric percentage of artefacts per horizon. Confirm absence as [0%]{.ceg}. For routine assessment, record each type present using @tbl-hr-artefacts and estimate percentage volume and median fragment size in centimetres.

Example: [Ra 5% 6-10 cm]{.ceg} for a horizon containing some chert flake tools.

For detailed assessment, for each fragment type, append any or all of the origin (@tbl-hr-aorg), fixation (@tbl-hr-afix), shape (@tbl-hr-ashp), and degradation (@tbl-hr-adeg) codes supplied below. Colour may also be described using common terms. Additional free-text notes and photos may also be useful.

Example: [Pl 10% 2-3 cm A SF F D]{.ceg} for a horizon containing some broken down remnants of horticultural film.

Artefact fragments smaller than 2 mm and non-solid soil contaminants are beyond the scope of this handbook.

```{r}
#| label: tbl-hr-artefacts
#| tbl-cap: "Artefact types"

dat_hr_artefs <-
  tribble(~Code, ~Name, ~Description,
         'As', 'Asphalt',  'Asphalt and other bitumen-containing materials',
         'Ba', 'Bone - altered', 'Human-altered bone fragments',
         'Br', 'Brick',    'Fired clay building materials, usually unglazed',
         'Cr', 'Ceramic',  'Fired clay objects, usually glazed',
         'Ce', 'Concrete', 'Concrete materials, cements',
         'Gl', 'Glass',    'Fragments from windows, containers, etc',
         'Pl', 'Plastics', 'Artificial polymers (non-textile)',
         'Ra', 'Rock - altered', 'Human-altered rock fragments',
         'Sa', 'Shell - altered', 'Human-altered shells',
         'Ta', 'Timber - altered', 'Human-altered timber',
         'Tn', 'Textiles - natural', 'Fabrics and carpets made from e.g. wool',
         'Tx', 'Textiles - artificial', 'Fabrics derived from or dominated by plastics',
         'Mt', 'Metals', 'Refined ore products',
         'Ms', 'Mine Spoil', 'Materials leftover from ore processing activities, excluding overburden',
         'Ws', 'Wastes', 'Other domestic or commercial rubbish')

tbl_hr_artefs <- gt(dat_hr_artefs) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_hr_artefs

```

```{r}
#| label: tbl-hr-aorg
#| tbl-cap: "Artefact origin (adapted from @weber2022)"

dat_hr_aorg <-
  tribble(~Code, ~Name, ~Description,
         'N', 'Not verifiable', 'Designation not possible / uncertain',
         'C', 'Consumer', 'Consumer-products (everyday-products) like tools, decorative items, etc', 
         'A', 'Agricultural', 'Agricultural products like straw bale nets, wire etc',
         'I', 'Industrial', 'Industrial products like broken pieces of industrial equipment, industrial waste',
         'B', 'Construction', 'All materials which were placed in soil for construction and supply purposes (e.g., pipes, power cables, geomembranes)')

tbl_hr_aorg <- gt(dat_hr_aorg) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_hr_aorg

```

```{r}
#| label: tbl-hr-afix
#| tbl-cap: "Fragment fixation (adapted from @weber2022)"

dat_hr_afix <-
  tribble(~Code, ~Name, ~Description,
         'NF', 'None',    'Artefacts are not included in soil aggregates, or soil is apedal',
         'SF', 'Solely',   'Some artefacts are incorporated into soil aggregates, but most are loose',
         'MF', 'Moderate', 'The majority of artefacts are incorporated into soil aggregates, but some pieces are loose',
         'CF', 'Complete', 'All artefacts are incorporated into soil aggregates')

tbl_hr_afix <- gt(dat_hr_afix) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_hr_afix

```

```{r}
#| label: tbl-hr-ashp
#| tbl-cap: "Artefact shape (adapted from @weber2022)"

dat_hr_ashp <-
  tribble(~Code, ~Name, ~Description,
         'P', 'Preserved', 'The original manufactured shape of the object is recognisable',  
         'B', 'Broken',  'The fragment is recognisable as having been part of a larger object',
         'F', 'Flat', 'Tabular or sheet-like',
         'A', 'Angular', 'Sharp-edged',
         'R', 'Rounded', 'Has edges that are not sharp',
         'S', 'Spherical', 'Ball shaped or ovoid')

tbl_hr_ashp <- gt(dat_hr_ashp) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_hr_ashp

```

```{r}
#| label: tbl-hr-adeg
#| tbl-cap: "Artefact degradation (adapted from @weber2022)"

dat_hr_adeg <-
  tribble(~Code, ~Name, ~Description,
         'N', 'Not verifiable', 'Designation not possible / uncertain',    
         'F', 'Fresh', 'Fresh artefacts with bright colours, unaffected shape, without signs of deterioration (cracks, broken edges, rough surfaces)', 
         'I', 'Incipient alteration', 'Artefacts with incipient deterioration surface such as first cracks, broken edges or roughened/grooved surfaces',
         'D', 'Degraded', 'Artefacts with clear indications of ageing such as pale colours or obvious discolouration, strong fragmentation, frayed edges',
         'S', 'Strongly degraded', 'Artefacts with strongly degraded surfaces, faded colours, soft surface, frayed areas or edges')

tbl_hr_adeg <- gt(dat_hr_adeg) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_hr_adeg

```

## Roots {#sec-root}

Plant roots affect water infiltration, bind and rework the soil as they grow, and exert strong local control over soil chemistry. Their position in the soil profile can provide information about ped development, particle packing, and the presence of pans. Accurate measurement of root volumes *in situ* is technically challenging [@measurin2012; @freschet2021], requiring extensive sampling to account for spatial, temporal and interspecies variability. Such sampling may be necessary for accurate assessment of total soil composition. The majority of the root mass, particularly fine roots of herbaceous species and grasses, will be found in the topsoil. For soil classification, presence of roots and their relative position within the soil are key characteristics.

Note that this section applies to the live root mass only.

### Root abundance {#sec-root-abund}

Estimation of root volume from cross-sectional analyses is not accurate, and so the following method, reproduced from @internationalorganizationforstandardizationa must be considered at best a relative assessment of root abundance.

Abundance is communicated as a count per square decimetre (100 cm^2^), undertaken on a pit or exposure face. For very thin horizons, a 50 cm horizontal linear transect at the horizon midpoint may be used instead. Multiple split-open cores may also be examined provided the minimum 100 cm^2^ area can be examined.

For rapid assessment, report the total number of roots. For routine assessment, report counts separately for the size classes in @tbl-hr-rsz. Always confirm absence of roots by recording a [0]{.ceg}.

E.g. [V 30, F 5, M 5, C 0]{.ceg} for a mixture of very fine, fine and medium roots.

With adequate size data, count per area can be converted to an areal percentage, e.g. 10 roots with diameter 5 mm in a 10 x 10 cm space gives a total area of $\pi × 0.25^{2} × 10 = {2\ cm^{2}}$, or 2%.

The count data may be classified using @tbl-hr-rtab.

```{r}
#| label: tbl-hr-rsz
#| tbl-cap: "Plant root size classes (adapted from @internationalorganizationforstandardizationa)"

dat_hr_rsz <-
  tribble(~Code,    ~Name,  ~Description,
            'V', 'Very Fine', '<0.5',
            'F',   'Fine',         '≥0.5–<2.0',
            'M', 'Medium',      '≥2.0–<5.0',
            'C', 'Coarse',        '≥5.0')

tbl_hr_rsz <- gt(dat_hr_rsz) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_label(Description = 'Size range (mm)') |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_hr_rsz

```

### Root position {#sec-root-pos}

Where roots are present in a horizon, record how the roots are spatially distributed using the codes in @tbl-hr-rpos. For rapid assessment, record the dominant position of all roots. For routine assessment, split by size. Root position cannot be reliably observed in augered samples, or in cores where structural units are known to be larger than the core diameter.

E.g. [F B, C T]{.ceg} for a soil where fine roots are confined to voids but coarse roots are not.

```{r}
#| label: tbl-hr-rpos
#| tbl-cap: "Plant root position"

dat_hr_rpos <-
  tribble(~Code,        ~Name,  ~Description,
            'W',     'Within', 'Roots are primarily located within peds',
            'B',    'Between', 'Roots are primarily located between peds, within fissures or along bedding planes',
            'T', 'Throughout', 'Roots are evenly distributed within and between peds',
            'U',    'Unknown', 'Sampling method did not allow reliable observation')

tbl_hr_rpos <- gt(dat_hr_rpos) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red', 
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_hr_rpos

```

### Other parameters {#sec-root-other}

Information about particular plant species contributing to the root mass and spatial organisation beyond the positional classes above (e.g. presence of mats) may prove useful in certain contexts. No coding system is provided for these; record as free-text notes.

## Voids and pores {#sec-voidpore}

Voids and pores are an integral part of soil structure, forming the spaces that occur between and within the aggregates described in @sec-hr-stru. They may contain air, water, or both, and play a key role in aeration, drainage, and root penetration. They are also habitat for soil meso- and microfauna. Horizon voids exclude open surface cracks, which are noted in @sec-pr-crack, and fine pores, which are described in @sec-pore. Infill features, where voids have previously been created but are no longer empty of solid material, are described in @sec-pr-inf. Voids and pores range in size from many centimetres down to nano-scale gaps in clay minerals. Only the larger end of the scale can be reasonably assessed in the field. As with measurement of roots, void space is most accurately determined in the laboratory.

### Voids {#sec-void}

Voids comprise pores \>2 mm wide. Dimensions of large, isolated voids like burrows can be measured directly and volume calculated. Areal percentage of smaller inter-ped spaces can be assessed across a cut face using the abundance charts in @sec-in-abundpics or by counting void openings. Use a minimum area of 100 cm^2^ for this assessment.

For rapid assessment, estimate the areal percentage of the horizon volume occupied by void space. Always confirm absence of voids by recording a [0]{.ceg}.

For detailed assessment, identify void types using @tbl-hr-voidtype and estimate the percentage of each type seen. Optionally, add connectivity using @tbl-hr-voidcon. Note that void connectivity may vary with soil water status.

Example: [10% IT C]{.ceg} for the void space in a strongly pedal, polyhedral topsoil.

```{r}
#| label: tbl-hr-voidtype
#| tbl-cap: "Type of soil void"

dat_hr_voidtype <-
  tribble(~Code, ~Name, ~Description,
          'BP', 'Bedding plane', 'A parting or discontinuity due to stratification during deposition of sedimentary particles.', 
          'CH', 'Channel', 'Tubular void of biological origin that is less than 20 mm in diameter (e.g. insect burrow)',
          'TU', 'Tunnel',  'Tubular void of biological origin that is 20 mm or more in diameter (e.g. rabbit burrow) ',
          'FS', 'Fissure', 'Planar void with a width much smaller than length and depth. Fissures represent the release of strain caused by drying and usually bound individual structural units and form repetitively.',
          'IT', 'Interstitial', 'Irregular voids occupying space between aggregates that do not fit together',
          )

tbl_hr_voidtype <- gt(dat_hr_voidtype) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red', 
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20)) |>
  sub_missing(missing_text = '')

tbl_hr_voidtype
```

```{r}
#| label: tbl-hr-voidcon
#| tbl-cap: "Connectivity of soil voids"

dat_hr_voidcon <-
  tribble(~Code, ~Name, ~Description,
          'I', 'Isolated', 'Void spaces are closed off from each other',
          'P', 'Partly connected', 'Void spaces are sometimes connected to each other within the horizon, or continue into neighbouring horizons',
          'C', 'Continuous',  'Void spaces are well connected and form a continuous network')

tbl_hr_voidcon <- gt(dat_hr_voidcon) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20)) |>
  sub_missing(missing_text = '')

tbl_hr_voidcon
```

### Pores {#sec-pore}

Pores are open voids \<2 mm in diameter. Macropores are ≥0.75 mm, and micropores are \<0.75 mm. This division is based on the diameter above which pores are regarded as generated by soil fauna. Pores can be observed on freshly broken ped faces or broken faces of a block cut from an apedal massive horizon, but will not be observable in loose, apedal single-grain materials. Cut surfaces in moist or wet soils may smear, obscuring pore openings.

For routine assessment, note presence or absence of any pores visible without magnification. Ideally, make observations from at least 2-5 small samples from within a horizon.

For detailed assessment, count the number of macropores and micropores separately over a minimum area of 100 cm^2^. This may necessitate sampling several peds or small blocks from the target horizon. Counting micropores will require the use of a magnifier.

[size example diagram here]{.diags}

## Substrate-specific features {#sec-reg}

Horizons that have experienced limited weathering may exhibit geogenic features that relate to their parent material and its mode of emplacement. These can affect water movement, slope stability and plant growth [@wysocki2005; @juilleret2016].

### Bedding {#sec-reg-bed}

Transported soil parent materials may comprise multiple thin layers of alternating texture or lithology. These layers can be unconsolidated (e.g. lamina or cross-bedding in aeolian and fluvial sediments), partly or fully consolidated, and may subsequently have been compressed, folded and/or faulted. Where layers of subsolum material have been identified (@sec-pr-wdeg) and bedding is apparent within a layer, measure the median angle of the bedding off horizontal and note the degree of consolidation. For unconsolidated sediments, record the texture groups involved using @tbl-hr-earthgrps. If the bedding is consolidated, record the dominant lithologies (@sec-pm-lith) involved. Optionally, add a thickness range for the bedding in centimetres.

E.g. [25°, Sm/Ss, 10-30 cm]{.ceg} for thin, very firm marine sediments, or [2°, C/S, 0.1-0.5 cm]{.ceg} for clay-sand lamina.

### Fractures {#sec-reg-fra}

Rock at the base of the profile may be significantly fractured by tectonic and weathering processes. Where lithic or paralithic layers (@sec-pr-wdeg) have been identified and fractures are apparent, note fracture type using @tbl-hr-fractyp, modal angle off horizontal in degrees, and median spacing in centimetres. For single fractures, record spacing as [0 cm]{.ceg}.

```{r}
#| label: tbl-hr-fractyp
#| tbl-cap: "Fracture types"

dat_hr_fractyp <-
  tribble(~Code, ~Name, ~Description,
          'F', 'Fault', 'Tectonically induced fractures that may cut across bedding. Bedding displacement will be apparent along the fault',
          'C', 'Crush', 'Fracture patterns induced by crushing and shearing forces. Fractures will have relatively close spacing and irregular orientation.',
          'P', 'Pressure', 'Fracture patterns induced by release of overburden pressure. Fractures are likely to be near-horizontal or parallel to the surface and may be curved.')

tbl_hr_fractyp <- gt(dat_hr_fractyp) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20)) |>
  sub_missing(missing_text = '')

tbl_hr_fractyp
```

[diagrams or photos - one for each type]{.diags}
