---
title: "Soil Horizon description"
editor: source
---

```{r}
library(tidyverse) 
library(gt)
library(gtExtras)
library(ggtern)
library(extrafont)
loadfonts(quiet = TRUE)
```

## Identification



## Composition

Each soil horizon is first described in terms of its broad composition. Composition is divided into 

- Mineral fraction
    - fine earth (mineral particles < 2mm)
    - rock fragments (mineral particles > 2mm)
- Biological fraction
   - Fine organic matter
   - Coarse organic matter (timber, plant roots)
- Liquid fraction (moisture content)
- Voids

These fractions are all measured as percentage by volume.

[complicated exploding pie chart goes here]{.diag}

## Architecture

The architecture of the physical components - how they are spatially arranged - is described in terms of the fine earth structure, as well as visible surface features and...

## Consistence 

Composition and architecture give rise to particular soil physical behaviours 

## Origins

## Parent material origin

Parent material origin, or mode of emplacement, describes how soil parent materials have arrived at the point of observation. Multiple modes may have operated in different parts of the soil profile. 

Record one parent material origin code per soil horizon. 

Parent material origin is summarised across the top 100 cm of the soil profile as part of defining the [NZSC Soil Family]().

```{r}
#| label: tbl-ss-pmo
#| tbl-cap: "Parent material origin"

dat_ss_pmo <-
  tribble(~Code, ~Name, ~Description,
          'An', 'Anthropic', 'Deposits made by the direct actions of humans, including truncation, mixing, or deposition.',
          'Cl', 'Colluvium', 'Weathered soil and rock material mantling slopes that has been transported primarily by gravity and sheet wash.',
          'Fl', 'Alluvium', 'Sediments that have been deposited by streams, rivers and other running water.',
          'Gl', 'Glacial till', 'Poorly stratified, poorly sorted rock fragments, sand and mud, surface or near-surface deposits resulting from the transportation by and deposition from ice or meltwater from beneath or in close proximity to glacial ice.',
          'Lc', 'Lacustrine', 'Formed in and around lake beds; comprising extremely fine sediment deposited under very low-flow freshwater conditions.',
          'Mr', 'Marine' , 'Unconsolidated sediments saturated by brackish or saline water.',
          'Lh', 'Lahar', 'A flow of volcanic material, both ash and coarser products, mixed with water; often caused by the spilling over of a crater lake. Also used to describe the mounds formed on volcanic ring-plains where lahars stop.',
          'Tp', 'Tephra', 'A general term for all solid (rather than molten) materials ejected from a volcano during an eruption: includes boulders, lapilli and ash.',
          'Li', 'Litter', 'Partly or fully decomposed forest litter, equal to or greater than 30 cm depth that is **not** almost continually saturated in the natural state (**F** and **H** horizons).',
          'Pt', 'Peat', 'Partly decomposed plant remains in a water-saturated environment, such as a bog',
          'Lo', 'Loess', 'A blanket deposit of silt-sized particles (0.002--0.06 mm diameter); usually carried by wind from dry riverbeds or outwash plains during glacial and post-glacial periods.',
          'Sa', 'Aeolian sand', 'Wind-deposited sand-sized particles (0.05--2.0 mm diameter), i.e. dune sand.',
          'Rw', 'Highly Weathered Rock', 'Highly to completely weathered rock, formed in place by chemical weathering.',
          'Rx', 'Rock', '*In situ* unweathered to moderately weathered lithified geologic material comprising one or more minerals showing rock structure.',
          'Uk', 'Unknown', 'Origin cannot be determined with certainty.'
          )

tbl_ss_pmo <- gt(dat_ss_pmo) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(20), Name ~pct(20))

tbl_ss_pmo
```




