# Consistence {#sec-cons}

```{r}
library(dplyr)
library(tibble)
library(gt)
library(purrr)
library(stringr)
```

## Background {#sec-cons-bg}

Consistence is a general term for the field-assessed behaviour of soil materials in response to applied physical stress over a range of soil-water states from wet to dry. The consistence properties described for mineral soil materials are strength, failure, fluidity, penetration resistance, packing, sensitivity, induration, plasticity, stickiness, dilatancy, and dry coherence. For organic materials, rubbed fibre content is relevant.

### Soil-water content {#sec-cons-bg-swc}

Some consistence properties (strength, failure, penetration resistance) vary markedly with soil-water content. The soil moisture status of the material should therefore be recorded at the time of description (see @sec-pr-sws) . Consistence descriptions on the same soils at different moisture states allow the change in consistence with moisture to be described. Other consistence properties are described at reference water contents.

### Sampling for consistency tests {#sec-cons-bg-samp}

Consistency tests should be performed down-profile, with at least one sample tested per horizon. Samples should be taken from the mid-point of each horizon. Additional samples in transition zones may also be of interest. For most of these tests, a 30 mm sample cube of soil (27 cm^3^) must be carefully extracted, preferably with a sharp knife. Keeping the sample size and shape consistent helps produce consistent results. Tests which involve remoulding the soil may use smaller volumes, but must use no less than \~3 cm^3^ (a 15 mm cube).

Within each horizon, the consistency tests that can and should be undertaken will depend on the cohesion (@sec-dn-prep) of the soil materials in that horizon. @tbl-cn-ct illustrates which tests are appropriate for different soil materials.

```{r}
#| label: tbl-cn-ct
#| tbl-cap: "Consistence tests for different kinds of soil material"

dat_cn_ct <-
  tribble(~Test, ~grav, ~ncoh, ~coh, ~org, ~mstat,
          '[Strength - unconfined mass](#sec-cn-stru)', 'If test cube can be formed', 'No', 'Yes', 'Yes', 'Any',
          '[Strength - soil peds](#sec-cn-stra)', 'No', 'No', 'If pedal', 'No', 'Any',
          '[Strength - remoulded](#sec-sn-strr)', 'No', 'No', 'Yes', 'No', 'Moist or Wet',
          '[Strength - at plastic limit](#sec-sn-strp)', 'No', 'No', 'Yes', 'No', 'Moist or Wet',
          '[Sensitivity](#sec-sn-sens)', 'No', 'No', 'Yes', 'No', 'Moist or Wet',
          '[Plasticity](#sec-cn-pla)', 'Yes', 'No', 'Yes', 'No', 'Wet',
          '[Stickiness](#sec-cn-sti)', 'Yes', 'Yes', 'Yes', 'Yes', 'Wet',
          '[Dilatency](#sec-cn-dil)', 'No', 'Yes', 'Yes', 'No', 'Moist to Wet',
          '[Penetration resistance](#sec-cn-pres)', 'Yes', 'Yes', 'Yes', 'Yes', 'Any',
          '[Particle packing - excavation](#sec-cn-pack)', 'Yes', 'No', 'No', 'No', 'Non-saturated',
          'Particle packing - blade insertion', 'No', 'Yes', 'Yes', 'Yes', 'Non-saturated',
          '[Mode of failure](#sec-cn-fail)', 'No', 'No', 'Yes', 'Yes', 'Any',
          '[Fluidity](#sec-cn-flu)', 'No', 'No', 'Yes', 'Yes', 'Wet or Saturated',
          '[Induration](#sec-cn-ind)', 'No', 'No', 'No', 'No', 'Any',
          '[Rubbed fibre](#sec-cn-fib)', 'No', 'If organic', 'If organic', 'Yes', 'Wet or Saturated'
          )

tbl_cn_ct <- gt(dat_cn_ct) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(style = 'italic',
                                   color = 'grey80')),
            locations = purrr::map( c('grav', 'ncoh', 'coh', 'org') ,
                     \(x){cells_body(
                       columns = x,
                       rows = stringr::str_detect(!!sym(x),"No")
                     )})) |>
  fmt_markdown(columns = 1) |>
  cols_label(grav = 'Mineral, rock fragments ≥ 15%', 
             ncoh = 'Mineral, non-cohesive, rock fragments < 15%',
             coh  = 'Mineral, cohesive, rock fragments < 15%',
             org  = 'Organic',
             mstat = 'Moisture status') |>
  cols_width(Test ~pct(20))

tbl_cn_ct
```

::: {#nte-consnzsc .callout-note appearance="minimal" collapse="true" title="Consistence tests and the New Zealand Soil Classification (NZSC)"}
For classifying using the New Zealand Soil Classification, particular tests down to or at a certain depth may be required to progress through the Key to Soil Orders. At the soil order level, these are:

-   **Gley Soils:** must not have any moderately or very fluid (@sec-cn-flu) layers within 30 cm of the soil surface. Unconfined soil strength (@sec-cn-stru) and cementation (@sec-cn-ind) tests may be required to confirm absence of an underlying fragipan.
-   **Ultic Soils:** need an assessment of soil strength (@sec-cn-stru) in the B horizon under moist conditions. May need assessment of failure mode (@sec-cn-fail) over a range of moisture conditions to rule out oxidic or cutanoxidic horizons. May need assessment of sensitivity (@sec-cn-sens) and stickiness (@sec-cn-sti) to rule out allophanic soil materials.
-   **Podzol Soils:** may need unconfined strength (@sec-cn-stru) and failure mode (@sec-cn-fail) tests to help identify a Podzolic B horizon, as well as the presence of humus- and ortstein-pans.
-   **Allophanic Soils:** may need assessment of sensitivity (@sec-cn-sens) and stickiness (@sec-cn-sti) to help confirm the presence of allophanic soil materials.
-   **Melanic Soils:** may require stickiness assessment (@sec-cn-sti) in the B horizon.
-   **Semiarid Soils:** may require cementation tests (@sec-cn-ind) to confirm fragipan absence.
-   **Oxidic Soils:** May need assessment of ped strength (@sec-cn-stra), soil strength (@sec-cn-stru), and failure mode (@sec-cn-fail) over a range of moisture conditions to confirm presence of an oxidic horizon. Stickiness (@sec-cn-sti) and plasticity (@sec-cn-pla) may also be helpful.
-   **Granular Soils:** May need assessment of ped strength (@sec-cn-stra), soil strength (@sec-cn-stru), and failure mode (@sec-cn-fail) over a range of moisture conditions to confirm presence of a cutanoxidic horizon. Stickiness (@sec-cn-sti) and plasticity (@sec-cn-pla) may also be helpful.
-   **Pallic Soils:** may require a negative cementation test (@sec-cn-ind) to help identify any fragipans. May require assessment of failure mode (@sec-cn-fail) to help confirm Brittle-B horizon presence. May require ped strength assessment in the B horizon.
-   **Recent Soils:** must not have any moderately or very fluid (@sec-cn-flu) layers within 30 cm of the soil surface.

Strength, fluidity, stickiness, mode of failure, and rubbed fibre content tests are also common diagnostic features at the group and subgroup level in the NZSC.

:::

## Cohesion {#sec-dn-prep}

The nature of the soil materials in each horizon determines what tests can and should be done. The first step is to determine if the sample is cohesive.

-   **Cohesive soil materials** may have plasticity and stickiness characteristics and have a strength which can be measured in unconfined compression. All listed properties, except particle packing, can be described for non-stony, cohesive soils.
-   **Non-cohesive soil materials** have no shear resistance other than that developed by internal friction between grains. Non-cohesive soil materials, by definition, have no unconfined dry strength or plasticity, and the only property that can be meaningfully described is degree of packing or particle packing.

The range of properties which can be described in stony, cohesive soils varies with rock fragment content. In practice, this depends on the feasibility of obtaining samples for the required tests.

The distinction between cohesive and non-cohesive is recognised by performing remoulding (@sec-cn-strr), plasticity (@sec-cn-pla), and dilatancy (@sec-cn-dil) tests in the following sequence (@fig-cn-cohes). Where doubt still remains after these tests have been completed, the dry coherence test (@sec-cn-dryco) may be necessary.

```{mermaid}
%%| label: fig-cn-cohes
%%| fig-cap: Determining sample cohesion

flowchart TD
  q1["Can a sample cube be formed from remoulded material?"] -->|yes| q2["Is the material plastic?"]
  q2 ---->|yes| a1(["**Cohesive**"])
  q1 -->|no| q3["Is the material plastic?"]
  q3 -->|yes| q4
  q2 -->|no| q4["Does the material have a fast positive reaction to the dilatancy test?"]
  q4 -->|no| q5["Does the material have a distinctly negative reaction to the dilatancy test?"]
  q4 --->|yes| a2
  q3 ---->|no| a2(["**Non-cohesive**"])
  q5 -->|yes| a1
  q5 -->|no| a2

```

### Dry coherence {#sec-cn-dryco}

This test may be used as the final arbiter for distinguishing cohesive from non-cohesive soils if other test results are inconclusive, e.g. in fine silt.

*Procedure*

1.  Remould a 30 mm test sample into a cube, and air-dry.
2.  If the cube collapses under its own weight or after being lightly touched, treat the soil as non-cohesive.

## Strength {#sec-cn-str}

Strength is also known as 'resistance to crushing'. This property is assessed using hand or body-weight strength, so the user must calibrate themselves against the classes in @tbl-cn-str. @drohan2020 discuss various options for calibration, including the use of pinch gauges or push-pull gauges to learn what different levels of force feel like.

Each of the tests in this section are evaluated against @tbl-cn-str.

```{r}
#| label: tbl-cn-str
#| tbl-cap: "Strength (resistance to crushing) classes"

dat_cn_str <-
  tribble(~Code, ~Name, ~Failure_cond, ~amt_force_n,
          1, 'Very Weak', 'Very gentle force between straight thumb and forefinger', '< 8',
          2, 'Weak', 'Gentle force between straight thumb and forefinger', '8-20',
          3, 'Slightly firm', 'Moderate force between straight thumb and forefinger', '20-40',
          4, 'Firm', 'Strong/maximum force between straight thumb and forefinger', '40-75',
          5, 'Very firm', 'Cannot crush between locked hands of an average person. Can crush with gentle force underfoot', '75-150',
          6, 'Hard', 'Full body weight (~80 kg) applied slowly', '150-800',
          7, 'Very hard', 'Withstands full body weight', '> 800'
          )

tbl_cn_str <- gt(dat_cn_str) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_label(Failure_cond = 'Failure condition',
             amt_force_n = 'Force required (N)') |>
  cols_width(Code ~pct(10), Name ~pct(20), amt_force_n ~pct(20))

tbl_cn_str
```

### Unconfined soil strength {#sec-cn-stru}

Unconfined soil strength describes how strongly soil material resists crushing when unsupported by surrounding material. It provides a simple field indicator of structural integrity in cohesive soils.

-   Test at observed field moisture condition.
-   If a test cube cannot be formed, the result is strength class [1]{.ceg} 'Very Weak' (see @tbl-cn-str). This can happen when soil pedality is weak or absent, but also when pedality is strong and peds are small.
-   This test cannot be reliably performed on auger-extracted samples.

*Procedure*

1.  Perform the test on a minimally disturbed 30 mm cube.
2.  Hold the cube between the thumb and forefinger. Both digits must be straight and fully extended.
3.  Pinch down. If the cube can be crushed, rate the difficulty using categories [1-4]{.ceg} in @tbl-cn-str.
4.  If the cube cannot be crushed by hand, attempt to crush it slowly underfoot, against a hard surface. Rate the difficulty using categories [5-7]{.ceg} in @tbl-cn-str.

### Unconfined aggregate strength {#sec-cn-stra}

Unconfined aggregate strength describes the physical integrity of soil aggregates (peds) under natural conditions. It provides a simple field indicator of soil structural stability in response to management practices and different environmental conditions.

-   Test at observed field moisture condition.
-   Only peds 30 mm or larger can be reliably tested; peds larger than 50 mm should be trimmed down to 30 mm.
-   This test cannot be reliably performed on auger-extracted samples.

*Procedure*

1.  Perform the test on an extracted ped.
2.  Hold the ped between the thumb and forefinger. Both digits must be straight and fully extended.
3.  Pinch down. If the ped can be crushed, rate the difficulty using categories [1]{.ceg}[–]{.ceg}[4]{.ceg} in @tbl-cn-str.
4.  If the ped cannot be crushed by hand, attempt to crush it slowly underfoot, against a hard surface. Rate the difficulty using categories [5–7]{.ceg} in @tbl-cn-str.

### Remoulded soil strength {#sec-cn-strr}

This test determines the strength of the soil after significant disturbance and reworking, disrupting the structure and any other binding materials like plant roots.

-   Test at observed field moisture condition.
-   Perform the test on an extracted 30 mm cube or sample of equivalent volume.
-   This test may be performed on auger-extracted samples.

*Procedure*

1.  Perform the test on an extracted 30 mm cube or sample of equivalent volume.
2.  Place the sample in a small plastic bag and seal with the air excluded (this maintains field moisture content while remoulding). Work the sample until its structure is completely broken down. This may take several minutes.
3.  Still in the bag, remould the soil into a cube and assess its strength against @tbl-cn-str.
4.  Use @tbl-cn-sens to also record the ratio of unconfined to remoulded strength ('sensitivity', @sec-cn-sens).

## Mode of failure {#sec-cn-fail}

Failure is the manner in which unconfined soil yields to increasing stress. This property varies with soil moisture content.

-   Test at observed field moisture condition.
-   Perform the test on a minimally disturbed 30 mm cube, or an aggregate.
-   If diagnosing Oxidic or Cutanoxidic soil material for the New Zealand Soil Classification, repeat the test under dry to wet moisture conditions.

*Procedure*

1.  Perform the unconfined strength test according to @sec-cn-stru.
2.  At strengths of class [4]{.ceg} or less, mode of failure can be observed. Assess and record per @tbl-cn-fail.

```{r}
#| label: tbl-cn-fail
#| tbl-cap: "Failure mode classes"

dat_cn_fail <-
  tribble(~Code, ~Name, ~Description,
           'N', 'Not determinable', 'Soil is either too dry or too wet to reliably test',
           'V', 'Very friable', 'Test specimen cannot be formed, or crumbles under very slight stress on crushing within the hand, into aggregates predominantly <2 mm in size. In most instances the test specimen is difficult to obtain.',
          'F', 'Friable', 'Test sample cannot be formed, or crumbles under very slight stress on crushing within the hand, into aggregates predominantly >2 mm in size, or crumbles under slight stress on crushing within the hand into aggregates predominantly <2 mm in size.',
          'B', 'Brittle', 'Under slowly increasing pressure between the extended thumb and forefinger, a test specimen retains its size and shape with few or no cracks until it fractures abruptly into aggregates predominantly > 2 mm in size.',
          'S', 'Semi-deformable', 'Under increasing pressure between the extended thumb and forefinger the test specimen as a whole is perceptibly compressed in the direction along which pressure is exerted, but cracks develop and the specimen ruptures before it has been compressed to half its original thickness. Before rupture, parts crumble and fall away if the specimen is shaken while held between the thumb and forefinger.',
          'D', 'Deformable', 'Under slowly increasing pressure between the extended thumb and forefinger, the test specimen can be compressed to at least half its original thickness without the appearance of cracks and without rupture.')

tbl_cn_fail <- gt(dat_cn_fail) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red', 
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_cn_fail
```

## Particle Packing {#sec-sn-pack}

Particle packing assesses the stability of a non-cohesive soil.

-   Test only on soils where unconfined strength (@sec-cn-stru) cannot be assessed e.g., gravelly, cobbly or bouldery soils, or loose sands
-   This test requires a large pit or exposure
-   Do not test at wet or saturated field moisture conditions, as this is unsafe

*Procedure*

1.  Expose the soil profile.
2.  Assess ease of excavating and maintaining a vertical face using @tbl-cn-pack.

```{r}
#| label: tbl-cn-pack
#| tbl-cap: "Particle-packing classes for non-cohesive soils"

dat_cn_fail <-
  tribble(~Code, ~Name, ~Description,
          1, 'Loose', 'Easily dislodged by spade or pick; does not maintain a vertical face',
          2, 'Compact', 'Intermediate between loose and dense',
          3, 'Dense', 'Difficult to dislodge except with spade and by removing individual fragments; maintains a stable vertical face', 
          4, 'Very dense', 'Very difficult to dislodge except by pick and removing individual fragments; maintains a stable vertical face')

tbl_cn_fail <- gt(dat_cn_fail) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_cn_fail
```

::: {#nte-singbl .callout-note appearance="minimal" collapse="true" title="Use of the Singleton Blade"}
The Singleton blade was a device invented in the 1980s by the New Zealand pedologist Peter Singleton to improve on packing density assessment methods available at the time. The device was thin, flexible piece of metal with a notched blade that was inserted into the pit face, leaving a longer end protruding. A pocket penetrometer was then used to push the protruding end of the blade sideways until it reached 45° or the blade fell out of the pit face. The penetrometer reading was empirically correlated to packing density. Mention of this device can be found in many New Zealand soil publications from \~1980-2000 [@griffiths1999] but has not been made commercially available and is now rarely used.
:::

## Penetration resistance {#sec-cn-pres}

Penetration resistance is the capacity of the soil, in its natural confined state, to resist penetration by rigid objects. This property is commonly related to strength by empirical means (@tbl-cn-penr).

-   This test requires a large pit or exposure.

*Procedure*

1.  Expose a soil profile.
2.  Assess ease of penetrating a vertical face using @tbl-cn-penr.

::: {#nte-ppen .callout-note appearance="minimal" collapse="true" title="Use of pocket penetrometers"}
Pocket-sized penetrometers with a 6 mm flat tip are often recommended for assessing penetration resistance of cohesive materials in the field. These devices can usually measure no more than 400-1000 kPa and so may not be adequate for measurements of hard materials.

If a pocket penetrometer is used, take at least 10 readings from across the target horizon and record a median and range for resistance. The soil material being measured must be at or near field capacity when measured for the readings to be valid.
:::

```{r}
#| label: tbl-cn-penr
#| tbl-cap: "Penetration resistance classes"

dat_cn_penr <-
  tribble(~Code, ~Name, ~Description, ~kpa,
          1, 'Extremely low', 'Easily penetrated by fist', '0-500',
          2, 'Very low', 'Easily penetrated by thumb', '500 - 1000',
          3, 'Low', 'Readily indented by thumb but penetrated with considerable effort. Imprinted by finger with some effort.', '1000 - 1500',
          4, 'Moderate', 'Readily indented by thumbnail', '1500-2200',
          5, 'High', 'Barely imprinted by thumb or forefinger', '2200 - 3000',
          6, 'Very high', 'Indented by thumbnail. No imprint by thumb or finger', '3100 - 4000',
          7, 'Extremely high', 'Barely indented by thumbnail', '> 4000')

tbl_cn_penr <- gt(dat_cn_penr) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red', 
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20), kpa ~pct(20)) |>
  cols_label(kpa = 'Pressure (kPa)')

tbl_cn_penr
```

## Sensitivity {#sec-cn-sens}

The sensitivity of a soil reflects the extent to which its structure contributes to its strength, and how that strength is reduced when the structure is destroyed. In soils with an unconfined strength class of [4]{.ceg} firm or greater, sensitivity is reported as the ratio of unconfined soil strength (@sec-cn-stru) to remoulded strength (@sec-cn-strr). For soils with an unconfined strength class of 1–3, the 'shear test' is performed. The two methods are assigned equivalent classes in @tbl-cn-sens.

-   Test soil at a moist or wet state.
-   Perform the test on a minimally disturbed 30 mm sample cube.
-   This test cannot be reliably performed on auger-extracted samples.

*Procedure*

1.  Hold the sample cube between extended thumb and forefinger as for @sec-cn-str.
2.  Pinch and press the sample to create a lateral, 'shear' force, moving the finger forward and down over the thumb.
3.  Assess the sample behaviour against @tbl-cn-sens.

::: {.callout-warning appearance="minimal"}
Note that this test is not analogous to the shear vane test of @newzealandgeotechnicalsociety2005 and cannot be used as a substitute.
:::

```{r}
#| label: tbl-cn-sens
#| tbl-cap: "Sensitivity (strength loss on remoulding) classes"

dat_cn_sens <-
  tribble(~Code, ~Name, ~rel_loss, ~Description,
          1, 'Non-sensitive', '0-1', 'the sample does not smear under shear force, and may soften only slightly',
          2, 'Weakly sensitive', '2', 'the sample smears and skids under moderate shear force (20-40 kPa)',
          3, 'Moderately sensitive', '3', 'the sample suddenly turns fluid under moderate shear force (20-40 kPa) and the fingers skid. Some free water is evident on fingers after shearing',
          4, 'Very sensitive', '>3', 'the soil suddenly turns fluid and slippery under only gentle shear force (<20 kPa). Free water is easily seen on the fingers after shearing')

tbl_cn_sens <- gt(dat_cn_sens) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red', 
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_label(rel_loss = 'Relative strength loss on remoulding (classes)',
             Description = 'Shear test behaviour') |>
  cols_width(Code ~pct(10), Name ~pct(20), rel_loss ~pct(20))

tbl_cn_sens
```

## Plasticity {#sec-cn-pla}

Plasticity is the capacity of soil material to change shape continuously when stress is applied and to retain that shape after the removal of stress. Plasticity can help distinguish clayey from silty textures, with silts having relatively low plasticity.

-   Test soil at a wet state.
-   Perform the test on 3-4 cm^3^ of soil (a 15–20 mm cube).
-   This test may be performed on auger-extracted samples.

*Procedure*

1.  Thoroughly remould the sample, keeping it in a wet state.
2.  Adjust the water content to slightly above the plastic limit. In practice this water content is slightly above that at which a 4-mm diameter roll of sample crumbles but less than that when sample becomes sticky (@sec-cn-sti), if stickiness is exhibited.
3.  Attempt to roll the sample out into a rod of 40 mm length. Assess the minimum achievable diameter of the rod and its behaviour using @tbl-cn-plas. To test whether the roll will support its own weight, attempt to pick it up from one end and see if the roll breaks.

```{r}
#| label: tbl-cn-plas
#| tbl-cap: "Plasticity classes"

dat_cn_plas <-
  tribble(~Code, ~Name, ~Description,
          1, 'Non-plastic', 'A roll 40 mm long and 6 mm thick that supports its own weight cannot be formed.',
          2, 'Slightly plastic', 'A roll 40 mm long and 6 mm thick can be formed and will support its own weight, but a roll 4 mm thick will not support its own weight.',
          3, 'Moderately plastic', 'A roll 40 mm long and 4 mm diameter can be formed, and will support its own weight, but a roll 2 mm thick will not support its own weight.',
          4, 'Very plastic', 'A roll 40 mm long and 2 mm thick can be formed and will support its own weight.')

tbl_cn_plas <- gt(dat_cn_plas) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_cn_plas
```

## Stickiness {#sec-cn-sti}

Stickiness is the degree to which remoulded soil adheres to external objects. Stickiness can suggest the presence of clay (particularly 2:1 phyllosilicates) and has agricultural and mechanical implications, e.g. root crop harvesting can be difficult in sticky materials, as can excavation.

-   Test soil at a wet state.
-   Perform the test on 3-4 cm^3^ of soil (a 15–20 mm cube).
-   This test may be performed on auger-extracted samples.

*Procedure*

1.  Thoroughly remould the sample, keeping it in a wet state.
2.  Press the remoulded material between thumb and forefinger, and note its adherence to the fingers.
3.  Adjust the water content by working the sample in the hand to remove water, or by adding water as necessary to achieve the maximum stickiness that the material will exhibit.
4.  Assess maximum stickiness and record using @tbl-cn-stik.

```{r}
#| label: tbl-cn-stik
#| tbl-cap: "Stickiness classes"

dat_cn_stik <-
  tribble(~Code, ~Name, ~Description,
          1, 'Non-sticky', 'After release of pressure, practically no soil material adheres to the thumb or forefinger',
          2, 'Slightly sticky', 'After release of pressure, soil material adheres perceptibly to both thumb and forefinger but, as the digits are separated, it tends to come off one or the other cleanly. The material is not stretched appreciably when the thumb and finger are separated',
          3, 'Moderately sticky', 'After release of pressure, soil material adheres to both thumb and forefinger and tends to stretch somewhat and pull apart rather than pulling free from either digit',
          4, 'Very sticky', 'After release of pressure, soil material adheres so strongly to both thumb and forefinger that it is decidedly stretched when they are separated, and finally pulls apart leaving part of the soil on thumb and part on the forefinger.')

tbl_cn_stik <- gt(dat_cn_stik) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_cn_stik
```

## Dilatancy {#sec-cn-dil}

Dilatancy occurs due to volume change during shearing and is used as a criterion for distinguishing non-cohesive from cohesive soil materials. The test can also help determine the relative dominance of silt and clay (silty textures are more dilatant,

-   Test soil at a moist to wet state.
-   Perform the test on 3-4 cm^3^ of soil (a 15–20 mm cube).
-   This test may be performed on auger-extracted samples.

*Procedure*

1.  Adjust the soil to a moist to wet state.
2.  Roll into a ball \~30 mm diameter.
3.  Hold sample in upturned palm and slap the back of the hand containing the sample several times with the other hand, or shake the hand vigorously from side to side.

A dilatant reaction is shown by a shiny surface caused by water extrusion, and by a dull surface when the palm of the hand is folded to partly remould the sample, causing water to be drawn into the sample. Dilatant samples will also lose their ball shape, flattening down. Record dilatant behaviour as a binary, [true/false]{.ceg}.

## Fluidity {#sec-cn-flu}

Fluidity tests whether saturated soils have an unconfined strength that is largely eliminated by saturation. Such materials were generally emplaced under water, have never been drained, have a fine texture that minimises gravity-based settling, and are not overlain by enough other sediment to have been compressed.

-   Test soil at a saturated state.
-   Only test soils that are saturated under field conditions.
-   Perform the test on \~200 cm^3^ of soil (a large handful).
-   This test may be performed on auger-extracted samples.

*Procedure*

1.  Extract the sample and squeeze firmly in the hand.
2.  Evaluate the sample behaviour using @tbl-cn-fluid.

```{r}
#| label: tbl-cn-fluid
#| tbl-cap: "Fluidity classes"

dat_cn_fluid <-
  tribble(~Code, ~Name, ~Description,
          1, 'Slightly fluid', 'Some material tends to flow into the spaces between the fingers, but after full pressure is exerted most of the residue is left in the hand',
          2, 'Moderately fluid', 'Soil material flows easily between the fingers, but a small residue is left in the hand after full pressure is exerted',
          3, 'Very fluid', 'Soil material flows like a slightly viscous liquid between the fingers, and very little or no residue is left in the hand after full pressure is exerted')

tbl_cn_fluid <- gt(dat_cn_fluid) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red', 
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_cn_fluid
```

## Induration {#sec-cn-ind}

Induration is the degree to which soil particles are held together by some combination of cementation and close packing.

-   Perform the test on a minimally disturbed 30 mm cube, or aggregate.
-   This test may also be performed on individual nodules and other solids \< 30 mm.
-   This test cannot be reliably performed on auger-extracted samples.
-   This test requires a minimum one hour to complete so may be done immediately following fieldwork.

*Equipment*

-   Sponge (\~50 mm diameter, flat)
-   Deionised (DI) water
-   Sample container at least 40-50 mm deep

*Procedure*

1.  Test the soil strength at field-moist condition according to @sec-cn-stru.
2.  Place the sample on a wet sponge until thoroughly moistened, then immerse the sample in DI water for 1 hour.
3.  If the sample has not slaked (see @sec-ft-slak), retest the soil strength of the wet sample according to @sec-cn-stru.
4.  Assess the behaviour of the sample using @tbl-cn-ind.

```{r}
#| label: tbl-cn-ind
#| tbl-cap: "Induration classes"

dat_cn_ind <-
  tribble(~Code, ~Name, ~Description,
          1, 'Non-indurated', 'Sample slakes within one hour',
          2, 'Very weakly indurated', 'Has not slaked, strength class of 2-4',
          3, 'Weakly indurated', 'strength class of 5-6',
          4, 'Strongly indurated', 'strength class of 7',
          5, 'Very strongly indurated', 'strength class of >7; does not break when struck by a sharp blow with a hammer.')

tbl_cn_ind <- gt(dat_cn_ind) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red', 
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_cn_ind
```


## Rubbed fibre content {#sec-cn-fib}

Rubbed fibre content is an indicator of the degree of organic decomposition and can be used to help determine whether organic material qualifies as fibric, mesic or humic (@sec-tex-orgmod, @sec-hor-o).

- This test can be performed on any sample with an appreciable amount of visible and decomposing organic matter. Live plant roots should be removed before performing the test.
- This test can be performed on auger-extracted samples. However, exercise caution when extracting samples from below the water table as they may easily become cross-contaminated by overlying material.
- Perform the test on \~100 cm^3^ of soil (a small handful)
- Perform the test on wet materials.

*Procedure*

1. Extract the sample and squeeze to remove excess water
2. Rub the wet sample 10 times between thumb and forefinger, or knead a ball of sample ten times firmly in the palm of the hand.
3. Examine the proportion of visible plant fibres.

Note that larger pieces of rotting solid or soft timber (e.g. buried branches) should also be included in the total proportion of fibrous material.

```{r}
#| label: tbl-cn-fib
#| tbl-cap: "Fibric organic material classes"

dat_cn_ind <-
  tribble(~Code, ~Name, ~Description,
          1, 'None',     'No appreciable fibre',
          2, 'Minor',    'Fibrous material < 15%',
          3, 'Moderate', 'Fibrous material 15-75%',
          4, 'Major',    'Fibrous material > 75%')

tbl_cn_ind <- gt(dat_cn_ind) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red', 
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_cn_ind
```
