---
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
library(dplyr)
library(tibble)
library(gt)
```

# Secondary features {#sec-secondary}

Secondary soil features develop as a result of pedogenesis. Many create barriers to water movement or alter chemistry in a way that affects soil biota.

## Concentrations {#sec-hr-conc}

[These are bodies of pedogenic origin enriched in a particular constituent that are not associated with a particular surface within the soil. They may contrast sharply with the surrounding material in strength, composition or internal organisation, or differences with the surrounding material may be slight.]{.milne} 

For rapid assessment, record concentration form and abundance percentage using @tbl-hr-concform.

For detailed assessment, note form, type, abundance in percentage, and size range in mm. Colour is not required as it is usually determined by type and may have already been recorded as a colour pattern (@sec-col-patt). Abundance and size can be classified using @tbl-in-size and @tbl-in-abgen respectively.

Example: [NO F 10% 6-8 mm]{.ceg} for scattered small ironstone nodules.

```{r}
#| label: tbl-hr-concform
#| tbl-cap: "Concentration forms"

dat_hr_concform <-
  tribble(~Code, ~Name, ~Description,
          'CC', 'Concretion', 'hard concentrations with a layered internal structure, usually rounded',
          'NO', 'Nodule', 'hard concentrations with no internal structure, usually irregular',
          'CT', 'Crystal', 'A mineral arranged as interlocking or twinned macroscopic crystals or as a single macroscopic crystal, e.g., gypsum.',
          'SS', 'Soft masses', 'A non-cemented concentration that can seldom be removed from the soil as a discrete unit')

tbl_hr_concform <- gt(dat_hr_concform) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_hr_concform
```

```{r}
#| label: tbl-hr-conctype
#| tbl-cap: "Concentration types (adapted from @nationalcommitteeonsoilandterrain2023)"

dat_hr_conctype <-
  tribble(~Code, ~Name, ~Description,
          'E',  'Earthy', 'loamy or silty texture within a sandy or clayey matrix',
          'A',  'Argillaceous', 'clayey texture within a lower-clay matrix',
          'F',  'Ferruginious', 'Iron-dominated',
          'M',  'Manganiferous', 'Manganese-dominated',
          'N',  'Ferromanganiferous', 'Dominated by a mix of Fe and Mn',
          'K',  'Calcareous', 'Dominated by calcium carbonate (CaCO~3~)',
          'Y',  'Gypseous', 'Dominated by calcium sulfate (gypsum, CaSO~4~)',
          'Z',  'Salty', 'Dominated by salts more soluble than Calcium carbonate or calcium sulfate',
          'S',  'Sulfurous', 'Dominated by sulfur, e.g. jarosite in acid sulfate soils',
          'O',  'Organic', 'Organic texture in a low-organics matrix',
          'UK', 'Unknown', 'Composition cannot be reliably determined'
          )

tbl_hr_conctype <- gt(dat_hr_conctype) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  fmt_markdown(columns = 3) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_hr_conctype
```

## Coatings {#sec-hr-coat}

Coatings develop around other soil features, e.g. covering the surfaces of grains or peds, or lining the interior of voids. Some are chemical precipitates like concentrations, some form from material left behind as other material is selectively removed, and some are biological in origin (e.g. some ant species line their burrow tunnel walls with fine material).

For rapid assessment note the presence of coating type(s) using @tbl-hr-coattype.

For detailed assessment, optionally also note their abundance as percentage of soil available surfaces covered.

e.g. [OG P 80%]{.ceg} for well-developed 

```{r}
#| label: tbl-hr-coattype
#| tbl-cap: "Types of surface coating"

dat_hr_coattype <-
  tribble(~Code, ~Name, ~Description,
          'CC', 'Clay coats', 'Appear to have more clay than the soil matrix. They are often different in colour from the adjacent matrix and are usually recognisable in sandy or loamy soils, but can be difficult to recognise in clayey soils.',
          'ZC', 'Silt coats', 'Silt grains concentrated at surfaces; they may be residual concentrations resulting from clay removal, or may have moved in from horizons above',
          'SC', 'Sand coats', 'Sand grains concentrated at surfaces; they may be residual concentrations resulting from silt and/or clay removal, or may have moved in from horizons above',
          'OG', 'Organic coats', 'Surface features that have a moist colour value of 4 or less and are rich in organic matter compared to the interior of the solid.',
          'SQ', 'Sesquioxide coats', 'Films of sesquioxides on interfaces or impregnating interfaces. They vary in colour depending on the composition and degree of oxidation and hydration. They are difficult to identify chemically, and the different kinds cannot normally be distinguished in the field. Ferri-manganiferous coats, however, are normally very dark brown or black and effervesce vigorously with hydrogen peroxide',
          'MN', 'Manganese coats', 'dark coatings of precipitated manganese oxides.',
          'CB', 'Carbonate coats', 'Precipitated calcium carbonate (CaCO~3~)')

tbl_hr_coattype <- gt(dat_hr_coattype) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  fmt_markdown(columns = 3) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_hr_coattype
```

## Pans {#sec-pan}

Pans form when particular chemicals precipitate out of the soil solution and cement a layer of mineral particles (perhaps as a further accumulation and joining together of coatings and/or concentrations), or when particular particle size distributions and/or shape range become very tightly packed. Many pans are diagnostic in the NZSC.

For rapid assessment, record only the pan type using @tbl-hr-pantype. It is assumed that pans recorded this way are strongly developed enough to be classification-relevant. Note also that many (but not all) pans have dedicated conventional horizon names, and in those cases the horizon names these can be used for rapid assessment instead.

Pans can be described in detail in terms of type, thickness, continuity, structure, strength and cementation. Pans comprise distinct horizons, so their thickness is defined using horizon boundaries. Their strength and cementation are assessed using @tbl-cn-str and @tbl-cn-ind. Codes for type, continuity and structure are given below. Note that continuity can only be observed in large pits or exposures.

```{r}
#| label: tbl-hr-pantype
#| tbl-cap: "Types of Pan"

dat_hr_pantype <-
  tribble(~Code, ~Name, ~Description, ~Horizon,
          'C', 'Cultivation pan', 'Pan comprising compressed soil, induced by cultivation (e.g. use of heavy farm machinery)', '',
          'D', 'Densipan', 'Silty layers with a very high bulk density and extremly slow permeability', '\\*d (E horizons only)',
          'H', 'Humus-pan', 'pan cemented by aluminium and organic matter. Common in Podzols; sometimes known as \'coffee rock\'', 'Bhm',
          'F', 'Iron pan', 'interlocking nodules of Fe(/Mn) precipiates (bog iron), often at a texture contrast boundary', '',
          'K', 'Calcareous pan', 'pans created by the precipitation of calcium carbonate (CaCO~3~)', '\\*km (B or BC horizons)',
          'O', 'Ortstein pan', 'Pan cemented by iron and organic matter. Common in Podzols', 'Bsm',
          'P', 'Placic', 'a very thin (< 10 mm), dark, iron-dominated pan with sharp boundaries', 'Bfm',
          'Q', 'Silica pan', 'pans created by the precipitation of silica (SiO~2~)', '\\*qm (not A horizons)',
          'X', 'Fragipan', 'an earthy, high density, erosion-resistant layer common to Pallic soils', '\\*x (B or BC horizons)',
          'UK', 'Unknown', 'Pan origin cannot be clearly determined', ''
          )

tbl_hr_pantype <- gt(dat_hr_pantype) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  fmt_markdown(columns = c(3,4)) |>
  cols_width(Code ~pct(10), Name ~pct(20), Horizon ~pct(20))

tbl_hr_pantype
```

```{r}
#| label: tbl-hr-pancont
#| tbl-cap: "Pan continuity [@nationalcommitteeonsoilandterrain2023]"

dat_hr_pancont <-
  tribble(~Code, ~Name, ~Description,
          'C', 'Continuous', 'Extends as a layer with little or no break across 1 m or more.',
          'D', 'Discontinuous', 'Broken by cracks but original orientation of fragments is preserved.',
          'B', 'Broken', 'Broken by cracks and fragments are disoriented')

tbl_hr_pancont <- gt(dat_hr_pancont) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_hr_pancont
```

```{r}
#| label: tbl-hr-panstr
#| tbl-cap: "Pan structure (adapted from @nationalcommitteeonsoilandterrain2023)"

dat_hr_panstr <-
  tribble(~Code, ~Name, ~Description,
          'V', 'Massive', 'No recognisable structure',
          'S', 'Vesicular', 'Sponge-like structure having large pores, which may or may not be filled with softer material',
          'C', 'Concretionary', 'Spheroidal concretions cemented together',
          'N', 'Nodular', 'Irregular nodules cemented together',
          'P', 'Platy', 'Plate-like units cemented together')

tbl_hr_panstr <- gt(dat_hr_panstr) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_hr_panstr
```

## Stress features {#sec-hr-stress}

Stress features develop along fissures or on ped faces in soils that shrink and swell as water content changes. Stress features are usually shiny but do not differ in colour from the soil matrix. Magnification may be required to observe them clearly. Stress features will not develop in sandy soil textures (@sec-tex-gen).

Stress features can be optionally noted during a detailed soil description. Describe type and abundance as percentage of ped faces covered.

Example: [SL 30%]{.ceg} in a well-developed Vertic Melanic soil.

```{r}
#| label: tbl-hr-pftype
#| tbl-cap: "Types of pressure face"

dat_hr_pftype <-
  tribble(~Code, ~Name, ~Description,
          'SL', 'Slickensides', 'Smooth faces with linear grooves, from peds sliding past each other.',
          'PF', 'Pressure faces', 'Smooth planar faces from peds pressing against each other without sliding.' )

tbl_hr_pftype <- gt(dat_hr_pftype) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_hr_pftype
```

## Biological features {#sec-hr-bio}

Biological secondary features are mostly created by plant roots as they grow into the soil, and soil fauna as they create burrows.

For rapid assessment, note type(s) using the codes in @tbl-hr-biotype

For detailed assessment, also note the areal percentage of the exposed horizon face occupied by the feature.

Example: [T 30%]{.ceg} for infilled earthworm burrows creating an A/Bw horizon.

```{r}
#| label: tbl-hr-biotype
#| tbl-cap: "Types of secondary biological feature"

dat_hr_biotype <-
  tribble(~Code, ~Name, ~Description,
          'T', 'Tunnel', 'Infilled earthworm or insect burrow, < 20 mm in diameter',
          'B', 'Burrow', 'Infilled former macrofauna burrow, >20 mm in diameter',
          'C', 'Casts',  'Earthworm wastes forming balls of fine soil material')

tbl_hr_biotype <- gt(dat_hr_biotype) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_hr_biotype
```
