---
title: "Soil Horizon description"
editor: source
---

```{r}
library(tidyverse) 
library(gt)
library(gtExtras)
library(ggtern)
library(extrafont)
loadfonts(quiet = TRUE)
```

## Identification



## Composition

Each soil horizon is first described in terms of its broad composition. Composition is divided into 

- Mineral fraction
    - fine earth (mineral particles < 2mm)
    - rock fragments (mineral particles > 2mm)
- Biological fraction
   - Fine organic matter
   - Coarse organic matter (timber, plant roots)
- Liquid fraction (moisture content)
- Voids

These fractions are all measured as percentage by volume.

[complicated exploding pie chart goes here]{.diag}

## Architecture

The architecture of the physical components - how they are spatially arranged - is described in terms of the fine earth structure, as well as visible surface features and...

## Consistence 

Composition and architecture give rise to particular soil physical behaviours 

## Origins

## Parent material origin

Parent material origin, or mode of emplacement, describes how soil parent materials have arrived at the point of observation. Multiple modes may have operated in different parts of the soil profile. 

Record one parent material origin code per soil horizon, using the codes in @tbl-ss-pmo. At the horizon level, the [An]{.ceg} Anthropic class can be subdivided by origin using the codes in @tbl-hr-antype (e.g. [Ann]{.ceg} for a fill pad in a residential housing development).

```{r}
#| label: tbl-hr-antype
#| tbl-cap: "Subtypes of Anthropic soil materials"

dat_hr_antype <-
  tribble(~Code, ~Name, ~Description,
          'm', 'Māori', 'Soil modifications dating from Māori arrival in New Zealand, particularly pre-modern soil management practices',
          'p', 'Pakeha', 'Soil modifications dating from Pakeha (non-Māori) arrival in New Zealand up to modern times (e.g. European agricultural practices)',
          'n', 'Modern', 'Anthropic modifications resulting from modern techniques or addition of modern materials to the soil (e.g. plastics)'
          )

tbl_hr_antype <- gt(dat_hr_antype) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(20), Name ~pct(20))

dat_hr_antype
```

Parent material origin is also summarised across the top 100 cm of the soil profile as part of defining the [NZSC Soil Family](in_06_nzsc_familysibling.qmd#sec-fam-pmo). 






