# Horizon architecture {#sec-hr-arch}

```{r}
library(dplyr)
library(tibble)
library(gt)
```

Each soil horizon may include any or all of the following major components:

-   Solid fraction
    -   Fine earth
    -   Rock Fragments
    -   Artefacts
-   Biological fraction
    -   Fine organic matter
    -   Coarse organic matter
    -   Living plant roots
-   Liquid fraction
    -   Soil solution
-   Void space fraction
    -   Voids
    -   Pores
        -   Macropores
        -   Micropores

These components can be communicated as compositional percentage by volume, adding to 100%. Volumetric water content is best determined in the laboratory, and so is not described further in this handbook.

[complicated exploding pie chart or similar goes here]{.diags}

Soil horizon architecture involves describing the nature and the physical arrangement of the above components.

## Structure {#sec-hr-stru}

Structure refers to the physical arrangement of the fine-earth fraction (mineral materials \< 2 mm). The type of structure present in a horizon and the extent of its development is determined by the fine-earth texture (@sec-hr-eatxt), mineralogy, chemistry, the action of soil biota, weathering, and land use history. Well-developed structural units are commonly referred to as 'peds' and can be identified by observing any or all of the following:

-   repeating, regular units of similar shape and size
-   small gaps or planes of physical weakness between units
-   outer surfaces that have a distinct appearance when compared to the ped interior.

Peds can range in size from \<1 mm to \>1 m and sometimes 'nest', where larger units break down into smaller units. Given their potential size range, the method of profile exposure determines whether structures can be reliably observed and what maximum unit size will be detectable.

### Method: Assessing structure in the field

-   To assess the structure in an [E]{.ceg} or [P2]{.ceg} type profile (@tbl-pr-gear), insert a spade horizontally near the base of the horizon and push down on the handle, levering out a section of soil. Drop the spade blade with the soil, approximately 0.3 m onto a hard surface, e.g., the blade of another spade. Remove major roots that bind the mass together and again note the attributes of the newly produced solids. Note that this procedure will not necessarily disaggregate very moist or wet soils.
-   Structure may also be observed *in situ* by using a small hand tool to gently pick back the profile face. See @sec-dig. Structure of hard layers may need to be assessed in place, or by careful removal of individual units, and will usually require an [E]{.ceg} or [P2]{.ceg} type profile.
-   For smaller exposures ([P1]{.ceg}, [C1]{.ceg}, [C2]{.ceg}), remove a handful of soil from the extracted horizon and carefully disaggregate the soil mass to segregate all peds. Do not destroy peds to create fragments, or disaggregate existing apedal material. Note the attributes of the solids as instructed below. When coring, also use a spade to extract a block of adjacent surface material so that at least the topmost horizon can be assessed in detail.
-   Assessing structure is generally impossible for [A]{.ceg} augered profiles, as the extraction method destroys it. However, the surface horizon should still be extracted with a spade and assessed if possible.

Direct measurement of peds will help estimate size. Weighing, or filling volumetric containers, may help estimate abundance.

### Structure origin {#sec-str-origin}

Structural development can be controlled by different processes depending on vertical position within the profile and land use history [@anderson2024]. Record the dominant mode of origin for the structural units observed in each horizon using the options in @tbl-hr-form.

```{r}
#| label: tbl-hr-form
#| tbl-cap: "Type of structural development (adapted from @anderson2024)"

dat_hr_form <-
  tribble(~Code, ~Name,  ~Description,
          'P', 'Pedogenic', 'Structural units meet the definition of natural soil peds and occur largely within the solum',
          'G', 'Geogenic', 'Structural units are inherited from parent materials that have undergone limited weathering e.g. bedding planes, and occur largely below the solum',
          'A', 'Anthropogenic', 'Structural units have been induced by human activity, usually within the solum and most commonly in surface horizons',
          'U', 'Unknown', 'Type of structural development could not be determined with confidence')

tbl_hr_form <- gt(dat_hr_form) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_hr_form

```

### Structure development {#sec-hr-stru-pedality}

Degree of structural development (also frequently referred to as 'pedality' or 'aggregation') provides information about permeability, in concert with texture and firmness.

For rapid assessment, record degree of overall structural development once per horizon using @tbl-hr-pedality. For routine or detailed assessment, record degree of development for each nested structural unit observed (if more than one is seen; see @sec-hr-stru-nest).

```{r}
#| label: tbl-hr-pedality
#| tbl-cap: "Degree of structural development"

dat_hr_pedality <-
  tribble(~p, ~Code, ~Name,  ~Description,
          'Apedal',  'G', 'Single grain', 'Material is structureless or has <15% identifiable structural units, is uncemented and loosely packed, and falls apart into single grains when disturbed',
          'Apedal', 'V', 'Massive', 'Material is structureless or has <15% identifiable structural units, is uncemented and packed closely enough to hold together until substantially disturbed, and has no fissures spaced closer than 200 mm',
         'Pedal', 'W', 'Weak', 'Material where 15–25% of the material comprises coherent structural units',
          'Pedal', 'M', 'Moderate', 'Material where 25–75% of the material comprises coherent structural units',
          'Pedal', 'S', 'Strong', 'Material where >75% of the material comprises coherent structural units')

tbl_hr_pedality <- gt(group_by(dat_hr_pedality, p)) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%',
    row_group.as_column = TRUE
  ) |>
  tab_style(style = list(cell_text(weight = 'bold',
                                   v_align = 'middle')),
            locations = cells_row_groups()) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 2)) |>
  cols_width(p ~pct(15), Code ~pct(10), Name ~ pct(20))
  
tbl_hr_pedality

```

### Structure shape {#sec-hr-stru-shp}

Where distinct structural units can be identified (development types [W]{.ceg}, [M]{.ceg}, [S]{.ceg}), record their dominant shape using @tbl-hr-pedshp.

+------------+----------------+----------------------------------------------------------------------------------------------------------------------------------------+-----------------------------+
| Code       | Name           | Description                                                                                                                            | Orientation                 |
+:===========+:===============+:=======================================================================================================================================+:============================+
| [AB]{.ceg} | Angular blocky | -   bounded by relatively flat smooth, roughly equal faces                                                                             | None                        |
|            |                | -   number of faces variable                                                                                                           |                             |
|            |                | -   most vertices angular                                                                                                              |                             |
|            |                | -   usually much accommodation to the faces of surrounding aggregates                                                                  |                             |
+------------+----------------+----------------------------------------------------------------------------------------------------------------------------------------+-----------------------------+
| [BL]{.ceg} | Blocky         | -   bounded by undulating rough faces                                                                                                  | None                        |
|            |                | -   number of faces variable                                                                                                           |                             |
|            |                | -   many vertices rounded                                                                                                              |                             |
|            |                | -   limited accommodation to the faces of surrounding aggregates                                                                       |                             |
+------------+----------------+----------------------------------------------------------------------------------------------------------------------------------------+-----------------------------+
| [PH]{.ceg} | Polyhedral     | -   bounded by relatively flat smooth, unequal faces                                                                                   | None                        |
|            |                | -   more than six faces; most vertices angular                                                                                         |                             |
|            |                | -   usually much accommodation to the faces of surrounding aggregates                                                                  |                             |
|            |                | -   re-entrant angles between adjoining faces - peds 'lock' together                                                                   |                             |
+------------+----------------+----------------------------------------------------------------------------------------------------------------------------------------+-----------------------------+
| [SR]{.ceg} | Spheroidal     | -   bounded by curved or very irregular faces;                                                                                         | None                        |
|            |                | -   limited accommodation to the faces of surrounding aggregates                                                                       |                             |
|            |                | -   sometimes with many visible pores                                                                                                  |                             |
+------------+----------------+----------------------------------------------------------------------------------------------------------------------------------------+-----------------------------+
| [LT]{.ceg} | Lenticular     | -   bounded by curved faces                                                                                                            | Horizontal or shallow angle |
|            |                | -   overlapping, lens-shaped aggregates generally parallel to the soil surface that are thick at the centre and taper toward the edges |                             |
|            |                | -   usually much accommodation to the faces of surrounding aggregates                                                                  |                             |
+------------+----------------+----------------------------------------------------------------------------------------------------------------------------------------+-----------------------------+
| [PM]{.ceg} | Prismatic      | -   bounded by relatively flat faces                                                                                                   | Vertical                    |
|            |                | -   vertically elongated units with angular vertices and flat tops                                                                     |                             |
|            |                | -   much accommodation to the faces of surrounding aggregates                                                                          |                             |
+------------+----------------+----------------------------------------------------------------------------------------------------------------------------------------+-----------------------------+
| [CO]{.ceg} | Columnar       | -   bounded by relatively flat faces;                                                                                                  | Vertical                    |
|            |                | -   vertically elongated units with angular to rounded vertices and rounded (domed) tops                                               |                             |
+------------+----------------+----------------------------------------------------------------------------------------------------------------------------------------+-----------------------------+
| [PL]{.ceg} | Platy          | -   bounded by relatively flat horizontal faces                                                                                        | Horizontal                  |
|            |                | -   much accommodation to the faces of surrounding aggregates                                                                          |                             |
+------------+----------------+----------------------------------------------------------------------------------------------------------------------------------------+-----------------------------+
| [WL]{.ceg} | Wedge-like     | -   bounded by flat faces                                                                                                              | Horizontal                  |
|            |                | -   interlocking wedges or lenses that terminate in pronounced angular vertices                                                        |                             |
|            |                | -   ends of vertices may be missing                                                                                                    |                             |
|            |                | -   much accommodation to the faces of surrounding aggregates                                                                          |                             |
+------------+----------------+----------------------------------------------------------------------------------------------------------------------------------------+-----------------------------+

: Structural shapes and identifying characteristics (adapted from @milne1995, @iussworkinggroupwrb2022, and @ncst2024) {#tbl-hr-pedshp .striped cols-width=\[10,20,50,20\]}

[add diagrams]{.diags}

::: {#nte-nut .callout-note collapse="true" title="Shorthand terms for structural units" appearance="minimal"}
Some pedogenic structural terms that don't seem to fit into the above schema are still in common use.

-   Crumb structure: a very fine structure seen in many silty to loamy horizons. The appearance is as of breadcrumbs, with 0.5–2 mm rounded peds loosely to firmly packed, depending on depth and drainage. Larger weak subangular blocky structures in well-drained allophanic soils almost always break down into moderately or strongly developed crumb structure.
-   Nut structure: a larger version of crumb, usually on the order of 5–20 mm in diameter, often encased in a network of grass roots. Nut structures are more common in topsoils but can also occur in well-drained loamy subsoils.

Both structures are more correctly described as type [P]{.ceg} with [SR]{.ceg} shape. 'Crumbs' have a size range of about 1 ± 0.5 mm and 'nuts' are about 15 ± 10 mm. Their use as a shorthand description is convenient, but the terms should not be used in databases because they conflate size and shape in a way that the modern schemas seek to avoid.

Similarly, the terms 'clod' and 'fragment', previously used to signify tillage-induced structures greater and less than 100 mm in length respectively, also conflate size and shape. These should more correctly be described as type [A]{.ceg} structures with (usually) [BL]{.ceg} or [PH]{.ceg} shape.
:::

### Structure size {#sec-hr-stru-size}

Record the median length of the structural units observed, along their longest axis. Record in mm with a range estimate, e.g. [25 ± 2 mm]{.ceg}.

For rapid assessment, rather than recording a numeric value, the dominant size may be noted using the NZSC Functional Horizon convention (@tbl-in-fhstr): [C]{.ceg} for Coarse, [F]{.ceg} for Fine.

#### Nesting units {#sec-hr-stru-nest}

Structural units are commonly nested, with large structures breaking down to smaller ones that may also be a different type and shape. For highly detailed descriptions, multiple sets of nesting peds may be described in terms of their origin, development, shape and size.

For example, [P S PR 80 ± 10]{.ceg} breaking to [P M BL 20 ± 2]{.ceg}.

## Fragments {#sec-frag}

Fragments are mineral or non‑mineral components \> 2 mm within a soil horizon. They provide information about soil development, particularly mode of emplacement (@sec-pr-orig). They also reduce the soil’s water‑holding capacity and, when large and abundant enough, can restrict land use and management practices.

For very rapid assessment, estimate total fragment abundance (rocks and non-rock materials as well as any artefacts) as an integer percentage of the whole horizon volume. Where fragments are absent, always record [0%]{.ceg} as confirmation.

### Rock fragments {#sec-rockfrag}

Rock fragments are soil mineral particles \> 2 mm. For rapid assessment, use the texture class code modifiers provided for rock fragments in @sec-hr-rftex, and confirm absence by recording an abundance of [0%]{.ceg}.

Example: [SL(G)]{.ceg} for a sandy loam soil with 5-35% gravels (size range 2-20 mm).

For routine assessment, append an estimate of rock fragment abundance as integer percentage of the whole horizon volume, as well as dominant lithology class using @tbl-ss-pm and shape using @tbl-hr-rockfrshp.

Example: [SL(G) 15% Hs R]{.ceg} for a sandy loam soil with 15% rounded greywacke gravels.

For detailed assessment, one may optionally continue to add detail by recording volumetric abundance, lithology and shape for each of the three rock fragment size classes given in @tbl-hr-rocksz. The three additional gravel size subdivisions in @tbl-hr-gvquals may also be used. Picking and sieving from a known volume of soil will likely be required for sufficient accuracy when making detailed assessments.

```{r}
#| label: tbl-hr-rockfrshp
#| tbl-cap: "Coarse fragment shape classes"

dat_hr_rockfrshp <-
  tribble(~Code, ~Name, ~Description,
         'R', 'Rounded',   'Rock fragments considerably smoothed by transport processes',
         'A', 'Angular',   'Rock fragments with little smoothing by transport processes',
         'M',   'Mixed',   'Both rounded and angular fragments mixed together') 

tbl_hr_rockfrshp <- gt(dat_hr_rockfrshp) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_hr_rockfrshp
```

### Natural fragments {#sec-hr-nonrock}

Biologically-derived coarse materials can alter the soil's water holding capacity and other parameters just as rock fragments can, although in general they can themselves hold more water. They will also break down faster.

Where present in significant amounts (\> 5% by volume), record each type present and estimate volume and median fragment size in centimetres.

```{r}
#| label: tbl-hr-nonrock
#| tbl-cap: "Non-rock fragment types"

dat_hr_nonrock <-
  tribble(~Code, ~Name, ~Description,
         'Sh', 'Shell',    'Seashells or chitinous material',
         'Bo', 'Bone',     'Skeletal fragments',
         'Ti', 'Timber',   'Dead wood from trees and shrubs, coarse roots',
         'Ch', 'Charcoal', 'Burnt organic materials',
         'Gu', 'Gum',      'Kauri gum')

tbl_hr_nonrock <- gt(dat_hr_nonrock) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_hr_nonrock

```

### Artefacts {#sec-hr-artef}

Artefacts can both function like rock fragments and signify strong human influence on the soil profile.

Where present in significant amounts, for rapid assessment simply record total volumetric percentage of artefacts per horizon. Confirm absence as [0%]{.ceg}. For routine assessment, record each type present using @tbl-hr-artefacts and estimate percentage volume and median fragment size in centimetres.

Example: [Ra 5% 6-10 cm]{.ceg} for a horizon containing some chert flake tools.

For detailed assessment, for each fragment type, append any or all of the origin (@tbl-hr-aorg), fixation (@tbl-hr-afix), shape (@tbl-hr-ashp), and degradation (@tbl-hr-adeg) codes supplied below. Colour may also be described using common terms. Additional free-text notes and photos may also be useful.

Example: [Pl 10% 2-3 cm A SF F D]{.ceg} for a horizon containing some broken down remnants of horticultural film.

Artefact fragments smaller than 2 mm and non-solid soil contaminants are beyond the scope of this handbook.

```{r}
#| label: tbl-hr-artefacts
#| tbl-cap: "Artefact types"

dat_hr_artefs <-
  tribble(~Code, ~Name, ~Description,
         'As', 'Asphalt',  'Asphalt and other bitumen-containing materials',
         'Ba', 'Bone - altered', 'Human-altered bone fragments',
         'Br', 'Brick',    'Fired clay building materials, usually unglazed',
         'Cr', 'Ceramic',  'Fired clay objects, usually glazed',
         'Ce', 'Concrete', 'Concrete materials, cements',
         'Gl', 'Glass',    'Fragments from windows, containers, etc',
         'Pl', 'Plastics', 'Artificial polymers (non-textile)',
         'Ra', 'Rock - altered', 'Human-altered rock fragments',
         'Sa', 'Shell - altered', 'Human-altered shells',
         'Ta', 'Timber - altered', 'Human-altered timber',
         'Tn', 'Textiles - natural', 'Fabrics and carpets made from e.g. wool',
         'Tx', 'Textiles - artificial', 'Fabrics derived from or dominated by plastics',
         'Mt', 'Metals', 'Refined ore products',
         'Ms', 'Mine Spoil', 'Materials leftover from ore processing activities, excluding overburden',
         'Ws', 'Wastes', 'Other domestic or commercial rubbish')

tbl_hr_artefs <- gt(dat_hr_artefs) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_hr_artefs

```

```{r}
#| label: tbl-hr-aorg
#| tbl-cap: "Artefact origin (adapted from @weber2022)"

dat_hr_aorg <-
  tribble(~Code, ~Name, ~Description,
         'N', 'Not verifiable', 'Designation not possible / uncertain',
         'C', 'Consumer', 'Consumer-products (everyday-products) like tools, decorative items, etc', 
         'A', 'Agricultural', 'Agricultural products like straw bale nets, wire etc',
         'I', 'Industrial', 'Industrial products like broken pieces of industrial equipment, industrial waste',
         'B', 'Construction', 'All materials which were placed in soil for construction and supply purposes (e.g., pipes, power cables, geomembranes)')

tbl_hr_aorg <- gt(dat_hr_aorg) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_hr_aorg

```

```{r}
#| label: tbl-hr-afix
#| tbl-cap: "Fragment fixation (adapted from @weber2022)"

dat_hr_afix <-
  tribble(~Code, ~Name, ~Description,
         'NF', 'None',    'Artefacts are not included in soil aggregates, or soil is apedal',
         'SF', 'Solely',   'Some artefacts are incorporated into soil aggregates, but most are loose',
         'MF', 'Moderate', 'The majority of artefacts are incorporated into soil aggregates, but some pieces are loose',
         'CF', 'Complete', 'All artefacts are incorporated into soil aggregates')

tbl_hr_afix <- gt(dat_hr_afix) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_hr_afix

```

```{r}
#| label: tbl-hr-ashp
#| tbl-cap: "Artefact shape (adapted from @weber2022)"

dat_hr_ashp <-
  tribble(~Code, ~Name, ~Description,
         'P', 'Preserved', 'The original manufactured shape of the object is recognisable',  
         'B', 'Broken',  'The fragment is recognisable as having been part of a larger object',
         'F', 'Flat', 'Tabular or sheet-like',
         'A', 'Angular', 'Sharp-edged',
         'R', 'Rounded', 'Has edges that are not sharp',
         'S', 'Spherical', 'Ball shaped or ovoid')

tbl_hr_ashp <- gt(dat_hr_ashp) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_hr_ashp

```

```{r}
#| label: tbl-hr-adeg
#| tbl-cap: "Artefact degradation (adapted from @weber2022)"

dat_hr_adeg <-
  tribble(~Code, ~Name, ~Description,
         'N', 'Not verifiable', 'Designation not possible / uncertain',    
         'F', 'Fresh', 'Fresh artefacts with bright colours, unaffected shape, without signs of deterioration (cracks, broken edges, rough surfaces)', 
         'I', 'Incipient alteration', 'Artefacts with incipient deterioration surface such as first cracks, broken edges or roughened/grooved surfaces',
         'D', 'Degraded', 'Artefacts with clear indications of ageing such as pale colours or obvious discolouration, strong fragmentation, frayed edges',
         'S', 'Strongly degraded', 'Artefacts with strongly degraded surfaces, faded colours, soft surface, frayed areas or edges')

tbl_hr_adeg <- gt(dat_hr_adeg) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_hr_adeg

```

## Roots {#sec-root}

Plant roots affect water infiltration, bind and rework the soil as they grow, and exert strong local control over soil chemistry. Their position in the soil profile can provide information about ped development, particle packing, and the presence of pans. Accurate measurement of root volumes *in situ* is technically challenging [@measurin2012; @freschet2021], requiring extensive sampling to account for spatial, temporal and interspecies variability. This may be necessary for accurate assessment of total soil composition. The majority of the root mass, particularly fine roots of herbaceous species and grasses, will be found in the topsoil. For soil classification, presence of roots and their relative position within the soil are key characteristics.

Note that this section applies to the live root mass only.

### Root abundance {#sec-root-abund}

Estimation of root volume from cross-sectional analyses is not accurate, and so the following method, reproduced from @internationalorganizationforstandardizationa must be considered at best a relative assessment of root abundance.

Abundance is communicated as a count per square decimetre (100 cm^2^), undertaken on a pit or exposure face. For very thin horizons, a 50 cm horizontal linear transect at the horizon midpoint may be used instead. Multiple split-open cores may also be examined provided the minimum 100 cm^2^ area can be examined.

For rapid assessment, report the total number of roots. For routine assessment, report counts separately for the size classes in @tbl-hr-rsz. Always confirm absence of roots by recording a [0]{.ceg}.

E.g. [V 30, F 5, M 5, C 0]{.ceg} for a mixture of very fine, fine and medium roots.

With adequate size data, count per area can be converted to an areal percentage, e.g. 10 roots with diameter 5 mm in a 10 x 10 cm space gives a total area of $\pi × 0.25^{2} × 10 = {2\ cm^{2}}$, or 2%.

The count data may be classified using @tbl-hr-rtab.

```{r}
#| label: tbl-hr-rsz
#| tbl-cap: "Plant root size classes (adapted from @internationalorganizationforstandardizationa)"

dat_hr_rsz <-
  tribble(~Code,    ~Name,  ~Description,
            'V', 'Very Fine', '<0.5',
            'F',   'Fine',         '≥0.5–<2.0',
            'M', 'Medium',      '≥2.0–<5.0',
            'C', 'Coarse',        '≥5.0')

tbl_hr_rsz <- gt(dat_hr_rsz) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_label(Description = 'Size range (mm)') |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_hr_rsz

```

### Root position {#sec-root-pos}

Where roots are present in a horizon, record how the roots are spatially distributed using the codes in @tbl-hr-rpos. For rapid assessment, record the dominant position of all roots. For routine assessment, split by size. Root position cannot be reliably observed in augered samples, or in cores where structural units are known to be larger than the core diameter.

E.g. [F B, C T]{.ceg} for a soil where fine roots are confined to voids but coarse roots are not.

```{r}
#| label: tbl-hr-rpos
#| tbl-cap: "Plant root position"

dat_hr_rpos <-
  tribble(~Code,        ~Name,  ~Description,
            'W',     'Within', 'Roots are primarily located within peds',
            'B',    'Between', 'Roots are primarily located between peds, within fissures or along bedding planes',
            'T', 'Throughout', 'Roots are evenly distributed within and between peds',
            'U',    'Unknown', 'Sampling method did not allow reliable observation')

tbl_hr_rpos <- gt(dat_hr_rpos) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red', 
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_hr_rpos

```

### Other parameters {#sec-root-other}

Information about particular plant species contributing to the root mass and spatial organisation beyond the positional classes above (e.g. presence of mats) may prove useful in certain contexts. No coding system is provided for these; record as free-text notes.

## Voids and pores {#sec-voidpore}

Voids and pores are an integral part of soil structure, forming the spaces that occur between and within the aggregates described in @sec-hr-stru. They may contain air, water, or both, and play a key role in aeration, drainage, and root penetration. They are also habitat for soil meso- and microfauna. Horizon voids exclude open surface cracks, which are noted in @sec-pr-crack, and fine pores, which are described in @sec-pore. Infill features, where voids have previously been created but are no longer empty of solid material, are described in @sec-pr-inf. Voids and pores range in size from many centimetres down to nano-scale gaps in clay minerals. Only the larger end of the scale can be reasonably assessed in the field. As with measurement of roots, void space is most accurately determined in the laboratory.

### Voids {#sec-void}

Voids comprise pores \>2 mm wide. Dimensions of large, isolated voids like burrows can be measured directly and volume calculated. Areal percentage of smaller inter-ped spaces can be assessed across a cut face using the abundance charts in @sec-in-abundpics or by counting void openings. Use a minimum area of 100 cm^2^ for this assessment.

For rapid assessment, estimate the areal percentage of the horizon volume occupied by void space. Always confirm absence of voids by recording a [0]{.ceg}.

For detailed assessment, identify void types using @tbl-hr-voidtype and estimate the percentage of each type seen. Optionally, add connectivity using @tbl-hr-voidcon. Note that void connectivity may vary with soil water status.

Example: [10% IT C]{.ceg} for the void space in a strongly pedal, polyhedral topsoil.

```{r}
#| label: tbl-hr-voidtype
#| tbl-cap: "Type of soil void"

dat_hr_voidtype <-
  tribble(~Code, ~Name, ~Description,
          'BP', 'Bedding plane', 'A parting or discontinuity due to stratification during deposition of sedimentary particles.', 
          'CH', 'Channel', 'Tubular void of biological origin that is less than 20 mm in diameter (e.g. insect burrow)',
          'TU', 'Tunnel',  'Tubular void of biological origin that is 20 mm or more in diameter (e.g. rabbit burrow) ',
          'FS', 'Fissure', 'Planar void with a width much smaller than length and depth. Fissures represent the release of strain caused by drying and usually bound individual structural units and form repetitively.',
          'IT', 'Interstitial', 'Irregular voids occupying space between aggregates that do not fit together',
          )

tbl_hr_voidtype <- gt(dat_hr_voidtype) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red', 
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20)) |>
  sub_missing(missing_text = '')

tbl_hr_voidtype
```

```{r}
#| label: tbl-hr-voidcon
#| tbl-cap: "Connectivity of soil voids"

dat_hr_voidcon <-
  tribble(~Code, ~Name, ~Description,
          'I', 'Isolated', 'Voids are not connected to each other, and may only reach up to the soil surface',
          'P', 'Partly connected', 'Voids are sometimes connected to each other within the horizon, or continue into neighbouring horizons',
          'C', 'Continuous',  'Void spaces are well connected and form a continuous network')

tbl_hr_voidcon <- gt(dat_hr_voidcon) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20)) |>
  sub_missing(missing_text = '')

tbl_hr_voidcon
```

### Pores {#sec-pore}

Pores are open voids \<2 mm in diameter. Macropores are ≥0.75 mm, and micropores are \<0.75 mm. This division is based on the diameter above which pores are regarded as biologically generated. Pores can be observed on freshly broken ped faces or broken faces of a block cut from an apedal massive horizon, but will not be observable in apedal single-grain materials. Cut surfaces in moist or wet soils may smear, obscuring pore openings.

For routine assessment, note presence or absence of any pores visible without magnification. Ideally, make observations from at least 2-5 small samples from within a horizon.

For detailed assessment, count the number of macropores and micropores separately over a minimum area of 100 cm^2^. This may necessitate sampling several peds or small blocks from the target horizon. Counting micropores will require the use of a magnifier.

[size example diagram here]{.diags}

## Substrate-specific Features {#sec-reg}

Horizons that have experienced limited weathering may exhibit geogenic features that relate to their parent material and its mode of emplacement. These can affect water movement, slope stability and plant growth [@wysocki2005; @juilleret2016].

### Bedding {#sec-reg-bed}

Transported soil parent materials may comprise multiple thin layers of alternating texture or lithology. These layers can be unconsolidated (e.g. lamellae or cross-bedding in aeolian and fluvial sediments) partly or fully consolidated, and may subsequently have been compressed, folded and/or faulted. Where layers of subsolum material have been identified (@sec-pr-wdeg) and bedding is apparent within a layer, measure the median angle of the bedding off horizontal and note the degree of consolidation. For unconsolidated sediments, record the texture groups involved using @tbl-hr-earthgrps. If the bedding is consolidated, record the dominant lithologies (@sec-pm-lith) involved. Optionally, add a thickness range for the bedding in centimetres.

E.g. [25°, Sm/Ss, 10-30 cm]{.ceg} for thin, very firm marine sediments, or [2°, C/S, 0.1-0.5 cm]{.ceg} for clay-sand lamellae.

### Fractures {#sec-reg-fra}

Rock at the base of the profile may be significantly fractured by tectonic and weathering processes. Where lithic or paralithic layers (@sec-pr-wdeg) have been identified and fractures are apparent, note fracture type using @tbl-hr-fractyp, modal angle off horizontal in degrees, and median spacing in centimetres. For single fractures, record spacing as [0 cm]{.ceg}.

```{r}
#| label: tbl-hr-fractyp
#| tbl-cap: "Fracture types"

dat_hr_fractyp <-
  tribble(~Code, ~Name, ~Description,
          'F', 'Fault', 'Tectonically induced fractures that may cut across bedding. Bedding displacement will be apparent along the fault',
          'C', 'Crush', 'Fracture patterns induced by crushing and shearing forces. Fractures will have relatively close spacing and irregular orientation.',
          'P', 'Pressure', 'Fracture patterns induced by release of overburden pressure. Fractures are likely to be near-horizontal or parallel to the surface and may be curved.')

tbl_hr_fractyp <- gt(dat_hr_fractyp) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20)) |>
  sub_missing(missing_text = '')

tbl_hr_fractyp
```

[diagrams or photos - one for each type]{.ceg}
