---
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
library(dplyr)
library(tibble)
library(gt)
library(gtExtras)
```

# Soil architecture {#sec-hr-arch}

Soil architecture comprises presence and arrangement of fine earth, stones, roots, and other materials. The presence and connectivity of voids is also described.

## Structure {#sec-hr-stru}

Structure refers to the physical arrangement of the fine-earth fraction. Structure is determined by mineralogy, chemistry, the action of soil biota and land use history.

-   To assess the structure in an [E]{.ceg} or [P2]{.ceg} type profile (@tbl-pr-gear), insert a spade horizontally near the base of the horizon and push down on the handle, levering out a section of soil. Drop the spade blade with the soil, approximately 0.3 m onto a hard surface, e.g., the blade of another spade. Remove major roots that bind the mass together and again note the attributes of the newly produced solids. Note that this procedure will not necessarily disaggregate very moist or wet soils.
-   For smaller exposures ([P1]{.ceg}, [C1]{.ceg}, [C2]{.ceg}), remove a handful of soil from the extracted horizon and carefully disaggregate the soil mass to segregate all peds. Do not destroy peds to create fragments, or disaggregate existing apedal material. Note the attributes of the solids as instructed below. For smaller exposures, also use a spade to extract surface material so that at least the topmost horizon can be assessed in detail.
-   Assessing structure is generally impossible for [A]{.ceg} augered profiles, as the extraction method destroys it. However, the surface horizon should still be extracted with a spade and assessed if possible.

Sieving or direct measurement of solids may help estimate size. Weighing, or filling volumetric containers, may help estimate abundance of different classes of solids.

### Pedality {#sec-hr-stru-pedality}

Pedality refers to the degree of structural development evident within the horizon, independent of size or shape. Record pedality once per horizon using @tbl-hr-pedality.

```{r}
#| label: tbl-hr-pedality
#| tbl-cap: "Degree of pedality"
#| classes : no-stripe

dat_hr_pedality <-
  tribble(~Code, ~Name,  ~Description,
          'G', 'Single grain', 'Material is structureless and incoherent, with more than 85% by weight of discrete primary particles ranging in size from sand to very coarse gravel.',
          'V', 'Massive', 'Material is structureless but coherent, with no fissures at spacings of less than 200 mm',
          'W', 'Weak', 'Material where 15-25% of the soil comprises coherent peds',
          'M', 'Moderate', 'Material where 25-75% of the soil comprises coherent peds',
          'S', 'Strong', 'Material where >75% of the soil comprises coherent peds',
          'D', 'Disturbed', 'Material where ped structure has been altered or detroyed by disturbance, e.g. recent tillage or animal pugging')

tbl_hr_pedality <- gt(dat_hr_pedality) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_hr_pedality

```

### Ped shape {#sec-hr-stru-shp}

Where natural peds can be identified (pedality types [W]{.ceg}, [M]{.ceg}, [S]{.ceg}), describe their dominant shape using @tbl-hr-pedshp. Use @tbl-hr-clodshp for [D]{.ceg} pedality.

+------------+----------------+----------------------------------------------------------------------------------------------------------------------------------------+-----------------------------+
| Code       | Name           | Description                                                                                                                            | Orientation                 |
+:===========+:===============+:=======================================================================================================================================+=============================+
| [AB]{.ceg} | Angular blocky | -   Bounded by relatively flat smooth, roughly equal faces                                                                             | None                        |
|            |                | -   number of faces variable                                                                                                           |                             |
|            |                | -   most vertices angular                                                                                                              |                             |
|            |                | -   usually much accommodation to the faces of surrounding aggregates                                                                  |                             |
+------------+----------------+----------------------------------------------------------------------------------------------------------------------------------------+-----------------------------+
| [BL]{.ceg} | Blocky         | -   Bounded by undulating rough faces                                                                                                  | None                        |
|            |                | -   number of faces variable                                                                                                           |                             |
|            |                | -   many vertices rounded                                                                                                              |                             |
|            |                | -   limited accommodation to the faces of surrounding aggregates                                                                       |                             |
+------------+----------------+----------------------------------------------------------------------------------------------------------------------------------------+-----------------------------+
| [PH]{.ceg} | Polyhedral     | -   Bounded by relatively flat smooth, unequal faces                                                                                   | None                        |
|            |                | -   more than six faces; most vertices angular                                                                                         |                             |
|            |                | -   usually much accommodation to the faces of surrounding aggregates                                                                  |                             |
|            |                | -   re-entrant angles between adjoining faces - peds 'lock' together                                                                   |                             |
+------------+----------------+----------------------------------------------------------------------------------------------------------------------------------------+-----------------------------+
| [SR]{.ceg} | Spheroidal     | -   bounded by curved or very irregular faces;                                                                                         | None                        |
|            |                | -   limited accommodation to the faces of surrounding aggregates                                                                       |                             |
|            |                | -   sometimes with many visible pores                                                                                                  |                             |
+------------+----------------+----------------------------------------------------------------------------------------------------------------------------------------+-----------------------------+
| [LT]{.ceg} | Lenticular     | -   Bounded by curved faces                                                                                                            | Horizontal or shallow angle |
|            |                | -   overlapping, lens-shaped aggregates generally parallel to the soil surface that are thick at the centre and taper toward the edges |                             |
|            |                | -   usually much accommodation to the faces of surrounding aggregates                                                                  |                             |
+------------+----------------+----------------------------------------------------------------------------------------------------------------------------------------+-----------------------------+
| [PM]{.ceg} | Prismatic      | -   Bounded by relatively flat faces                                                                                                   | Vertical                    |
|            |                | -   vertically elongated units with angular vertices and flat tops                                                                     |                             |
|            |                | -   much accommodation to the faces of surrounding aggregates                                                                          |                             |
+------------+----------------+----------------------------------------------------------------------------------------------------------------------------------------+-----------------------------+
| [CO]{.ceg} | Columnar       | -   Bounded by relatively flat faces;                                                                                                  | Vertical                    |
|            |                | -   vertically elongated units with angular to rounded vertices and rounded (domed) tops                                               |                             |
+------------+----------------+----------------------------------------------------------------------------------------------------------------------------------------+-----------------------------+
| [PL]{.ceg} | Platy          | -   Bounded by relatively flat horizontal faces                                                                                        | Horizontal                  |
|            |                | -   much accommodation to the faces of surrounding aggregates                                                                          |                             |
+------------+----------------+----------------------------------------------------------------------------------------------------------------------------------------+-----------------------------+
| [LF]{.ceg} | Lentiform      | -   Bounded by curved faces                                                                                                            | Horizontal or shallow-angle |
|            |                | -   overlapping, lens-shaped aggregates generally parallel to the soil surface that are thick at the centre and taper toward the edges |                             |
|            |                | -   usually much accommodation to the faces of surrounding aggregates                                                                  |                             |
+------------+----------------+----------------------------------------------------------------------------------------------------------------------------------------+-----------------------------+
| [WL]{.ceg} | Wedge-like     | -   Bounded by flat faces;                                                                                                             | Horizontal                  |
|            |                | -   interlocking wedges or lenses that terminate in pronounced angular vertices;                                                       |                             |
|            |                | -   ends of vertices may be missing;                                                                                                   |                             |
|            |                | -   much accommodation to the faces of surrounding aggregates                                                                          |                             |
+------------+----------------+----------------------------------------------------------------------------------------------------------------------------------------+-----------------------------+

: Natural ped shapes and identifying characteristics (adapted from @milne1995, @iussworkinggroupwrb2022, and @nationalcommitteeonsoilandterrain2009) {#tbl-hr-pedshp cols-width=\[10,20,50,20\]}

[add diagrams]{.diags}

::: {#nte-nut .callout-note collapse="true" title="shorthand terms for soil structure"}
Some soil terms that don't seem to fit into the above schema are still in common use.

-   Crumb structure: a very fine structure seen in many silty to loamy horizons. The appearance is as of breadcrumbs, with 0.5-2mm rounded peds loosely to firmly packed, depending on depth and drainage. Larger weak subangular blocky structures in well-drained allophanic soils almost always break down into moderately or strongly developed crumb structure.
-   Nut structure: a larger version of crumb, usually on the order of 5-20 mm in diameter, often encased in a network of grass roots. Nut structures are more common in topsoils but can also occur in well-drained loamy subsoils.

Both structures are more correctly described as spheroidal with a characteristic size range of 1 ± 0.5 mm for crumb and 15 ± 10 mm for nut. Their use as a shorthand description is convenient, but the terms should not be used in databases because they conflate size and shape in a way that the modern schemas seek to avoid.
:::

+------------+-----------+------------------------------------------------------------+-------------+
| Code       | Name      | Description                                                | Orientation |
+============+===========+============================================================+=============+
| [CL]{.ceg} | Clod      | An artificially induced soil aggregate ≥ 100 mm in length  | Any         |
+------------+-----------+------------------------------------------------------------+-------------+
| [FM]{.ceg} | Fragment  | An artificially induced soil aggregate \< 100 mm in length | Any         |
+------------+-----------+------------------------------------------------------------+-------------+

: : Artificially induced pedality shapes and identifying characteristics {#tbl-hr-clodshp cols-width=\[10,20,50,20\]}

### Ped size {#sec-hr-stru-size}

Record the median length of the peds observed, along their longest axis. Record in mm with a range estimate, e.g. [25 ± 2 mm]{.ceg}

For rapid assessment, the dominant ped size may be recorded simply as [C]{.ceg} (coarse) or [F]{.ceg} (fine). The size cutoff between the two is set at 40 mm. See @tbl-in-size for a more detailed set of size classifications that may be applied to numeric size data.

#### Nesting peds {#sec-hr-stru-nest}

Peds are commonly nested, with large structures breaking down to smaller ones that may also be a different shape. For highly detailed descriptions, multiple sets of nesting peds may be described in terms of their development, shape and size.

For example, [S PR 80 ±  10]{.ceg} parting to [M BL 20 ±  2]{.ceg}.

# Roots and plant materials {#sec-rtpl}

## Roots {#sec-root}

Plant roots affect water infiltration and help define the functional horizon topsoil (see @sec-in-fh-ts). Their position in the soil profile can also provide information about ped development, particle packing, and the presence of pans. Note that this section does not apply to peats.

### Root abundance {#sec-root-abund}

Measure abundance by estimating the volumetric abundance of live roots in the visible soil horizon. For roots \< 2 mm it may be more practical to sample a cube \~100 mm on a side. When augering for rapid assessment, also use a spade to sample the topsoil.

Provide separate estimates for roots \< 2 mm, 2 - 10 mm and \> 10 mm. Record as whole percentages, confirming [0%]{.ceg} where no roots are observable in a given size range. The data may be classified using @tbl-hr-rtab.

```{r}
#| label: tbl-hr-rtab
#| tbl-cap: "Plant root abundance classes"
#| classes : no-stripe

dat_hr_rtab <-
  tribble(~Code,      ~Name,  ~Description,
              0,     'None',          '0%',
              1,      'Few',   '> 0 - 25%',
              2,   'Common',  '> 25 - 50%',
              3,     'Many',  '> 50 - 75%',
              4, 'Abundant',       '> 75%')

tbl_hr_rtab <- gt(dat_hr_rtab) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_hr_rtab

```

[practice change: Counting individual roots is impractical at both small and large size ranges. Switching to estimates of volumetric percentage. This brings roots into alignment with abundance assessment of other soil components (textures, stone, artefacts, voids), allowing total soil compositional data to be calculated.]{.majch}

### Root position {#sec-root-pos}

Where abundance is \> 0%, record how roots are distributed within the horizon using the codes in @tbl-hr-rpos. Position may not be reliably observable outside of pits or exposures, in which case record [U]{.ceg}.

```{r}
#| label: tbl-hr-rpos
#| tbl-cap: "Plant root position codes"
#| classes : no-stripe

dat_hr_rpos <-
  tribble(~Code,        ~Name,  ~Description,
            'W',     'Within', 'Roots are primarily located within peds.',
            'B',    'Between', 'Roots are primarily located between peds',
            'T', 'Throughout', 'Roots are spread throughout the soil mass.',
            'U',    'Unknown', 'Sampling method did not allow reliable observation.')

tbl_hr_rpos <- gt(dat_hr_rpos) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_hr_rpos

```

## Other parameters {#sec-root-other}

Notes about particular plant species contributing to the root mass and spatial organisation beyond the positional classes above (e.g. presence of mats) may prove useful in certain contexts. No coding system is provided for these.

## Timber and fibre {#sec-root-burt}

Where non-peat plant remains are encountered in significant volume (e.g. buried by volcanic or landslide activity), record as per @sec-hr-nonrock.

# Voids {#sec-void}

-   NB surface cracks already covered

## Pores {#sec-void-pore}

-   abundance
-   macro v micro sizes
-   inferred origin
-   direct and indirect observation for connectivity

## Other voids {#sec-void-other}

-   non-infilled space between coarse gravels (assessed via packing)
-   spaces between dry peds (where shrink-swell is significant)

```{r}
#| label: tbl-hr-cracktype
#| tbl-cap: "Crack persistence (WRB 2022)"

dat_hr_cracktype <-
  tribble(~Code, ~Name, ~Description,
          'NO', 'No cracks', NA, 
          'RT', 'Reversible cracks', 'Cracks that open and close with changing soil moisture',
          'IT', 'Irreversible cracks', 'Cracks that persist year-round')

tbl_hr_cracktype <- gt(dat_hr_cracktype) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(20), Name ~pct(20)) |>
  sub_missing(missing_text = '')

tbl_hr_cracktype
```

```{r}
#| label: tbl-hr-crackdepth
#| tbl-cap: "Crack depth (WRB 2022)"

dat_hr_crackdepth <-
  tribble(~Code, ~Name, ~Description,
          'AC', 'All', 'All cracks continue into the underlying layer', 
          'HC', 'Most', 'At least half, but not all of the cracks continue into the underlying layer',
          'SC', 'Some', 'At least one, but less than half of the cracks continue into the underlying layer',
          'NC', 'None', 'Cracks do not continue into the underlying layer')

tbl_hr_crackdepth <- gt(dat_hr_crackdepth) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(20), Name ~pct(20))

tbl_hr_crackdepth
```
