---
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
library(dplyr)
library(tibble)
library(gt)
```

# Secondary features {#sec-secondary}

Secondary soil features develop as a result of pedogenesis. Many create barriers to water movement or alter chemistry in a way that affects soil biota.

## Concentrations

These are bodies of pedogenic origin enriched in a particular constituent that are not associated with a particular surface within the soil. They may contrast sharply with the surrounding material in strength, composition or internal organisation, or differences with the surrounding material may be slight. Some are thin and sheetlike, others are blocky and some are irregular in shape. None are large or continuous enough to form a pan.

For rapid assessment, record concentration form and abundance % using @tbl-hr-concform.

For detailed assessment, note form, type, abundance in %, and size in mm. Colour is not required as it is usually determined by type and may have already been recorded as a colour pattern (@sec-col-patt). Abundance and size can be classified using @tbl-in-size and @tbl-in-abgen respectively.

Example: [NO F 10% 6-10 mm]{.ceg} for scattered small ironstone nodules.

```{r}
#| label: tbl-hr-concform
#| tbl-cap: "Concentration forms"

dat_hr_concform <-
  tribble(~Code, ~Name, ~Description,
          'CC', 'Concretion', 'hard concentrations with a layered internal structure, usually rounded',
          'NO', 'Nodule', 'hard concentrations with no internal structure, usually irregular',
          'CT', 'Crystal', 'A mineral arranged as interlocking or twinned macroscopic crystals or as a single macroscopic crystal, e.g., gypsum.',
          'SS', 'Soft masses', 'A non-cemented concentration that can seldom be removed from the soil as a discrete unit')

tbl_hr_concform <- gt(dat_hr_concform) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_hr_concform
```

```{r}
#| label: tbl-hr-conctype
#| tbl-cap: "Concentration types (adapted from @nationalcommitteeonsoilandterrain2023)"

dat_hr_conctype <-
  tribble(~Code, ~Name, ~Description,
          'E',  'Earthy', 'loamy or silty texture within a sandy or clayey matrix',
          'A',  'Argillaceous', 'clayey texture within a lower-clay matrix',
          'F',  'Ferruginious', 'Iron-dominated',
          'M',  'Manganiferous', 'Manganese-dominated',
          'N',  'Ferromanganiferous', 'Dominated by a mix of Fe and Mn',
          'K',  'Calcareous', 'Dominated by calcium carbonate (CaCO~3~)',
          'Y',  'Gypseous', 'Dominated by calcium sulfate (gypsum, CaSO~4~)',
          'Z',  'Salty', 'Dominated by salts more soluble than Calcium carbonate or calcium sulfate',
          'S',  'Sulfurous', 'Dominated by sulfur, e.g. jarosite in acid sulfate soils',
          'O',  'Organic', 'organic texture in a low-organics matrix',
          'UK', 'Unknown', 'composition cannot be reliably determined'
          )

tbl_hr_conctype <- gt(dat_hr_conctype) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  fmt_markdown(columns = 3) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_hr_conctype
```

## Pans {#sec-pan}

Pans form when particular chemicals precipitate out of the soil solution and cement a layer of mineral particles, or when particular particle size distributions and/or shape range become very tightly packed. Many pans are diagnostic in the NZSC.

Pans are described in terms of type, thickness, continuity, structure, strength and cementation. Pans comprise distinct horizons, so their thickness is defined using horizon boundaries. Their strength and cementation are assessed using @tbl-cn-str and @tbl-cn-ind. Codes for type, continuity and structure are given below. Note that continuity can only be observed in large pits or exposures.

For rapid assessment, record only the pan type using @tbl-hr-pantype. It is assumed that pans recorded this way are strongly developed enough to be classification-relevant. Note also that many (but not all) pans have dedicated horizon names, and in those cases these can be used for rapid assessment instead.

```{r}
#| label: tbl-hr-pantype
#| tbl-cap: "Types of Pan"

dat_hr_pantype <-
  tribble(~Code, ~Name, ~Description, ~Horizon,
          'C', 'Cultivation pan', 'Pan comprising compressed soil, induced by cultivation (e.g. use of heavy farm machinery)', '',
          'D', 'Densipan', 'Silty layers with a very high bulk density and extremly slow permeability', '\\*d (E horizons only)',
          'H', 'Humus-pan', 'pan cemented by aluminium and organic matter. Common in Podzols; sometimes known as \'coffee rock\'', 'Bhm',
          'F', 'Iron pan', 'interlocking nodules of Fe(/Mn) precipiates (bog iron), often at a texture contrast boundary', '',
          'K', 'Calcareous pan', 'pans created by the precipitation of calcium carbonate (CaCO~3~)', '\\*km (B or BC horizons)',
          'O', 'Ortstein pan', 'Pan cemented by iron and organic matter. Common in Podzols', 'Bsm',
          'P', 'Placic', 'a very thin (< 10 mm), dark, iron-dominated pan with sharp boundaries', 'Bfm',
          'Q', 'Silica pan', 'pans created by the precipitation of silica (SiO~2~)', '\\*qm (not A horizons)',
          'X', 'Fragipan', 'an earthy, high density, erosion-resistant layer common to Pallic soils', '\\*x (B or BC horizons)',
          'UK', 'Unknown', 'Pan origin cannot be clearly determined', ''
          )

tbl_hr_pantype <- gt(dat_hr_pantype) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  fmt_markdown(columns = c(3,4)) |>
  cols_width(Code ~pct(10), Name ~pct(20), Horizon ~pct(20))

tbl_hr_pantype
```


```{r}
#| label: tbl-hr-pancont
#| tbl-cap: "Pan continuity [@nationalcommitteeonsoilandterrain2023]"

dat_hr_pancont <-
  tribble(~Code, ~Name, ~Description,
          'C', 'Continuous', 'Extends as a layer with little or no break across 1 m or more.',
          'D', 'Discontinuous', 'Broken by cracks but original orientation of fragments is preserved.',
          'B', 'Broken', 'Broken by cracks and fragments are disoriented')

tbl_hr_pancont <- gt(dat_hr_pancont) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_hr_pancont
```


```{r}
#| label: tbl-hr-panstr
#| tbl-cap: "Pan structure (adapted from @nationalcommitteeonsoilandterrain2023)"

dat_hr_pancont <-
  tribble(~Code, ~Name, ~Description,
          'V', 'Massive', 'No recognisable structure',
          'S', 'Vesicular', 'Sponge-like structure having large pores, which may or may not be filled with softer
material',
          'C', 'Concretionary', 'Spheroidal concretions cemented together',
          'N', 'Nodular', 'Irregular nodules cemented together',
          'P', 'Platy', 'Plate-like units cemented together')

tbl_hr_pancont <- gt(dat_hr_pancont) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_hr_pancont
```

## Ped surface features

-   lamellae
-   clay coats
-   cutans and other infill types

## Stress features

-   pressure faces, slickensides, things of that nature

## Cryogenic features

-   for alpine areas and antarctic soils
