---
title: 'NZSC Family and Sibling'
output: html_document
editor_options: 
  chunk_output_type: console
---

Levels 4 and 5 of the NZSC, the Family and Sibling, provide useful summaries of key whole-profile parameters. 

## NZSC Family {#sec-in-nzscfam}

```{r}
library(dplyr)
library(tibble)
library(gt)
library(gtExtras)
```



### Soil Profile Material {#sec-fam-pm}

The soil profile material concept is a brief summary of the overall character of the profile, in particular its mineral and stone content and major erosional/depositional origin. Some information on the likely range of bulk density values is implicit.

@tbl-in-pmat below defines the available classes and should be used as a key. For Organic soils, skip the 'M' codes and start at [Sl]{.ceg}. Choose one code only per profile. Applying the key accurately requires familiarity with the major NZSC Diagnostic Criteria. Links to glossary definitions are provided below, but the source material [@hewitt2010] should also be consulted.

```{r}
#| label: tbl-in-pmat
#| tbl-cap: "Soil Profile Material"

dat_in_pmat <-
  tribble(~Code, ~Name, ~ Description,
          'Mf', 'Fragmental', 'Rock fragments with minimal fine earth from within 20 cm to > 60 cm depth',
          'Ml' ,'Lithic', 'Other soils with a coherent- or shattered [lithic contact](LINKME) within 45 cm',
          'Mp', 'Paralithic', 'Other soils with a [paralithic contact](LINKME) within 45 cm',
          'Mt', 'Tephric', 'Other soils with at least 30 cm of [tephric soil material](LINKME) within 100 cm',
          'Mr', 'Rounded-stony', 'Other soils with ≥ 35% by volume [rounded or sub-rounded](LINKME) rock fragments from within 45 cm to > 100 cm.',
          'Ma', 'Angular-stony', 'Other soils with ≥ 35% by volume [angular or sub-angular](LINKME) rock fragments from within 45 cm to > 100 cm.',
          'Mm', 'Moderately deep on rock', 'Other soils with a lithic or paralithic contact at < 100 cm.',
          'Mg', 'Moderately deep on rock fragments', 'Other soils with ≥ 35% by volume rounded or sub-rounded rock fragments from within 45-90 cm, extending continuously to 100 cm with a cumulative thickness of ≥ 50 cm (unless over rock).',
          'Md', 'Stoneless', 'Other soils with weighted average < 3% rock fragments by volume in the top 100 cm.',
          'Ms', 'Stony', 'Other mineral soils.',
          'Sl', 'Organic-lithic', 'Organic soils with a coherent- or shattered lithic contact within 45 cm.',
          'Sp', 'Organic-paralithic', 'Other organic soils with a paralithic contact within 45 cm.',
          'Sd', 'Organic-deep', 'Other organic soils with continuous peat throughout the top 100 cm.',
          'So', 'Organic-other', 'Other organic soils.')

tbl_in_pmat <- gt(dat_in_pmat) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(20), Name ~pct(20))

tbl_in_pmat

```

Profile material class places restrictions on other parts of the family and sibling definition. These are discussed as they arise below.

### Rock class {#sec-fam-rcl}

```{r}
#| label: tbl-ss-pm
#| tbl-cap: "NZSC Lithology"

dat_ss_pm <-
  tribble(~Code, ~Name,
    'Cb',	'Carbonaceous Rock (Coal)',
    'Sq',	'Soft Quartzitic Sediments',
    'Hq',	'Hard Quartzitic Sediments',
    'Sm',	'Soft mudstone',
    'Ss',	'Soft sandstone',
    'Hm',	'Hard mudstone',
    'Hs',	'Hard sandstone',
    'Tm',	'Tuffaceous mudstone',
    'Ts',	'Tuffaceous sandstone',
    'Sc',	'Schist',
    'Ls',	'Soft calcareous',
    'Li',	'Hard calcareous',
    'Ma',	'Marble',
    'Rh',	'Rhyolite',
    'Ig',	'Ignimbrite',
    'An',	'Andesite',
    'Ba',	'Basalt',
    'Gr',	'Granite and Gneiss',
    'Di',	'Diorite',
    'Ga',	'Gabbro',
    'Fe',	'Iron Sands')
      
tbl_ss_pm <- gt(dat_ss_pm) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(20))

tbl_ss_pm

```

::: callout-note
'Soft' rocks in this context can at most be broken readily with a light hammer blow, and can be scraped and peeled by a pocket knife. 'Hard' rocks can at best be scratched with a knife but not scraped or peeled, and require a firm hammer blow to crack [@lynn1991, p. 11].
:::

#### Rock class of stone and bedrock {#sec-fam-rclr}

Rock class of in-profile stone and of bedrock (where encountered) uses the lithology codes in @tbl-ss-pm, with specific restrictions around how they are applied to the whole profile. These are:

-   [Md]{.ceg} soils effectively have no stone or bedrock. Leave the property undefined.
-   Rock class is only defined in reference to stones \> 2 mm and bedrock (where encountered within 100 cm). The rock class of the fine earth fraction is accounted for below.
-   The 'deep and stoneless' classes ([Mt]{.ceg}, [Md]{.ceg}, [So]{.ceg} and [Sd]{.ceg}) can still have a total of \< 3% by volume stones in the top 100 cm.
-   Include pumice in rock class only where pumice strength is more than 'extremely weak' [@lynn1991, p. 10] - i.e., it can only be broken by hand with difficulty. If the pumice's lithology is uncertain, use the dominant lithology of the probable volcano of origin (e.g. [Rh]{.ceg} in the Taupō Volcanic Centre and [An]{.ceg} around Taranaki).
-   Where in-profile stone lithology contrasts with bedrock lithology, two codes separated by '[/]{.ceg}' can be used e.g. [Sc/Gw]{.ceg}.
-   Where two lithologies are close to co-dominant in a stony profile, two codes separated by '[+]{.ceg}' can be used e.g. [An+Rh]{.ceg}.
-   More complicated combinations are not allowed, e.g. no [An+Rh/Gw]{.ceg}.

#### Rock class of fines {#sec-fam-rclf}

The same rock class codes in @tbl-ss-pm can be applied to report the dominant lithology of the fine earth (\< 2 mm) fraction of the full profile. Again, some restrictions apply:

-   [Sd]{.ceg} soils and [Mf]{.ceg} soils effectively have no fine earth fraction. Leave the property undefined.
-   Where a lithological discontinuity exists in the profile, two codes separated by '[/]{.ceg}' can be used e.g. [Sc/Gw]{.ceg}.
-   Where two lithologies are close to co-dominant, two codes separated by '[+]{.ceg}' can be used e.g. [An+Rh]{.ceg}.
-   More complicated combinations are not allowed, e.g. no [An+Rh/Sm+Ss/Gw]{.ceg}.

### Family texture class {#sec-fam-tex}

The family texture class describes the dominant character of the top 60 cm, or to the profile material class-defining rock or gravel layer contact if that is shallower. Use the codes in @tbl-in-famtex. Note that 18% organic carbon corresponds to approximately 30% organic matter.

```{r}
#| label: tbl-in-famtex
#| tbl-cap: "Family texture classes"
#| classes : no-stripe

dat_in_famtex <-
  tribble(~Code, ~Name, ~Orgc, ~Silt, ~Clay, 
         'c', 'Clayey', '< 18', NA, '≥ 35', 
         's', 'Sandy', '< 18', '< 40', '< 8',
         'l', 'Loamy', '< 18', '< 40', '≥ 8 - < 35',
         'z', 'Silty' , '< 18', '≥ 40', '≥ 8 - < 35',
         'p', 'Peaty', '≥ 18', NA, NA)

tbl_in_famtex <- gt(dat_in_famtex) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  fmt_missing(missing_text = '-') |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_in_famtex

```

### Permeability class {#sec-fam-perm}

This class estimates the permeability of the top 100 cm in units of mm/hr. Use the codes in @tbl-in-famperm and follow the rules in @fig-in-famperm to assign permeability class.

```{r}
#| label: tbl-in-famperm
#| tbl-cap: "Family permeability classes"
#| classes : no-stripe

dat_in_sibdepth <-
  tribble(~Code, ~Name, ~Description, 
         's', 'Slow', '< 4 mm/hr', 
         'm', 'Moderate', '≥ 4 - < 72 mm/hr',
         'r', 'Rapid', '≥ 72 mm/hr')

tbl_in_sibdepth <- gt(dat_in_sibdepth) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(20), Name ~ pct(40))
  
tbl_in_sibdepth

```

Follow the flowchart to assign family permeability class:

```{mermaid}
%%| label: fig-in-famperm
%%| fig-cap: "NZSC Family permeability flowchart"

%%{ init: { 'flowchart': { 'curve': 'stepAfter' } } }%%
graph TD
  A[Identify the uppermost horizon with the slowest permeability,<br />or rock above 100 cm. This is Horizon X.]-->B[Does Horizon X have > 30 cm of<br /> contrasting permeability above?]
  B-->|Yes|C(Use m/s, r/s, r/m)
  B--> |No|D["Does the profile have a horizon ≥ 20 cm thick"<br /> below Horizon X with a different permeability class?]
  D-->|Yes|E(Use s/m, s/r or m/r)
  D-->|No|F(Use s, m, or r)
  
```

Some additional rules apply for particular soil profile material classes:

-   For soils with an [S\*]{.ceg} profile material class, assess family permeability class from 20-100 cm or to rock (i.e. ignore the surface 20 cm).
-   Soils with profile material class [Ml]{.ceg} or [Mp]{.ceg} and any soil with an [M]{.ceg} root barrier must use [m/s]{.ceg} or [r/s]{.ceg}.
    -   Exception: [Ml]{.ceg} with [F]{.ceg} barrier can't be [\*/s]{.ceg}.

### Parent material origin {#sec-fam-pmo}

Parent material origin (PMO) is the final defining factor of the soil Family. It describes how soil parent materials have arrived at the point of observation. Allowable codes are listed in @tbl-ss-pmo.

Multiple modes may have operated in different parts of the soil profile. Up to two may be recorded at Family level, in order of their occurrence down-profile (e.g. [Tp/Fl]{.ceg}). If PMO has been recorded at horizon level and more than two have been identified, the dominant two must be chosen on the basis of their cumulative thickness. Cumulative thicknesses \<20 cm should be ignored unless the layers include a significant pedological feature (e.g. a pan).

```{r}
#| label: tbl-ss-pmo
#| tbl-cap: "Parent material origin"

dat_ss_pmo <-
  tribble(~Code, ~Name, ~Description,
          'An', 'Anthropic', 'Deposits made by the direct actions of humans, including truncation, mixing, or deposition.',
          'Cl', 'Colluvium', 'Weathered soil and rock material mantling slopes that has been transported primarily by gravity and sheet wash.',
          'Fl', 'Alluvium', 'Sediments that have been deposited by streams, rivers and other running water.',
          'Gl', 'Glacial till', 'Poorly stratified, poorly sorted rock fragments, sand and mud, surface or near-surface deposits resulting from the transportation by and deposition from ice or meltwater from beneath or in close proximity to glacial ice.',
          'Lc', 'Lacustrine', 'Formed in and around lake beds; comprising extremely fine sediment deposited under very low-flow freshwater conditions.',
          'Mr', 'Marine' , 'Unconsolidated sediments saturated by brackish or saline water.',
          'Lh', 'Lahar', 'A flow of volcanic material, both ash and coarser products, mixed with water; often caused by the spilling over of a crater lake. Also used to describe the mounds formed on volcanic ring-plains where lahars stop.',
          'Tp', 'Tephra', 'A general term for all solid (rather than molten) materials ejected from a volcano during an eruption: includes boulders, lapilli and ash.',
          'Li', 'Litter', 'Partly or fully decomposed forest litter, equal to or greater than 30 cm depth that is **not** almost continually saturated in the natural state (**F** and **H** horizons).',
          'Pt', 'Peat', 'Partly decomposed plant remains in a water-saturated environment, such as a bog',
          'Lo', 'Loess', 'A blanket deposit of silt-sized particles (0.002--0.06 mm diameter); usually carried by wind from dry riverbeds or outwash plains during glacial and post-glacial periods.',
          'Sa', 'Aeolian sand', 'Wind-deposited sand-sized particles (0.05--2.0 mm diameter), i.e. dune sand.',
          'Rw', 'Highly Weathered Rock', 'Highly to completely weathered rock, formed in place by chemical weathering.',
          'Rx', 'Rock', '*In situ* unweathered to moderately weathered lithified geologic material comprising one or more minerals showing rock structure.',
          'Uk', 'Unknown', 'Origin cannot be determined with certainty.'
          )

tbl_ss_pmo <- gt(dat_ss_pmo) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(20), Name ~pct(20))

tbl_ss_pmo
```

## NZSC Sibling {#sec-in-sib}

These codes help define level 5 of the NZSC, the soil sibling.

### Soil Depth {#sec-in-sibdepth}

Soil depth from a sibling perspective is the depth to where digging becomes difficult, e.g.

-   A horizon with ≥ 35% stones by volume (V or X functional horizons, @sec-in-fh)
-   A soft- or hard-rock surface (M rooting barrier)
-   A shattered-lithic contact (F rooting barrier)
-   A pan (Q functional horizons, @sec-)
-   Firm strength with massive or coarse structure (\*Cf functional horizons)

Use the codes in @tbl-in-sibdepth to assign a depth class to a profile.

```{r}
#| label: tbl-in-sibdepth
#| tbl-cap: "Sibling depth classes"
#| classes : no-stripe

dat_in_sibdepth <-
  tribble(~Code, ~Name, ~Description, 
         'd', 'Deep', '> 100 cm', 
         'md', 'Moderately deep', '≥ 45 - < 100 cm',
         's', 'Shallow', '≥ 20 - < 45 cm',
         'vs', 'Very Shallow' , '< 20 cm')

tbl_in_sibdepth <- gt(dat_in_sibdepth) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(20), Name ~ pct(40))
  
tbl_in_sibdepth

```

### Topsoil stoniness

Topsoil stoniness assesses the amount of rock fragments (as % by volume) in the top 20 cm of the soil profile, including those resting on the soil surface.

A weighted average calculation is required if the top 20 cm contains more than one horizon. As an example,

```{r}
#| label: tbl-in-sibdcalc
#| classes : no-stripe

dat_in_sibdcalc <-
  tribble(~hthick, ~relthick, ~prop, ~stonepc, ~wavgstone,
          12, 12, 0.6, 15, 9,
          20, 8, 0.4, 25,10,
          NA, NA, NA, NA, 19)

tbl_in_sibdcalc <- gt(dat_in_sibdcalc) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  cols_label(hthick ~ "Horizon Thickness (cm)",
             relthick ~ "Relevant thickness (cm)",
			       prop ~ "Horizon proportion of top 20 cm",
             stonepc ~ "Horizon stone (%)",
             wavgstone ~ "Weighted average topsoil stone (%)" ) |>
  tab_style(style = list(cell_text(weight = 'bold')), 
                         locations = cells_body(rows = 3, columns = 5)) |>
  sub_missing(missing_text = '')
  
tbl_in_sibdcalc

```

Use the codes in @tbl-in-sibstone to assign stoniness.

```{r}
#| label: tbl-in-sibstone
#| tbl-cap: "Sibling topsoil stoniness classes"
#| classes : no-stripe

dat_in_sibstone <-
  tribble(~Code, ~Name, ~Description, 
         '1', 'Stoneless', '< 1 %', 
         '2', 'Slightly stony', '≥ 1 - < 5 %',
         '3', 'Moderately stony', '≥ 5 - < 35 %',
         '4', 'Very stony' , '≥ 35 % ')

tbl_in_sibstone <- gt(dat_in_sibstone) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(20), Name ~ pct(40))
  
tbl_in_sibstone
```

### Sibling texture classes

These classes are assessed over the top 100 cm, or to rock or gravel layer contact if those are shallower. Note that this is deeper than the family-level texture control section. The following rules apply:

-   [Mf]{.ceg}, [Mr]{.ceg}, [Ma]{.ceg}, [Ml]{.ceg}, and [Mp]{.ceg} soils must match their family texture code ([c]{.ceg}, [s]{.ceg}, [l]{.ceg}, or [z]{.ceg}).
-   [Sl]{.ceg} and [Sp]{.ceg} soils must use one of [Tp]{.ceg}, [Tc]{.ceg}, [Tl]{.ceg} or [Ts]{.ceg} to align with their family texture code of [p]{.ceg}.
-   [Sd]{.ceg} soils must use [Tp]{.ceg}.
-   Upper and lower textures can be defined for other soil profile classes, but only the dominant two are recorded. Both contributing layers must be at least 20 cm thick.
-   When there are more than two texture layers within the control section, then the texture profile is identified according to the uppermost texture contrast.
    -   For example, if a sandy horizon occurs at 80 cm depth and a clayey horizon overlays a silty horizon at 40 cm, then the texture contrast is reported as clayey over silty ([c/z]{.ceg}) and the sandy layer is ignored.
-   Silty and loamy horizons are not considered contrasting and can be added together where contiguous.
    -   For example, a profile with silty (0--15 cm) over loamy (15--25 cm) over sandy (25-90 cm) is considered to have a contrasting upper layer of 25 cm, and is identified as silty over sandy ([z/s]{.ceg}). The texture specification uses the uppermost of the silty or loamy layers together with the lower layer.
-   Skeletal horizons ([k]{.ceg}) may only be used for [Ms]{.ceg}, [Mm]{.ceg}, [Mt]{.ceg} and [So]{.ceg} soils. The skeletal material must be non-tephric.
-   Skeletal horizons in moderately deep gravelly soils ([Mg]{.ceg}) are ignored in favour of describing the overlying fine sediments.

Use the codes in @tbl-in-sibtexture to assign sibling level texture class(es).

```{r}
#| label: tbl-in-sibtexture
#| tbl-cap: "Sibling texture classes"
#| classes : no-stripe

dat_in_sibtexture <-
  tribble(~Code, ~Name, ~Org, ~Silt, ~Clay, ~Stone, 
           'k', 'Skeletal',                  NA,     NA,            NA, '≥ 35',
          'Tp', 'Peat or litter',        '≥ 30',     NA,            NA, '< 35',
          'Tc', 'Clayey peat',    '≥ 18 - < 30',     NA,        '≥ 35', '< 35',
          'Tl', 'Loamy peat',     '≥ 18 - < 30',     NA,  '≥ 8 - < 35', '< 35',
          'Ts', 'Sandy peat',     '≥ 18 - < 30', '< 40',         '< 8', '< 35', 
          'c',  'Clayey',                '< 18',     NA,        '≥ 35', '< 35',
          's',  'Sandy',                 '< 18', '< 40',         '< 8', '< 35',
          'l',  'Loamy',                 '< 18', '< 40',  '≥ 8 - < 35', '< 35', 
          'z',  'Silty',                 '< 18', '≥ 40',         '< 3', '< 35') 

tbl_in_sibtexture <- gt(dat_in_sibtexture) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_label(Org ~ 'Organic Carbon %',
             Silt ~ 'Silt %',
             Clay ~ 'Clay %', 
             Stone ~ 'Stone %')|>
  sub_missing(missing_text = '') |>
  cols_width(Code ~pct(10), Name ~ pct(30))
  
tbl_in_sibtexture
```

### Drainage {#sec-in-pdrng}

Profile drainage is determined from horizon depths and per-horizon drainage ratings completed according to @sec-in-hdrng.

```{r}
#| label: tbl-in-pdrng
#| tbl-cap: "Profile drainage rules"

dat_in_pdrng <-
  tribble(   
    ~Code,      ~Name,              ~Description,
    'VP', 'Very Poor', 'Profiles with very poorly drained horizons starting <10cm from the land surface',
    'PO',    'Poor', 'Profiles with poorly drained horizons starting <30cm from the land surface',
    'IP',  'Imperfect', 'Profiles with poorly drained horizons starting 30-60cm from the land surface, *or*
    Profiles with imperfectly drained horizons starting <30cm from the land surface',
    'MW', 'Moderately Well', 'Profiles with moderately well-drained horizons starting <90cm from the surface, *or*
    Profiles with poorly-drained horizons starting 60-90cm from the surface, *or*
    Profiles with imperfectly drained horizons starting 30-90cm from the surface', 
    'WE', 'Well', 'All horizons starting within 90cm of the surface are well-drained.'  )

tbl_in_pdrng <- gt(dat_in_pdrng) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(weight = 'bold', color = 'red')),
            locations = cells_body(columns = 1)) |>
  fmt_markdown(columns = 3)

tbl_in_pdrng

```

Workers classifying drainage for NZSC Sibling construction have no need to subdivide within those profile ratings. For other types of work e.g. agricultural suitability assessment, it may be useful to recognise subgroups – e.g. for a deep-rooted, wetness-intolerant crop, there’s a difference between a profile that is at its *worst* imperfectly drained in the top 1 m, vs one that is dominantly imperfectly drained but has poor drainage below 60 cm.

#### Helper information for profile drainage {#sec-in-pdrng-xtra}

@tbl-in-drngl describes some potential additional signifiers of profile drainage condition that might be considered. These may be useful when estimating drainage in a new landscape that has not yet been examined in detail.

```{r}
#| label: tbl-in-drngl
#| tbl-cap: "Profile drainage signifiers"
#| classes : no-stripe
#| column: page

dat_in_drngl <-
  tribble(~Code, ~Name, ~relel, ~Landform, ~Slope, ~Veg, ~Other, 
         'VP', 'Very poor', 'Very low (0-1m)', 'Waning/minimal/enclosed flats and depressions', 'Level to very gentle', 'Exclusively wetland plants', 'Persistent surface water ponding or seeps nearby',
         'PO', 'Poor', 'Low (0-10m)', 'Waning/minimal slopes and flats', 'Low-slope or level ground', 'Wetland plants', 'Seasonal ponding or seeps nearby',
         'IP', 'Imperfect', 'Variable', 'Slopes, hollows, flats', 'Gentle to moderately steep', 'Any', 'More common below high dropoffs',
         'MW', 'Moderately Well', 'Variable', 'Most landforms', 'Gentle to moderately steep', 'Any' ,'',
         'WE', 'Well', 'Moderate to high (>10 m)', 'crests, ridges and spurs', 'Gentle to very steep' , 'Any, dry-tolerant may outcompete', '')

tbl_in_drngl <- gt(dat_in_drngl) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_label(relel ~ 'Elevation above local drainage',
             Veg ~ 'Vegetation') |>
  cols_width(Code ~pct(10), Name ~ pct(20))
  
tbl_in_drngl

```

#### Natural vs Artificial drainage {#sec-in-pdrng-alt}

Profile drainage is often modified by human activity. Many of these modifications require ongoing maintenance. Use the codes in @tbl-in-natdrng to signify the degree of drainage modification.

```{r}
#| label: tbl-in-natdrng
#| tbl-cap: "Profile drainage modification level"
#| classes : no-stripe

dat_in_natdrng <-
  tribble(~Code, ~Name, ~Description, 
         'N', 'Natural', 'No evidence of human impacts on the local drainage regime', 
         'M', 'Minor', 'Drainage regime passively altered by e.g. conversion to or from forest; or ambiguous evidence of any human impacts beyond clearing',
         'A', 'Major', 'Clear evidence of directed human alteration to the drainage regime e.g. drain or dam construction adjacent to site, flow diversions, infill, levelling or contouring works')

tbl_in_natdrng <- gt(dat_in_natdrng) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(20), Name ~ pct(20))
  
tbl_in_natdrng

```
