---
title: "Horizon Properties"
editor: visual
---

```{r}
library(dplyr)
library(tibble)
library(gt)
library(gtExtras)
```

## Permeability ratings

-   as currently used - import from S-Map manual

## Drainage

Soil drainage is an assessment of how fast water leaves a soil profile relative to supply. This characteristic can be controlled by morphological characteristics that occur in a range of combinations.

Drainage should first be assessed on a per-horizon basis. These interpretations are later summarised at a profile level (see @sec-in-pdrng). @tbl-in-intdrng below summarises the common characteristics associated with each horizon drainage class. A 'balance of evidence' approach should be taken when assigning a drainage class to a horizon. When in doubt, choose the poorest class available.

```{r}
#| label: tbl-in-intdrng
#| tbl-cap: "Horizon-level drainage characteristics"
#| column: page

dat_in_intdrng <-
  tribble(~Code, ~Name, ~HNames, ~sws, ~mcol, ~Mottles, ~Texture, ~StrGr, ~secfts, ~cons,
          'VP', 'Very poor', 'O*, *r', 'Saturated or wet', 'Dark or gley', '<2%', 'Peaty, clayey, silty. Rarely sandy, very rarely stony', 'Apedal', 'Few to None', 'Deformable', 
          'PO', 'Poor', '*r, *g', 'Wet', 'Dark, gley, grey, pale', '2-50%', 'Clayey, silty, occasionally peaty and more often just highly organic (e.g. drained ex-peat). Less comonly sandy or stony', 'Apedal to weak', 'Few to none', 'Deformable or semi-deformable',
          'IP', 'Imperfect', '*(g)', 'Wet to Moist', 'Dark, grey, yellow or pale', '25-75%', 'Any non-peaty',  'Apedal to moderate', 'Fe and/or Mn-rich segregations, nodules and pans', 'Friable to semi-deformable',
          'MW', 'Moderately Well', '*(f)', 'Wet to Dry', 'Any non-gley', '2-50% (non-gley)', 'Any', 'Any', 'Mottles more often in ped interiors', 'Very friable to brittle', 
          'WE', 'Well', 'No redox suffixes', 'Moist to Dry', 'Any non-gley', 'None', 'Variable but usually not clayey', 'Any', '<2% (non-gley)', 'Very friable to brittle')

tbl_in_intdrng <- gt(dat_in_intdrng) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  tab_style(style = list(cell_text(weight = 'bold')),
            locations = cells_body(columns = 2)) |>
  cols_label(HNames = 'Horizon Names',
             sws    = 'Typical Moisture',
             mcol   = 'Matrix colour',
             StrGr  = 'Structure Grade',
             secfts = 'Secondary features',
             cons   = 'Consistence')

tbl_in_intdrng
```

### Classifying topsoil drainage

Morphological features associated with drainage status can be hard to spot in A horizons, mostly due to their darker colour. **When otherwise in doubt**, a well developed A horizon should be assigned a drainage class one better than the horizon below it (usually a B, E, or C horizon). A poorly developed A horizon should receive the same class assignment as the horizon below. Use the 'distinct topsoil' NZSC diagnostic criteria to identify well developed A horizons. 

The reasoning behind this rule of thumb is that the lack of topsoil development may be partially explained by lack of contrast in the drainage regime between the topsoil and the layer below.

<!-- example diagram here --->

::: callout-note
Soils that are persistently wet may also become anoxic. This is more common in low-permeability soils, where slow percolation allows enough time for oxygen to be consumed entirely. 
<!--- add more --->
:::
