# Reference data

```{r}
library(dplyr)
library(tibble)
library(gt)
library(gtExtras)
```

## Naming/numbering sites and site collections

-   tips for robust systems for projects of varying scale
-   tips for working across multiple teams
-   distinguishing site IDs from database IDs (e.g. you might want a UUID but you don't want to be writing one out by hand)
-   tagging by eg. transect membership

## Date/Time

See @sec-ss-dt for recording requirements.

## Authors

-   pick one responsible officer for the description, note the others
-   note: privacy law - don't put this data in public-facing downloads without anonymising

## Exposure type

The method by which the soil was exposed for observation affects which parameters can be recorded, or at least recorded reliably.

```{r}
#| label: tbl-pr-gear
#| tbl-cap: "Method of profile exposure"

dat_pr_gear <-
  tribble(~Code, ~Name,
          'P1', 'Quick pit – up to 0.5 m wide and generally no deeper than 1m',
          'P2', 'Full pit > 1 m wide, exposing the complete profile to regolith or bedrock',
          'E', 'Exposure – road cutting or similar, chipped back and cleaned',
          'C1', 'Relatively undisturbed soil core, < 50 mm diameter',
          'C2', 'Relatively undisturbed soil core, ≥ 50 mm diameter',
          'A', 'Auger sample (hand or mechanical, including e.g. post hole diggers)')

tbl_pr_gear <- gt(dat_pr_gear) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(20))

tbl_pr_gear
```

-   [**A**]{style="color: red"} samples generally cannot be used to determine soil structure due to the degree of disturbance.
-   Combined sampling is common in field mapping, e.g. small pit excavation to \~0.50-0.75 m and augering to a required depth in relatively well-understood landscapes. Record these as [**P1**]{style="color: red"}.
-   Gouge-style augers used in soft marine sediments may be included in [**C2**]{style="color: red"} even though they commonly narrow to \< 50 mm at their maximum depth.
-   [**C1**]{style="color: red"} style samplers are common in paddock-scale monitoring and research, where they are used for rapid surface measurement. These are not generally suitable for mapping as they cannot sample below the topsoil.

[add some gear pics/diagrams here]{style="background-color: lightgreen"}
