# The soil site {#sec-site}

```{r}
library(dplyr)
library(tibble)
library(gt)
```

## Defining the site {#sec-pr-site}

Soil profiles are described at sites. The 'site' concept is necessarily somewhat flexible to account for the range of sampling purposes and methods. Generally speaking, it is considered to include the immediate surroundings of a profile description, up to a radius of \~10 m. As such, it will generally nest within a landform (@sec-gmph-lfm) unless the site selection method (@sec-ref-sampstrat) was prescriptive and landform-blind.

Larger sampling plots or long transects that cross multiple landforms (or even landscapes) should be treated as a series of related sites, rather than a single site.

The soil profile description starts with an observation of the site's surface and works downwards towards the parent material. The chief tasks are assessing the surface, delineating pedologic horizons (and any additional geogenic layers, if present) and describing their morphological character.

Before beginning a soil profile description, it is essential to record some additional reference data.

## Reference data {#sec-ref}

Reference data is metadata that clearly identifies a soil profile description. These data enable the description to be linked with other profiles, samples, analytical results and photographs. Reference data includes:

-   an identifier
-   who completed the description
-   when and where it was located
-   why the description was needed
-   how the location was chosen
-   what equipment was used
-   depth of observation

These data are discussed in turn below.

### Site identifier {#sec-ref-id}

Assign each profile description an identifier that is at least unique at a whole-of-project level, and preferably at a whole-of-organisation level. There are many ways to build identifier systems (see @nte-uuid), but a good system usually has a predictable structure, is open-ended, and is easy to remember. Incorporating location data into the identifier can help avoid confusion between multiple projects within an organisation.

e.g. a region code followed by a sequential number: [WLG_001]{.ceg}.

This kind of identifier is 'semi-unique' in that there is still some potential for confusion if the data is combined with datasets from other sources - for example, a vegetation survey group working independently in the same region may have decided to use the same naming protocol for their sites.

::: {#nte-uuid .callout-note collapse="true" appearance="minimal" title="Unique identifiers"}
Profile identifier formats are often a compromise between robustness and simplicity. Well-designed databases use truly unique identifiers for every piece of data they hold. Under contemporary best practice, these are usually 'UUIDs', 128-bit hexadecimal codes (see @davis2024). Although technically robust, these are difficult to work with day-to-day, and most workers will find themselves preferring to interact with simpler semi-unique identifiers. In a soil profile database, semi-unique identifiers should be implemented as UUID aliases. End users should primarily interact with the aliases, and the UUIDs should remain hidden.
:::

::: {#nte-teamsites .callout-note collapse="true" appearance="minimal" title="Working in teams"}
Semi-unique identifiers require active management when working with multiple field teams operating in parallel. This process avoids the risk of two sites being given the same ID on the same day. Each team should be assigned a 'block' of identifiers to work through, large enough to cover the intended number of profile descriptions. Unused numbers may need be flagged in databases (e.g. in project-level metadata) to prevent future users erroneously thinking that site data is missing.
:::

::: {#nte-sitetags .callout-note collapse="true" appearance="minimal" title="Other profile groupings"}
It may be desirable to be able to group profile descriptions in more complex ways within a database. Examples of these might include

-   descriptions belonging to a transect or plot
-   descriptions representing a chrono-, climo- or topo-sequence

A tagging system, where multiple labels can be added to a given site, can be useful in these cases. Otherwise, consider whether additional custom metadata fields should be implemented.
:::

### Description authors {#sec-ref-auth}

Each profile description must have a single responsible author. Record their name. Additional participants should also be noted - soil description is often a group activity. Where sample chain of custody needs to be tracked, recording the name of the sampler may also be necessary.

::: {#cau-priv .callout-caution collapse="true" appearance="minimal" title="Data privacy"}
New Zealand data privacy laws are strict. Personally identifying information must be stored securely and not distributed without a good reason. Personal details such as names may not be publishable without explicit consent. Data storage solutions must respect these requirements.
:::

### Site location {#sec-ref-loc}

Record location coordinates as eastings and northings in New Zealand Transverse Mercator, e.g [1586090E, 5405720N]{.ceg}.

Location coordinates are the minimum requirement for site location; see @sec-loc for a more detailed discussion.

### Date of observation {#sec-ref-dt}

Record the date of observation using the ISO 8601 standard, [YYYY-MM-DD]{.ceg} [@internationalorganizationforstandardization] e.g., [2025-01-01]{.ceg}. Time is not essential but is recommended, as it can be useful for correlating site descriptions with photographs or other data recorded using digital means (e.g. GPS tracks, environmental sensor data). If recording time, use the 24-hour format, e.g. [14:35]{.ceg} for 2:35 pm.

### Site purpose {#sec-ref-why}

The original task for which a site was observed can affect what data is recorded and is relevant to data re-use decisions.

Record the reason for observing the site using the schema in @tbl-ss-reason. This parameter does not need to be recorded in the field.

```{r}
#| label: tbl-ss-reason
#| tbl-cap: "Site purpose"

dat_ss_reason <-
  tribble(~Code, ~Name, ~Description,
          'E', 'Mapping exploration', 'Reconnaissance sites, or sites described in the early stages of building knowledge to underpin, for example, a soil map. May have biased or opportunistic sample locations (e.g. chosen during fieldwork) and moderate location precision',
          'V', 'Mapping validation', 'Sites described when there is already a substantial knowledge base from previous work, with the purpose of gap‑filling or refining. May be low‑detail, as their main role is to confirm or disprove predictions',
          'R', 'Research', 'Sites described to support a scientific investigation, which may or may not be related to soil mapping. Often predetermined by a model and located with high precision',
          'M', 'Monitoring', 'Sites intended primarily for repeated, regular re‑sampling over time. This code implies ongoing measurements, with an expectation of minimal major disturbance between visits',
          'T', 'Reference', 'Calibration sites that provide a definitive description and analysis of a soil type, serving as benchmarks for consistent classification. Typically have high morphological detail and extensive analytical data, representing ‘ideal’ profiles for particular soil classes',
          'D', 'Development', 'Sites established to inform proposed or planned changes to the land, such as altering land use or undertaking construction. Often part of resource consent planning. The soil may be removed or sealed after observation, limiting data re‑use potential')

tbl_ss_reason <- gt(dat_ss_reason) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red', 
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_ss_reason
```

### Site selection method {#sec-ref-sampstrat}

Spatial sampling strategies vary in their statistical properties. Soil data is stored for long periods and is often re-used for multiple projects; specific re-uses may not be appropriate or may require that certain permissions be obtained.

Record the spatial sampling strategy used using the schema in @tbl-ss-sampstrat. This parameter does not usually need to be recorded in the field.

```{r}
#| label: tbl-ss-sampstrat
#| tbl-cap: "Spatial sampling strategies"

dat_ss_sampstrat <-
  tribble(~Code, ~Name, ~Description,
          'A', 'Arbitrary', 'Sites chosen while in the field, in response to direct observation of on-ground conditions. These are often practical and time-efficient but are also likely to over-sample the more easily accessible and/or interesting parts of the landscape, to the detriment of overall representativeness',
          'D', 'Data-driven', 'Sites chosen before a field visit using a loose set of rules or principles (e.g. “visit the major landforms on this farm but stay close to tracks”). May be subject to minor changes once on site',
          'M', 'Model-based', 'Sites chosen before a field visit using a deterministic and replicable set of rules, including rules for relocating sites where on-ground conditions make this necessary',
          'R', 'Random', 'Sites chosen by a randomised method, including stratified-random sampling',
          'T', 'Transect', 'Sites chosen to form a line over the land surface, with or without regular spacing',
          'G', 'Grid', 'Sites chosen to form a regular or semi-regular lattice over the land surface')

tbl_ss_sampstrat <- gt(dat_ss_sampstrat) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_ss_sampstrat
```

[A]{.ceg} Arbitrary, [D]{.ceg} data-driven, and [R]{.ceg} model-based sampling all represent attempts to use secondary environmental data sources to increase the efficiency of fieldwork, with the goal of minimising total sites while maximising the representativeness and quality of information gathered at each site. The success of these strategies depends on whether relevant and accurate model inputs are available.

[R]{.ceg} Random, [T]{.ceg} (regular) transect and [G]{.ceg} (regular) grid sampling effectively ignore environmental parameters, although stratified random sampling might partially incorporate such information. These strategies are useful when the spatial distribution of a parameter of interest is unknown or poorly understood, but can lead to sites in inefficient or unsafe locations.

Some of these methods can be combined - for example, data-driven transect choices with randomised locations along each. In such cases, the 'last choice' should be recorded - in this example, [R]{.ceg}. The details of the full sampling strategy should be discussed in project metadata.

::: {#nte-sitenum .callout-note collapse="true" appearance="minimal" title="How many sites?"}
Estimating the number of sites required to complete a project and choosing an optimal sampling strategy is outside the scope of this document. Advice on these topics can be found in @grealish2017 and @mckenzie2008.
:::

### Exposure type {#sec-pr-exptype}

The method by which the soil was exposed for observation affects which parameters can be recorded, or at least recorded reliably. Record profile exposure method using one of the options in @tbl-pr-gear.

```{r}
#| label: tbl-pr-gear
#| tbl-cap: "Method of profile exposure"

dat_pr_gear <-
  tribble(~Code, ~Name, ~Description,
          'P1', 'Minimal pit', 'Pit up to 0.5 m wide and generally no deeper than 1m',
          'P2', 'Standard pit', 'Pit or trench > 1 m wide, exposing the complete profile to regolith or bedrock',
          'E', 'Exposure', 'Road cutting or similar, chipped back and cleaned',
          'C1', 'Small Core', 'Relatively undisturbed soil core, < 50 mm diameter',
          'C2', 'Large Core', 'Relatively undisturbed soil core, ≥ 50 mm diameter',
          'A', 'Auger', 'Hand or mechanical auger, including e.g. post hole diggers')

tbl_pr_gear <- gt(dat_pr_gear) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_pr_gear
```

-   [A]{.ceg} samples cannot be used to reliably determine soil structure due to the degree of disturbance.
-   Combined sampling is common in field mapping, e.g. small pit excavation to \~50-75 cm followed by augering to a required depth in relatively well-understood landscapes. Record each method used and note the depth at which the method changed, e.g., [P1/A (75 cm)]{.ceg}.
-   Gouge-style augers used in soft marine sediments may be included in [C2]{.ceg} even though they commonly narrow to \< 50 mm at their maximum depth.
-   [C1]{.ceg} style samplers are common in paddock-scale monitoring and research, where they are used for rapid surface measurement. These are not generally suitable for mapping as they cannot sample below the topsoil.

Further discussion of appropriate tools for soil profile exposure may be found in @sec-digging.

### Depth of observation {#sec-pr-doa}

Record the maximum depth of observation from the soil surface in whole centimetres. This depth does not necessarily need to be recorded in the field as it can be determined from the lower depth of the last horizon in the profile description.

A full profile description ideally extends to bedrock or underlying geological material, but such materials may not always be accessible with the equipment and time available. More commonly, reaching the C horizon (@sec-hzn-convp) is the deepest practically achievable goal. The minimum depth of observation for most applications in New Zealand, where possible, is 100 cm. Classification systems like the *New Zealand Soil Classification* [@hewitt2025] and LUC [@lynn2009] rely on ruling out certain soil features within that depth range, as well as positively identifying others.

#### Stopping early {#sec-pr-rsa}

When the target depth cannot be achieved, note the reason for stopping early using the codes in @tbl-ss-stoppage.

```{r}
#| label: tbl-ss-stoppage
#| tbl-cap: "Stoppage reason"

dat_ss_stoppage <-
  tribble(~Code, ~Name, ~Description,
          'B', 'Bedrock', 'Underlying fresh rock mass encountered, no more soil or weathering substrate to describe',
          'R', 'Resistance', 'Materials too difficult to extract with available equipment',
          'C', 'Collapse', 'Excavation unstable or unsafe due to e.g. wetness or loose particle packing',
          'W', 'Water', 'Excavation reached the water table and cannot extract saturated material from below',
          'I', 'Identified', 'The depth required to confirm a target parameter of interest was reached',
          'O', 'Other', 'Circumstances unrelated to the soil itself caused early stoppage')

tbl_ss_stoppage <- gt(dat_ss_stoppage) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10))

tbl_ss_stoppage
```

## Site setting {#sec-ss-set}

With the reference data established, the next task is to describe the site's location in relation to its setting. As noted in earlier chapters, some of these parameters can be recorded remotely or after fieldwork, with reference to site notes and photographs.

The following requirements comprise a minimum dataset for any soil profile description. Data is gathered at the landform scale with any site-scale deviations noted. More specific measurements are taken at the site scale when observing the land surface (@sec-pr-surface).

### Location {#sec-ss-recloc}

Location coordinates are captured as reference data (@sec-ref-loc). Additionally, record elevation in meters, and note the sensor or data source used to obtain location and elevation data using @tbl-ss-locsrc and @tbl-ss-elevsrc respectively. If the equipment in use does not estimate expected error automatically, make a note of it manually.

[1586090E, 5405722N GS (462 ± 2 m A)]{.ceg} for a location obtained using a smartphone, and an elevation measured with a calibrated altimeter.

When establishing a monitoring site, record additional data relevant to relocating the site during future visits as described in @sec-loc-rel.

### Geomorphology {#sec-ss-recgmph}

Record one each of landscape, landform, and landform element, using the codes in @tbl-gmph-lsnames, @tbl-gmph-lfnames, and @tbl-gmph-elnames respectively. Optional detail may be added to landform and landform element codes using the full system described in @sec-gmph.

E.g., [Te Fn Sl]{.ceg} for a site on the side of a low-angle fan at the back of a terrace.

### Parent material {#sec-ss-}

Record the dominant parent material origin (@tbl-ss-pmo) and lithology of the landform on which the site is located.

E.g., [Tp Rh]{.ceg} for a landform blanketed in rhyolitic tephra.

### Climate and weather

Record the current weather using an option from @tbl-ss-weatherc, and optionally note recent rainfall using @tbl-ss-weatherp.

E.g. [SU 21°C, NW]{.ceg} for fine weather with no rainfall in the previous week.

### Biota

Record the dominant LUC vegetation class [see @lynn2009, appendix 3] for the landform on which the site is located.

E.g., [fP]{.ceg} for a site in podocarp forest.

### Land cover, use and management

Record the dominant land cover class of the landform using an option from @sec-gmph-lfm.

Record the dominant land use of the landform using an option from [@law2024]. Note also the degree to which the land use has impacted the soil profile using @tbl-ss-luimp.

E.g. [2.2.1 M]{.ceg} for a landform being used primarily for dairying.

Ancillary landuses may optionally be recorded if they are considered to be impacting the soil profile - for example, [2.7.3]{.ceg} for a landform on a dairy farm that is regularly receiving waste water.

Land management activities should be noted in free text when they are considered to be impacting the soil profile.
