# Profile description {#sec-site}

```{r}
library(dplyr)
library(tibble)
library(gt)
```

## Defining the site {#sec-pr-site}

Soil profiles are described at *sites*. The site concept is necessarily somewhat flexible to account for a wide range of observation purposes and sampling methods. Generally speaking it is considered to include the immediate surroundings of a profile description, up to a maximum radius of \~10 m, so it will generally sit within a single landform. Where a site is close to a landform edge and the processes operating on the neighbouring landform are not judged to affect the profile being described, those neighbouring landforms should be ignored.

Multiple profile descriptions made across larger sampling plots or along transects that cross multiple landforms (or even landscapes) should usually be treated as a series of related sites rather than a single site.

The soil profile description starts with an observation of the local land surface and works downwards through the soil profile towards the underlying parent material. The chief tasks are assessing the surface, delineating pedologic horizons (and any additional geogenic layers, if present), and describing their morphological character.

Before beginning a soil profile description it is essential to record some additional reference data.

## Reference data {#sec-ref}

Reference data are metadata that clearly identify a soil profile description. These data enable the description to be linked with other profiles, samples, analytical results, and photographs. Reference data include:

-   an identifier
-   who completed the description
-   where the description was located
-   when the description was done
-   why the description was needed
-   how the location was chosen
-   what equipment was used
-   depth of observation.

These data are discussed in turn below.

### Identifier {#sec-ref-id}

Assign each profile description an identifier that is at least unique at a whole-of-project level, and preferably at a whole-of-organisation level. There are many ways to build identifier systems (see @nte-uuid), but a good system usually has a predictable structure, is open-ended, and is easy to remember. Incorporating location data into the identifier can help avoid confusion between multiple projects within an organisation; for example, a region code followed by a sequential number: [WLG_001]{.ceg}.

This kind of identifier is 'semi-unique' in that there is still some potential for confusion if the data are combined with data sets from other sources; for example, a vegetation survey group working independently in the same region may have decided to use the same naming protocol for their descriptions.

::: {#nte-uuid .callout-note collapse="true" appearance="minimal" title="Unique identifiers"}
Profile description identifier formats are often a compromise between robustness and simplicity. Well-designed databases use truly unique identifiers for every piece of data they hold. Under contemporary best practice, these are usually 'UUIDs', 128-bit hexadecimal codes (see @davis2024). Although technically robust, these are difficult to work with day-to-day, and most workers will find themselves preferring to interact with simpler, semi-unique identifiers. In a soil profile database, semi-unique identifiers should be implemented as UUID aliases. End users should primarily interact with the aliases, and the UUIDs should remain hidden.
:::

::: {#nte-teamsites .callout-note collapse="true" appearance="minimal" title="Working in teams"}
Semi-unique identifiers require active management when working with multiple field teams operating in parallel. This process avoids the risk of two profile descriptions being given the same ID on the same day. Each team should be assigned a 'block' of identifiers to work through, large enough to cover the intended number of profile descriptions. Unused numbers may need to be flagged in databases (e.g. in project-level metadata) to prevent future users erroneously thinking that descriptions are missing.
:::

::: {#nte-sitetags .callout-note collapse="true" appearance="minimal" title="Other profile groupings"}
It may be desirable to be able to group profile descriptions in more complex ways within a database. Examples of these might include:

-   descriptions belonging to a transect or plot
-   descriptions representing a chrono-, climo- or topo-sequence.

A tagging system, whereby multiple labels can be added to a given description, can be useful in these cases. Otherwise, consider whether additional custom metadata fields should be implemented.
:::

### Authors {#sec-ref-auth}

Each soil profile description must have a single author responsible for it. Record their name. Additional participants should also be noted (soil description is often a group activity). Where sample chain of custody needs to be tracked, recording the name of the sampler may also be necessary.

::: {#cau-priv .callout-caution collapse="true" appearance="minimal" title="Data privacy"}
New Zealand data privacy laws are strict. Personally identifying information must be stored securely and not distributed without a good reason. Personal details such as names must not be publishable without explicit consent. Data storage solutions must respect these requirements.
:::

### Location {#sec-ref-loc}

Record location coordinates as eastings and northings in New Zealand Transverse Mercator (e.g. [1586090E, 5405720N]{.ceg}). Location coordinates are the minimum requirement for location (see @sec-ss-recloc for full requirements).

### Date of observation {#sec-ref-dt}

Record the date of observation using the ISO 8601 standard, [YYYY-MM-DD]{.ceg} [@internationalorganizationforstandardization]; for example, [2025-01-01]{.ceg}. Recording the time is not essential, but is recommended because it can be useful for correlating soil profile descriptions with photographs or other data recorded using digital means (e.g. GPS tracks, environmental sensor data). If recording time, use the 24-hour format; for example, [14:35]{.ceg} for 2:35 pm.

### Purpose {#sec-ref-why}

The original task for which a profile description was completed can affect what data were recorded and is relevant to data reuse decisions. Record the reason for observing the profile using the schema in @tbl-ss-reason. This parameter does not need to be recorded in the field.

```{r}
#| label: tbl-ss-reason
#| tbl-cap: "Profile description purpose"

dat_ss_reason <-
  tribble(~Code, ~Name, ~Description,
          'E', 'Mapping exploration', 'Reconnaissance descriptions, or profiles described in the early stages of building knowledge to underpin, for example, a soil map. They may have biased or opportunistic sample locations (e.g. chosen during fieldwork) and moderate location precision.',
          'V', 'Mapping validation', 'Profiles described when there is already a substantial knowledge base from previous work, with the purpose of gap‑filling or refining. They may be low‑detail, as their main role is to confirm or disprove predictions.',
          'R', 'Research', 'Profiles described to support a scientific investigation, which may or may not be related to soil mapping. These are often predetermined by a model and located with high precision.',
          'M', 'Monitoring', 'Profiles described at locations that will be visited more than once. This code implies ongoing measurements, with an expectation of minimal major disturbance between visits.',
          'T', 'Reference', 'Calibration profiles that provide a definitive description and analysis of a soil type, serving as benchmarks for consistent classification. They typically have high morphological detail and extensive analytical data, representing ‘ideal’ profiles for particular soil classes.',
          'D', 'Development', 'Profiles described to inform proposed or planned changes to the land, such as altering land use or undertaking construction. They are often part of resource consent planning. The soil may be removed or sealed after observation, limiting data reuse potential.')

tbl_ss_reason <- gt(dat_ss_reason) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red', 
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_ss_reason
```

### Location selection method {#sec-ref-sampstrat}

Spatial sampling strategies vary in their statistical properties. Soil data are stored for long periods and are often reused for multiple projects; specific reuses may not be appropriate or may require that certain permissions be obtained.

Record the spatial sampling strategy using the schema in @tbl-ss-sampstrat. This parameter does not usually need to be recorded in the field.

```{r}
#| label: tbl-ss-sampstrat
#| tbl-cap: "Spatial sampling strategies"

dat_ss_sampstrat <-
  tribble(~Code, ~Name, ~Description,
          'A', 'Arbitrary', 'Locations chosen while in the field in response to direct observation of on-ground conditions. These are often practical and time-efficient, but are also likely to over-sample the more easily accessible and/or interesting parts of the landscape, to the detriment of overall representativeness.',
          'D', 'Data-driven', "Locations chosen before a field visit using a loose set of rules or principles (e.g. 'Visit the major landforms on this farm but stay close to tracks'). They may be subject to minor changes once on-site.",
          'M', 'Model-based', 'Locations chosen before a field visit using a deterministic and replicable set of rules, including rules for relocating sites where on-ground conditions make this necessary.',
          'R', 'Random', 'Locations chosen by a randomised method, including stratified-random sampling.',
          'T', 'Transect', 'Locations chosen to form a line over the land surface, with or without regular spacing.',
          'G', 'Grid', 'Locations chosen to form a regular or semi-regular lattice over the land surface.')

tbl_ss_sampstrat <- gt(dat_ss_sampstrat) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_ss_sampstrat
```

[A]{.ceg} (arbitrary), [D]{.ceg} (data-driven), and [M]{.ceg} (model-based) sampling all represent attempts to use secondary environmental data sources to increase the efficiency of fieldwork, with the goal of minimising total workload while maximising the representativeness and quality of information gathered at each location. The success of these strategies depends on whether relevant and accurate model inputs are available.

[R]{.ceg} (random), [T]{.ceg} (regular) transect and [G]{.ceg} (regular) grid sampling effectively ignore environmental parameters, although stratified random sampling might partially incorporate such information. These strategies are useful when the spatial distribution of a parameter of interest is unknown or poorly understood, but can lead to profile descriptions being made in inefficient or unsafe locations.

Some of these methods can be used in a sequence that progressively narrows down the choice of location (e.g. data-driven transect locations across a whole study area, then randomised final locations along each transect). In such cases, the 'last choice' should be recorded on the profile description – in this example, [R]{.ceg}. The details of the full sampling strategy should be discussed in project metadata.

::: {#nte-sitenum .callout-note collapse="true" appearance="minimal" title="How many profiles?"}
Estimating the number of profile descriptions required to complete a project and choosing an optimal sampling strategy are outside the scope of this document. Advice on these topics can be found in @grealish2017 and @mckenzie2008.
:::

### Exposure type {#sec-pr-exptype}

The method by which the soil was exposed for observation affects which parameters can be recorded, or at least recorded reliably. Record the profile exposure method using one of the options in @tbl-pr-gear.

```{r}
#| label: tbl-pr-gear
#| tbl-cap: "Method of profile exposure"

dat_pr_gear <-
  tribble(~Code, ~Name, ~Description,
          'P1', 'Standard pit', 'Pit or trench >1 m wide, exposing the complete profile to regolith or bedrock',
          'P2', 'Minimal pit', 'Pit up to 0.5 m wide and generally no deeper than 1 m',
           'E', 'Exposure', 'Road cutting or similar, chipped back and cleaned',
          'C1', 'Large core', 'Relatively undisturbed soil core, ≥30 mm diameter',
          'C2', 'Small core', 'Relatively undisturbed soil core, <30 mm diameter',
           'A', 'Auger', 'Hand or mechanical auger, including post-hole diggers')

tbl_pr_gear <- gt(dat_pr_gear) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_pr_gear
```

-   [A]{.ceg} samples cannot be used to reliably determine soil structure due to the degree of disturbance.
-   Combined sampling is common in field mapping; for example, small pit excavation to \~50–75 cm followed by augering to a required depth in relatively well-understood landscapes. Record each method used and note the depth at which the method changed (e.g. [P2/A (75 cm)]{.ceg}).
-   Gouge-style augers used in soft marine sediments may be included in [C1]{.ceg} even though they may narrow to \< 30 mm at their maximum depth.
-   [C2]{.ceg} style samplers are common in paddock-scale monitoring and research, where they are used for rapid surface measurement. These are not generally suitable for mapping because they cannot sample below the topsoil.

Further discussion of appropriate tools for soil profile exposure can be found in @sec-gear-dig, and methods in @sec-dig.

### Depth of observation {#sec-pr-doa}

Record the maximum depth of observation from the soil surface in whole centimetres. This depth does not necessarily need to be recorded in the field because it can be determined from the lower depth of the last horizon in the profile description.

A full profile description ideally extends to bedrock or underlying geological material, but such materials may not always be accessible with the equipment and time available. More commonly, reaching the C horizon (@sec-hzn-convp) is the deepest practically achievable goal. The minimum depth of observation for most applications in New Zealand, where possible, is 100 cm. Classification systems like the *New Zealand Soil Classification* and the LUC [@lynn2009] rely on ruling out certain soil features within that depth range, as well as positively identifying others.

When the target depth cannot be achieved, note the reason for stopping early using the codes in @tbl-ss-stoppage.

```{r}
#| label: tbl-ss-stoppage
#| tbl-cap: "Stoppage reason"

dat_ss_stoppage <-
  tribble(~Code, ~Name, ~Description,
          'B', 'Bedrock', 'Underlying fresh rock mass encountered; no more soil or weathering substrate to describe',
          'R', 'Resistance', 'Materials too difficult to extract with available equipment',
          'C', 'Collapse', 'Excavation unstable or unsafe due to, for example, wetness or loose particle packing',
          'W', 'Water', 'Excavation reached the water-table and cannot extract saturated material from below',
          'I', 'Identified', 'The depth required to confirm a target parameter of interest was reached',
          'O', 'Other', 'Circumstances unrelated to the soil itself caused early stoppage')

tbl_ss_stoppage <- gt(dat_ss_stoppage) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10))

tbl_ss_stoppage
```

## Profile description setting {#sec-ss-set}

With the reference data established, the next task is to describe the profile's location in relation to its setting. As noted in earlier chapters, some of these parameters can be recorded remotely or after fieldwork, by referring to notes and photographs.

The following requirements comprise a minimum data set for any soil profile description. Data are gathered at the site scale. Further measurements are taken when observing the land surface (@sec-pr-surface).

### Location {#sec-ss-recloc}

Location coordinates are captured as reference data (@sec-ref-loc). Also, record elevation in metres, and note the sensor or data source used to obtain location and elevation data using @tbl-ss-locsrc and @tbl-ss-elevsrc, respectively. If the equipment in use does not estimate expected error automatically, estimate it manually.

**For example:** use [1586090E, 5405722N ± 3 GS (462 ± 5 m A)]{.ceg} for a location obtained using a smartphone, and an elevation measured with a calibrated altimeter.

When establishing a monitoring location, record additional data relevant to relocating it during future visits, as described in @sec-loc-rel.

### Geomorphology {#sec-ss-recgmph}

Record one each of landscape and landform using the codes in @tbl-gmph-lsnames, @tbl-gmph-lfnames respectively. If relevant, note any significant structural control on the landscape using @tbl-hr-slopeenc. 

For detailed sites, record the landform's slope and orientation, then record the landform element using @tbl-gmph-elnames, including any necessary modifiers available in @tbl-hr-slopeinc and @tbl-hr-slopeenc. These parameters are optional for routine descriptions.

**For example:** use [Te Fn Sl]{.ceg} for a site on the side of a low-angle fan at the back of a terrace.

Optional detail may be added to the landscape description using the options in @sec-gmph-lscpar, and to the landform using the options in @sec-gmph-lfmdsc.

### Parent material {#sec-ss-pm}

Record the dominant parent material origin (@tbl-ss-pmo) and lithology (@tbl-ss-pm).

**For example:** use [Tp RHYL]{.ceg} for a landform blanketed in rhyolitic tephra.

### Climate and weather {#sec-ss-clw}

Record the current weather using an option from @tbl-ss-weatherc, and note the temperature (optional). For routine or detailed recording, describe any recent rainfall using @tbl-ss-weatherp.

**For example:** use [SU 21°C, NW]{.ceg} for fine weather with no rainfall in the previous week.

### Biota {#sec-ss-bio}

For rapid assessment of vegetation, record the LCDB cover class (see @sec-ss-cvusmg). For routine work, record the dominant LUC vegetation class [see @lynn2009, Appendix 3]. For detailed work, complete a RECCE description.

**For example:** use [fP]{.ceg} for a profile description made in a podocarp forest.

Direct assessment of earthworm presence in the topsoil is sometimes desired [@shepherd2000; @shepherd2000a]. If so, record the number of earthworms observed in a \~20 cm cube of topsoil. Convert the count to a rate per square metre as $25 × n$. Accurate counts require moist soil conditions [@schon2023].

### Land cover, use, and management {#sec-ss-cvusmg}

Record the current dominant land-cover class using an option from @sec-lcov. Record the dominant land use using an option from @fig-ss-nzlum. Ancillary land uses may also be recorded if they are impacting the soil profile.

**For example:** use [2.7.3]{.ceg} for a profile description made in a paddock on a dairy farm that is regularly receiving waste water.

Prior land uses may be recorded if they are known and are considered to have been capable of affecting the profile (e.g. cropping prior to vineyard conversion). Time since land-use change should also be noted.

Land management activities (current and past, where known) should be noted in free text when they are considered to be affecting the soil profile. Minimise the use of jargon and clearly separate information on current and past land uses.

**For example:** in an orchard recently converted from sheep pasture, use [33]{.ceg}, [2.4.1]{.ceg} (previous [2.2.3]{.ceg}), *Drip irrigation, manual control, installed and planted 2022. Originally cleared 1920, other previous management details unknown.*
