```{r}
library(dplyr)
library(tibble)
library(gt)
library(gtExtras)
```

# Land surface

The land surface at a site is assessed at the landform scale.

## Surface cover

Record the following generalised surface cover components as a percentage composition of the area surrounding a point observation, or the area within a sampling plot.

```{r}
#| label: tbl-pr-surfcov
#| tbl-cap: "Surface cover types"

dat_pr_surfcov <-
  tribble(~Code, ~Name, ~Description,
          'BE', 'Bare earth', 'Sediment or soil, < 2 mm', 
          'BR', 'Bare rock', 'Rock fragments, shells, or bones, > 2 mm',
          'BO', 'Outcrop', 'Bare rock outcrop',
          'WT', 'Water', 'Open water, ice or snow',
          'PT', 'Peat', 'Plant material decomposing under wet to saturated conditions',
          'LH', 'Litter', 'Plant material decomposing under dry to moist conditions',
          'GR', 'Small plants', 'Grass, herbs, moss, sedges, annual crops',
          'TR', 'Large plants', 'Shrubs, trees (stems and roots), perennial crops',
          'SL', 'Sealed', 'Building, concrete, asphalt and similar')

tbl_pr_surfcov <- gt(dat_pr_surfcov) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(20), Name ~pct(20))

tbl_pr_surfcov
```

[GRAPHIC: Turn the above into a 3x3 grid with labels underneath conceptual diagrams]{style="background-color:lightgreen"}

These parameters should be recorded only to nearest 5-10%. It will be unusual to encounter more than three components at any one site.

```{r}

# draw donut chart with 10% breaks next to a couple of pictures for examples



```

## Surface water

Where surface water is present within the landform, as well as noting its areal dominance above, note the type and degree of persistence using the codes in @tbl-pr-swater and @tbl-pr-swatert below.

```{r}
#| label: tbl-pr-swater
#| tbl-cap: "Surface water character (WRB 2022)"

dat_pr_swater <-
  tribble(~Code, ~Name, ~Description,
          'S', 'Seawater', 'Saline ocean or brackish tidal waters', 
          'I', 'Surface water', 'Waters flowing over the surface from further inland',
          'G', 'Groundwater', 'Waters rising from adjacent springs or an elevated natural water table',
          'R', 'Rainfall', 'Rain falling on or near the site and perching on the surface',
          'A', 'Anthropic', 'Water from a manmade discharge location or backing up behind a dam')

tbl_pr_swater <- gt(dat_pr_swater) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(20), Name ~pct(20)) |>
  sub_missing(missing_text = '')

tbl_pr_swater
```

```{r}
#| label: tbl-pr-swater
#| tbl-cap: "Surface water persistence (WRB 2022)"

dat_pr_swater <-
  tribble(~Code, ~Name, ~Description,
          'P', 'Permanent', 'Surface water persists year-round', 
          'S', 'Seasonal',  'Surface water is present during wet seasons',
          'T', 'Temporary', 'Surface water is present due to a specific event')

tbl_pr_swater <- gt(dat_pr_swater) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(20), Name ~pct(20)) |>
  sub_missing(missing_text = '')

tbl_pr_swater
```

## Surface cracking

Soil surface cracking can occur for a number of reasons:

-   Soil parent material includes shrink-swell clays.
-   Sediment originally deposited underwater (e.g. during a flood) cracks on drying out.
-   Cracks can be induced by mass movement or earthquakes.

Within the landform, record surface crack presence/persistence using the codes in @tbl-pr-cracktype. If present, record spatial arrangement using the codes in @tbl-pr-crackshape. For detailed descriptions, report median width and spacing between cracks in cm.

e.g. [**RT P 0.5 cm, 15 cm**]{style="color: red"} for surface cracking patterns common to Vertic Melanic soils.

```{r}
#| label: tbl-pr-cracktype
#| tbl-cap: "Crack persistence (WRB 2022)"

dat_pr_cracktype <-
  tribble(~Code, ~Name, ~Description,
          'NO', 'No surface cracks', NA, 
          'RT', 'Reversible cracks', 'Cracks that open and close with changing soil moisture',
          'IT', 'Irreversible cracks', 'Cracks that persist year-round')

tbl_pr_cracktype <- gt(dat_pr_cracktype) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(20), Name ~pct(20)) |>
  sub_missing(missing_text = '')

tbl_pr_cracktype
```

```{r}
#| label: tbl-pr-crackshape
#| tbl-cap: "Crack spatial arrangement (WRB 2022)"

dat_pr_crackshape <-
  tribble(~Code, ~Name, ~Description,
          'P', 'Polygonal', 'Surface cracks forming closed shapes on the land surface', 
          'L', 'Linear', 'Surface cracks running roughly parallel to each other',
          'I', 'Irregular', 'Surface cracks with no discernible pattern')

tbl_pr_crackshape <- gt(dat_pr_crackshape) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(20), Name ~pct(20)) |>
  sub_missing(missing_text = '')

tbl_pr_crackshape
```

::: callout-note
Recent weather data is useful for interpreting surface crack information. See @sec-ss-recentrain.
:::


## Microrelief

Microrelief occurs within a landform, so its components are smaller than the landform itself - usually too small to be sensed remotely. Its presence helps signify particular soil types and influences many small-scale soil processes such as local water infiltration, accumulation of surface litter, and topsoil development.

### Natural Microrelief

Natural microrelief emerges from slope processes, biotic activity and climate effects. 

Record the dominant microrelief type using the codes in @tbl-pr-micreltype. For detailed descriptions, record the average height and spacing of repeating microrelief elements in cm.

```{r}
#| label: tbl-pr-micreltype
#| tbl-cap: "Types of microrelief (Aus 2023 and WRB 2022)"

dat_pr_micreltype <-
  tribble(~Code, ~Name, ~Description,
          'N', 'None',   'Landform surface is naturally smooth or level', 
          'G', 'Gilgai', 'Induced by shrink-swell clay activity',
          'S', 'Swamp',  'Hummocky landforms common in wetland areas, derived by biotic activity',
          'K', 'Karst',  'Induced by subsurface collapse in limestone-dominated landscapes',
          'A', 'Animal', 'Terracettes, trampling or pugging induced by larger animals (cows, pigs, ungulates)',
          'F', 'Frost',  'Induced by freeze-thaw activity')

tbl_pr_micreltype <- gt(dat_pr_micreltype) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(20), Name ~pct(20)) |>
  sub_missing(missing_text = '')

tbl_pr_micreltype
```

### Anthropogenic microrelief

Small-scale human alterations to the soil surface are common in agricultural areas (current and former). These include creation of some microrelief, but more commonly its removal. 

Record the dominant anthropogenic microrelief type using the codes in @tbl-pr-amicreltype. For detailed descriptions, record the average height and spacing of repeating elements in cm.

```{r}
#| label: tbl-pr-amicreltype
#| tbl-cap: "Types of Anthropogenic microrelief (Aus 2023 and WRB 2022)"

dat_pr_micreltype <-
  tribble(~Code, ~Name, ~Description,
          'N', 'None',        'Landform surface is unaltered by humans', 
          'S', 'Smoothed',    'Landform surface has been smoothed, infilled or levelled',
          'R', 'Raised Beds', 'Elevated areas have been created for specific crops',
          'T', 'Terraces and tracks', 'Hillsides have been cut/filled into terraces, or tracks cut in across the slope',
          'D', 'Drained',     'Open drainage ditches have been installed'
          # might be a few more to go here...
          )

tbl_pr_micreltype <- gt(dat_pr_micreltype) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(20), Name ~pct(20)) |>
  sub_missing(missing_text = '')

tbl_pr_micreltype
```

