# Profile

```{r}
library(dplyr)
library(tibble)
library(gt)
library(gtExtras)
```

## Reference data

Reference data clearly identifies a soil profile description within a larger dataset, linking it to other profiles, samples, analytical results and photographs. Reference data also covers:

-  who completed the description
-  when and where it was located (as [previously discussed](ss_01_location.qmd))
-  what level of detail was achieved
-  why the work was done
-  how the profile was exposed and examined.

## Naming/numbering sites and site collections

-   tips for robust systems for projects of varying scale
-   tips for working across multiple teams
-   distinguishing site ID's from database ID's (e.g. you might want a UUID but you don't want to be writing one out by hand)
-   tagging by e.g. transect membership

## Authors

Each profile description should have a single responsible author. Additional participants may also be noted. A short unique identifier for each officer is recommended, for example the last three or four letters of a surname followed by the first letter of a first name, e.g. [SMIJ]{.ceg} for John Smith. These may be linked to more robust [UUIDs]{.glossary} in a database.

## Date/Time

See @sec-ss-dt for recording requirements.

::: callout-caution
## Data privacy

New Zealand data privacy laws are strict. Personally identifying information must be stored securely and not distributed without a good reason. Personal details such as names may not be publishable without explicit consent. Data storage solutions must respect these requirements.

:::


## Depth of observation

- requirements to classify against the NZSC
- requirements for mapping
- requirements for research

## Equipment options

- for NZ conditions

## Profile exposure options

- how to dig a pit safely
- how to measure depth on significant slopes

## Exposure type

The method by which the soil was exposed for observation affects which parameters can be recorded, or at least recorded reliably.

```{r}
#| label: tbl-pr-gear
#| tbl-cap: "Method of profile exposure"

dat_pr_gear <-
  tribble(~Code, ~Name,
          'P1', 'Quick pit – up to 0.5 m wide and generally no deeper than 1m',
          'P2', 'Full pit > 1 m wide, exposing the complete profile to regolith or bedrock',
          'E', 'Exposure – road cutting or similar, chipped back and cleaned',
          'C1', 'Relatively undisturbed soil core, < 50 mm diameter',
          'C2', 'Relatively undisturbed soil core, ≥ 50 mm diameter',
          'A', 'Auger sample (hand or mechanical, including e.g. post hole diggers)')

tbl_pr_gear <- gt(dat_pr_gear) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(20))

tbl_pr_gear
```

-   [A]{.ceg} samples generally cannot be used to determine soil structure due to the degree of disturbance.
-   Combined sampling is common in field mapping, e.g. small pit excavation to \~0.50-0.75 m and augering to a required depth in relatively well-understood landscapes. Record these as [P1]{.ceg}.
-   Gouge-style augers used in soft marine sediments may be included in [C2]{.ceg} even though they commonly narrow to \< 50 mm at their maximum depth.
-   [C1]{.ceg} style samplers are common in paddock-scale monitoring and research, where they are used for rapid surface measurement. These are not generally suitable for mapping as they cannot sample below the topsoil.

[add some gear pics/diagrams here]{.diag}

### Regolith stuff

TBD