# Defining soil horizons {#sec-hor}

```{r}
library(dplyr)
library(tibble)
library(ggplot2)
library(gt)
```

A soil horizon is a distinct zone that can be identified within a soil profile, usually but not always parallel to the land surface. Soil horizons are distinguished by differences in texture, stoniness, structure, colour, consistence or other soil properties as described in the following sections. Delineated horizons should ideally represent depth intervals over which important soil attributes change minimally, although some short-range internal variation may occur. Conversely, the horizon boundaries should represent depth intervals over which there is a relatively high rate of change of important soil properties. Soil horizons emerge from a complex interaction of differentiating and homogenising forces, the detailed discussion of which is beyond the scope of this handbook. See @hartemink2020 and @palmer2025 for further reading.

Soil horizons are commonly grouped into two major zones - the topsoil, where biological activity is elevated and new additions of soil parent material may be received, and the subsoil, a zone where weathering processes are concentrated, generating new soil minerals and accumulating materials from above. Below these zones, collectively called the 'solum', a 'subsolum' of variably weathered soil parent materials continues, often for many meters. These materials can include paleosols buried by subsequent depositional events, unconsolidated deposits that have not been subject to strong weathering, and rocks weathering in place. In New Zealand, many landscapes are underlain by sedimentary rocks that were not strongly consolidated before becoming exposed to weathering forces, so the boundary between the solum and the subsolum can be particularly difficult to define.

While this handbook focuses primarily on describing the solum, the subsolum, when encountered, should be described as well. Subsolum materials play a key role in deep drainage, groundwater recharge, slope stability, and ecological support [@juilleret2016]. 'Horizon' boundaries within the subsolum will most commonly relate to lithological or depositional discontinuities.

::: {#nte-hzn-cls .callout-note collapse="true" appearance="simple" title="Soil horizons and classification systems"}
All soil classification systems use particular soil horizons to help locate a soil within the system. To be classification-relevant, such horizons must not be merely present but must also meet particular thresholds of development, e.g. by having a minimum thickness or a particular abundance range for a key property.
:::

## Defining the soil surface {#sec-pr-ssurf}

The soil surface has a slightly more restrictive definition than the land surface. The soil surface begins below the layer of fresh leaf litter or small living plants, and at the top of organic soil materials such as decomposing litter or peat. The mineral soil surface is further restricted, identified as the top of the first horizon that is dominated by mineral soil material. Horizon depths are defined with reference to the soil surface overall - thus only fresh litter and/or small plants should be scraped back before measuring.

[draw this]{.diags}

## Recording depths {#sec-pr-depths}

Horizon depths are recorded starting from the soil surface at 0 cm and working downwards cumulatively.

When examining single small cores or auger points (see @tbl-pr-gear, type [A]{.ceg}, [C1]{.ceg} and perhaps [C2]{.ceg}), record the observable upper and lower boundary depths of each horizon from the soil surface in whole centimetres (e.g. [28-38 cm]{.ceg}).

Where a wider exposure (see @tbl-pr-gear, type [E]{.ceg}, [P1]{.ceg}, [P2]{.ceg}) or multiple undisturbed cores are available and greater detail is desired, one may choose to record upper and lower boundary depths as a median and range, again in whole centimetres (e.g. [28 cm (27-29), 38 cm (37-40)]{.ceg}). At least 5 measurements should contribute to such an assessment. This practice may be useful for describing irregular, convolute or discontinuous horizons (@sec-pr-bdyshp).

Measure boundaries to and from the midpoint of each horizon transition zone (see @sec-pr-bdydist). Where depths are correctly measured, horizon thicknesses can be calculated after fieldwork.

::: {#imp-stp .callout-important collapse="true" appearance="simple" title="Defining horizons on steep slopes"}
On steeper slopes, measuring horizons from vertical exposures can exaggerate horizon thicknesses, as the soil horizons mantle the landscape perpendicular to the sloping surface [@prietzel2019]. This starts to matter at slopes \> 25°, and progressively becomes more important - especially when calculating volumetric stocks of soil components. @fig-hr-slpex shows how a vertical 30-cm depth represents a progressively shallower 'true depth' as slope increases.

Where depths are measured perpendicular to the surface, no adjustment needs to be made. To adjust vertically measured horizon depths to correct for slope, apply the formula $depth × cos(slope)$. Supply depth in cm and slope in radians. Slopes measured in degrees can be converted to radians using the formula $(slope × \pi) ÷ 180$. If this correction is made, record that it has been done in the site notes to prevent the horizon depths from being adjusted any further.

```{r}
#| label: fig-hr-slpex
#| fig-cap: "Depth exaggeration by slope."

deg2rad <- function(deg) {(deg * pi) / (180)}

slp_dat <- data.frame('Slope' = seq(0, 50, 1),
                      'Thickness' = 30 * cos(deg2rad(seq(0, 50, 1)))) |>
                        mutate('Ratio' = Thickness/30)

ggplot(slp_dat) +
  geom_line(aes(y = Thickness, x = Slope),
            linewidth = 1, colour = 'grey40') +
  labs(x= 'Slope (°)', y = 'Horizon thickness (cm)') +
  theme_minimal()

```
:::

## Boundary shape {#sec-pr-bdyshp}

Horizon boundary shape can only be observed on wider exposures (see @tbl-pr-gear, type [E]{.ceg}, [P1]{.ceg}, [P2]{.ceg}). Record using one of the options in @tbl-hr-bshp. Note that the boundary shape is associated with the lower boundary of any given horizon, so the lowest horizon described on a profile may have an undefined boundary shape if profile exposure stopped early (@sec-pr-rsa). In that case, use code [N]{.ceg}.

```{r}
#| label: tbl-hr-bshp
#| tbl-cap: "Horizon boundary shape"

dat_hr_bshp <-
  tribble(~Code, ~Name, ~Description,
          'S', 'Smooth', 'the boundary surface is a plane with few or no irregularities and usually occurs at the same depth across the profile face', 
          'W', 'Wavy', 'the boundary surface has broad, shallow, relatively regular pockets and none deeper than they are wide',
          'I', 'Irregular', 'the boundary surface has pockets which are deeper than they are wide but not recurved',
          'C', 'Convolute', 'the boundary surface has pockets which are deeper than they are wide and, in parts, recurved',
          'D', 'Discontinuous', 'the boundary is discontinuous, usually due to external disturbance',
          'N', 'Not observed', 'the lower horizon boundary could not be observed')

tbl_hr_bshp <- gt(dat_hr_bshp) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red', 
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_hr_bshp
```

## Boundary distinctness {#sec-pr-bdydist}

Horizon boundaries mark changes in soil properties over short vertical distances. The distinctness is variable, from very sudden to gradual, and can signify particular soil characteristics (e.g. a change in parent material). Horizon boundary distinctness is used when classifying soils. Measure the distinctness of a horizon's lower boundary in terms of its width. Report a median and range in 0.1 centimetre increments (e.g. [2.0 cm +/- 0.5 cm]{.ceg}).

For rapid assessment, horizon boundary distinctness can be noted using one of the options in @tbl-hr-bdst.

```{r}
#| label: tbl-hr-bdst
#| tbl-cap: "Horizon boundary distinctness"

dat_hr_bdst <-
  tribble(~Code, ~Name, ~Description,
          'S', 'Sharp', '< 0.5 cm', 
          'A', 'Abrupt', '≥0.5–<2.0 cm',
          'C', 'Clear', '≥2.0–<5.0 cm',
          'G', 'Gradual', '≥5.0–<10.0 cm',
          'D', 'Diffuse', '≥10–<30.0 cm',
          'N', 'Not observed', 'the lower horizon boundary could not be observed')

tbl_hr_bdst <- gt(dat_hr_bdst) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_hr_bdst
```

Where a transition is thicker than 30 cm, it should be defined as a new horizon. Use transitional horizon names as needed (see @sec-hor-tran).

[put a diagram unifying the previous three concepts here]{.diags}

## Infill features {#sec-pr-inf}

Infill features are interruptions to the horizonation pattern caused by relatively recent biological or physical disturbances. Infill features are large enough that they can usually only be clearly observed at exposures or large pit faces. The interruption may extend across several horizons, and the infilled material may be from adjacent horizon(s) or comprise reworked material from the affected horizon(s). Where infill features are present, note the type using the options in @tbl-hr-infl, the horizons affected, and the infill material using the options in @tbl-hr-infm.

E.g., [C (1-3) T]{.ceg} for infilled shrink-swell cracks extending from the surface to the third horizon.

Note that open subsurface voids should be described as per @sec-void, and open surface cracks as per @sec-pr-crack.

```{r}
#| label: tbl-hr-infl
#| tbl-cap: "Infill feature types"

dat_hr_infl <-
  tribble(~Code, ~Name, ~Description,
          'C', 'Crack',  'Infilled surface crack caused by shrink-swell clay movement',
          'T', 'Tunnel', 'Infilled animal burrow > 20 mm diameter',
          'B', 'Burrow', 'Infilled animal burrow <= 20 mm diameter', 

          'R', 'Root',   'Infilled root channel',
          'D', 'Drain',  'Infilled mole drain or similar deep-tillage feature')

tbl_hr_infl <- gt(dat_hr_infl) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(15))

tbl_hr_infl
```

```{r}
#| label: tbl-hr-infm
#| tbl-cap: "Infill feature material"

dat_hr_infm <-
  tribble(~Code, ~Name, ~Description,
          'T', 'Topsoil',  'Infilled by topsoil material (in non-topsoil horizons)',
          'S', 'Subsoil',  'Infilled by subsoil material (in topsoil or regolith horizons)',
          'O', 'Organic',  'Infilled by organic material',
          'L', 'Local',    'Infilled by reworked material from the same horizon',
          'M', 'Mixed',    'Infilled by mixed materials from multiple horizons',
          'U', 'Unknown',  'Infill material cannot be determined with confidence')

tbl_hr_infm <- gt(dat_hr_infm) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(15))

tbl_hr_infm
```

## Soil moisture status {#sec-pr-sws}

Report soil moisture status for each horizon at the time of observation. This parameter supports assessments of drainage and permeability, and contextualises consistence tests.

```{r}
#| label: tbl-hr-soilmst
#| tbl-cap: "Soil water status (adapted from @griffiths1985)."

dat_hr_bdst <-
  tribble(~Code, ~Name, ~Sands, ~Light, ~Medium, ~Clays,
          'D', 'Dry', 'Loose, single grain', 'Loose', 'Easily broken down to powder', 'Hard, baked, cracked',
          'T', 'Moderately moist', 'Will not form a ball', 'Forms weak ball, breaks easily', 'Forms a ball, very pliable', 'Forms a ball, somewhat pliable',
          'M', 'Moist', 'Forms very weak ball', 'Forms weak ball, breaks easily', 'Forms a ball, very pliable', 'Easily forms a ball',
          'W', 'Wet', 'Fluid, non-plastic, non-sticky', 'Slightly fluid, slightly plastic', 'Deformable, plastic', 'Semi-deformable, very plastic',
          'S', 'Saturated', 'Water films visible, or below water table',  '',  '', '')

tbl_hr_bdst <- gt(dat_hr_bdst) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  tab_style(style = list(cell_text(align='center', style='italic')),
            locations = cells_body(columns = 3:6, rows = 5)) |>
  tab_spanner(label = 'Behaviour of the fine earth fraction',
              columns = Sands:Clays) |>
  cols_label(Sands = '> 80% Sand',
             Light = html("< 80% Sand<br />< 18% Clay"),
             Medium = html('> 18% Clay<br />< 35% Clay'),
             Clays = '>35% Clay') |>
  cols_merge(columns = 3:6, hide_columns = FALSE, rows = 5 ) |>
  cols_width(Code ~pct(10), Name ~pct(15))

tbl_hr_bdst
```

## Depth to free water

Depth to free water can be determined from the upper depth of the uppermost [S]{.ceg} saturated horizon (@tbl-hr-soilmst), or may be recorded directly, in whole centimetres. Depending on the soil's transmissivity, the water table may be also become evident by the pit or core filling with water. In this case, wait for the water level to stabilise before measuring.

-   If the water table is not encountered, record [NA]{.ceg}
-   If the water table is encountered, use @tbl-hr-frw to code its position in the profile, then record its depth above or below the soil surface in whole centimetres. Do not record negative numbers to signify a water table above the surface.
-   If the water table is exactly at the soil surface, record ['B 0']{.ceg}.

e.g. ['B 80']{.ceg} for a profile saturated below 80 cm depth; ['A 1']{.ceg} for a saturated and barely-walkable peat.

```{r}
#| label: tbl-hr-frw
#| tbl-cap: "Free water location"

dat_hr_frw <-
  tribble(~Code, ~Name, ~Description,
          'NA', 'Not applicable', 'Water table not encountered',
          'B', 'Below', 'Water table observed below the soil surface',
          'A', 'Above', 'Water table observed at or above the soil surface')

tbl_hr_frw <- gt(dat_hr_frw) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red', 
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_hr_frw
```

## Parent materials {#sec-pr-pm}

Information about the lithology, origin and weathering degree of the soil parent materials is essential for accurate classification and helps contextualise analytical data.

### Lithology {#sec-pr-lith}

The lithology of the fine earth (mineral particles \<2 mm) and of rock fragments (if present) are noted separately for each soil horizon. Where subsolum layers are observed, their lithology is also noted. Use the codes in @tbl-ss-pm.

Example: [An, Hs]{.ceg} for a mix of eroded sedimentary rocks and volcanic ash.

### Weathering {#sec-pr-wdeg}

The effects of weathering and pedogenesis vary down-profile (usually, but not always, decreasing with depth). Horizons that have experienced limited weathering may require specific descriptors (e.g. texture (@sec-tex-rec) cannot usually be determined on paralithic material, but bedding (@sec-reg-bed) may need to be described), so the need for these should be assessed early in the description process. Use the guidance in @tbl-pr-wdeg to clearly identify the solum, as well as any subsolum materials encountered. The weathering status of rock material specifically may be assessed using the schema in @sec-rockfrag-wth.

+------------+----------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Code       | Name                 | Description                                                                                                                                                                                                                                                                                                              |
+:===========+:=====================+:=========================================================================================================================================================================================================================================================================================================================+
| [So]{.ceg} | Soil material        | Horizon characteristics include any of:                                                                                                                                                                                                                                                                                  |
|            |                      |                                                                                                                                                                                                                                                                                                                          |
|            |                      | -   anthropic or pedogenic structure (@sec-str-origin),                                                                                                                                                                                                                                                                  |
|            |                      | -   elevated biological activity,                                                                                                                                                                                                                                                                                        |
|            |                      | -   formation of pedogenic features (@sec-secondary), *and*                                                                                                                                                                                                                                                              |
|            |                      |                                                                                                                                                                                                                                                                                                                          |
|            |                      | a lack of residual geogenic structure.                                                                                                                                                                                                                                                                                   |
+------------+----------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [Rg]{.ceg} | Regolithic material  | 1.  Horizon has geogenic structure in ≥ 50% (by volume) and pedogenic structure in \< 50% of the volume of the fine earth fraction (or is apedal), *and*                                                                                                                                                                 |
|            |                      | 2.  shows evidence of the accumulation of transported material by:                                                                                                                                                                                                                                                       |
|            |                      |     a)  the presence of a lithological discontinuity at its base, *or*                                                                                                                                                                                                                                                   |
|            |                      |     b)  consisting of a mixture of rock fragments that does not have either the same lithology, or size, or shape, or weathering status and that can easily be dug with a spade or with a pickaxe, if the layer is slightly consolidated, or when air-dry, more than 50% (by volume) slakes in water within 1 hour, *or* |
|            |                      | 3.  is composed of openwork rock fragments.                                                                                                                                                                                                                                                                              |
+------------+----------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [Sr]{.ceg} | Saprolithic material | 1.  Horizon has geogenic structure in ≥ 50% (by volume) and pedogenic structure in \< 50% of the volume of the fine earth fraction (or is apedal), *and*                                                                                                                                                                 |
|            |                      | 2.  shows no evidence of accumulation of transported material, *and*                                                                                                                                                                                                                                                     |
|            |                      | 3.  material can easily be dug with a spade or with a pickaxe if the material is slightly consolidated in \>75% of the volume, or when air-dry, more than 50 % (by volume) will slake in water within 1 hour.                                                                                                            |
|            |                      |                                                                                                                                                                                                                                                                                                                          |
|            |                      | Roots may be present in the matrix and/or along remnant geogenic features.                                                                                                                                                                                                                                               |
+------------+----------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [Pr]{.ceg} | Paralithic material  | 1.  Horizon has \>50% (by volume) of geogenic structure, *and*                                                                                                                                                                                                                                                           |
|            |                      | 2.  is coherent to the extent that the material is difficult to dig with a spade (except for material occurring in-between geogenic features which may be chipped or scraped), *and*                                                                                                                                     |
|            |                      | 3.  air-dry samples (25–30 mm) do not slake in water for 1 hour (except for material occurring in between geogenic features), *and*                                                                                                                                                                                      |
|            |                      | 4.  shows signs of mechanical weakening in one or more of the following ways:                                                                                                                                                                                                                                            |
|            |                      |     a)  occurrence of roots along geogenic features like cracks, sedimentary bedding or metamorphic foliations but not in the matrix, *or*                                                                                                                                                                               |
|            |                      |     b)  to the naked eye visible fillings of *in situ* weathered materials, clays and/or other material from the above layer or horizon along geogenic features like cracks, sedimentary bedding or metamorphic foliations.                                                                                              |
+------------+----------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [Rr]{.ceg} | Lithic material      | 1.  Horizon has \> 90% (volume) of geogenic structure, *and*                                                                                                                                                                                                                                                             |
|            |                      | 2.  its coherence is such that the material cannot be dug with a spade, or air-dry samples (25–30 mm) will not slake in water within 1 hour, *and*                                                                                                                                                                       |
|            |                      | 3.  has one or more of the following:                                                                                                                                                                                                                                                                                    |
|            |                      |     a)  no visible chemical or physical weathering (to the naked eye) along geogenic features along cracks, sedimentary bedding or metamorphic foliations, *or*                                                                                                                                                          |
|            |                      |     b)  when chemical or physical weathering is visible (to the naked eye), it is only expressed by a discoloration or staining along geogenic features like cracks, sedimentary bedding or metamorphic foliations and representing ≤ 25% of the visible material, *and*                                                 |
|            |                      | 4.  has no roots - neither along geogenic features nor in the matrix.                                                                                                                                                                                                                                                    |
+------------+----------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

: Weathering status of soil-forming materials, adapted from @juilleret2016. {#tbl-pr-wdeg .striped .hover tbl-colwidths=\[10,20,70\]}

::: {#nte-stp .callout-note collapse="true" appearance="simple" title="Degree of weathering and the New Zealand Soil Classification"}
For the purposes of classifying soils according to the New Zealand Soil Classification, profiles overlying a layer of Lithic material as defined in @tbl-pr-wdeg would qualify for the 'lithic contact', while profiles overlying Saprolithic or Paralithic material would qualify for the 'paralithic contact'.
:::

### Origin {#sec-pr-orig}

Parent material origin, or mode of emplacement, describes how soil parent materials have arrived at the point of observation. Multiple modes may have operated in different parts of the soil profile. For soil materials that have weathered from rocks *in situ*, the mode of emplacement is recorded as the last major formation process that left the rocks in a place where they could start weathering into soil.

Record one parent material origin code per soil horizon, using the codes in @tbl-ss-pmo. At the horizon level, the [An]{.ceg} Anthropic class can be subdivided by origin using the codes in @tbl-hr-antype (e.g. [Ann]{.ceg} for a fill pad in a residential housing development).

```{r}
#| label: tbl-ss-pmo
#| tbl-cap: "Parent material origin"

dat_ss_pmo <-
  tribble(~Code, ~Name, ~Description,
          'An', 'Anthropic', 'Deposits made by the direct actions of humans, including truncation, mixing, or deposition.',
          'Cl', 'Colluvium', 'Unconsolidated and unsorted, usually weathered soil and rock material deposited on lower slopes, transported primarily by gravity assisted by water. These deposits may comprise slow accumulations or result from sudden slope failures',
          'Db', 'Debris avalanche', 'Deposits derived from the large-scale collapse of unstable slopes of stratovolcanoes, forming mounds and hummocky ground that can be unconsolidated or heat-welded.',
          'Fl', 'Alluvium', 'Sediments that have been deposited by streams, rivers and other running water.',
          'Gl', 'Glacial till', 'Poorly stratified, poorly sorted rock fragments, sand and mud, surface or near-surface deposits resulting from the transportation by and deposition from ice or meltwater from beneath or in close proximity to glacial ice.',
          'Lc', 'Lacustrine', 'Formed in and around lake bed deposits; comprises extremely fine sediment deposited under very low-flow and usually freshwater conditions.',
          'Mr', 'Marine' , 'Unconsolidated sediments saturated by brackish or saline water.',
          'Lh', 'Lahar', 'A flow of heterogenous volcaniclastic material mixed with water and deposited rapidly in river valleys associated with stratovolcanoes. Lahars may be triggered by volcanic eruption or by cone collapse in response to oversteepening.',
          'Tp', 'Tephra', 'Unconsolidated primary pyroclastic products of explosive volcanic eruptions encompassing all grain sizes (ash, lapilli and larger) and lithologic compositions. Both airfall and non-welded pyroclastic flow deposits are included.',
          
          'Lo', 'Loess', 'A blanket deposit of silt-sized particles (0.002--0.06 mm diameter); usually carried by wind from dry riverbeds or outwash plains during glacial and post-glacial periods.',
          'Sa', 'Aeolian sand', 'Wind-deposited sand-sized particles (0.05--2.0 mm diameter), i.e. dune sand.',
          'Sf', 'Solifluction', 'Unconsolidated, unsorted, usually weathered soil and rock material mantling hillslopes and derived from slow, viscous, gravity-driven downslope movement of waterlogged soil amd materials in terrains underlain by frozen soil or permafrost, either now or in the past.',
          'Ex', 'Extruded rock', 'Volcanic rocks that were emplaced on the land surface in a molten state',
          'In', 'Intruded rock', 'Plutonic rocks that were emplaced beneath the land surface in a molten state',
          'Rx', 'Indurated rock', 'Uplifted rocks with strong sedimentary induration and/or metamorphosis',
          'Rw', 'Non-indurated rock', 'Uplifted sedimentary rocks with weak to moderate induration and no metamorphosis',
          'Li', 'Litter', 'Plant material decomposing under intermittently wet to dry conditions',
          'Pt', 'Peat', 'Plant material decomposing under saturated conditions',
          'Uk', 'Unknown', 'Origin cannot be determined with certainty.'
          ) %>% 
  mutate(Type = c(rep('Transported', 12), rep('Residual', 4), 
                  rep('Organic', 2), 'Other'))

tbl_ss_pmo <- gt(group_by(dat_ss_pmo, Type)) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(weight = 'bold')),
            locations = cells_row_groups()) |>
  tab_style(style = list(cell_text(color  = 'red', 
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  fmt_markdown(columns = 4) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_ss_pmo
```

```{r}
#| label: tbl-hr-antype
#| tbl-cap: "Subtypes of Anthropic soil materials"

dat_hr_antype <-
  tribble(~Code, ~Name, ~Description,
          'm', 'Historic Māori', 'Soil modifications resulting from Māori cultural practices dating from Polynesian arrival in New Zealand, particularly agricultural practices',
          'p', 'Historic Pakeha', 'Soil modifications dating from Pakeha (non-Māori) arrival in New Zealand up to modern times (e.g. non-mechanised European agricultural and mining practices)',
          'n', 'Modern', 'Soil modifications resulting from modern mechanised agricultural practices, other land alteration practices (e.g. mining, terracing in urban environments) and/or addition of industrially produced materials to the soil (e.g. plastics).'
          )

tbl_hr_antype <- gt(dat_hr_antype) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red',
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_hr_antype
```
