---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Parent material

```{r}
library(dplyr)
library(tibble)
library(gt)
library(gtExtras)
```

Parent material can be explored in terms of geological units, lithology (rock type), and mineralogy.

## Geology

Geological mapping, particularly surface and near-surface stratigraphy, is an essential resource for understanding the landscape. However, the scale and focus of the available products don't always align with soil observation needs. Some caution is needed when interpreting geological maps in a soils context. Keep the following in mind:

-   Read the accompanying reports for a given area as well as looking at the maps. They contain a wealth of contextual information.
-   When working in a depositional area, get familiar with the 'source' geologies - upstream, uphill, upwind. Those will be the main contributors to the parent material, which will often be mapped with closer reference to age and particle size than lithology.

Geology mapping in New Zealand can be accessed through resources provided by [GNS Science - Te Pū Ao](https://www.gns.cri.nz/). This includes QMap, a unified 1:250,000 scale geological map of the country, and a large set of legacy geology maps produced at scales as detailed as 1:25,000.

## Lithology

The specific lithology, or rock type, of the parent material offers a more specific and soil-relevant categorisation than geology mapping generally supplies. The following options are in common use in New Zealand.

### NZLRI Rock Type

An identification and coding system for New Zealand rock types is available in [@lynn1991]. This system is tuned for land resource assessment, particularly assessment of erosivity, and is used in the NZ Land Resource Inventory and Land Use Classification methodology [@lynn2009]. The system is mentioned here as an option where consistency with LUC mapping is desired.

[repeat codes, or just xref??]{.todo}

### NZSC Parent Material {#sec-ss-pmat}

Level 4 of the NZSC, the soil Family [@webb2011], uses a shorter list of lithology codes along with a separate 'origin' code, whereas the NZLRI rock type conflates the two concepts. The lithology codes are available in @tbl-ss-pm, and the origin codes in @tbl-ss-pmo.

## Other lithology classifications

### Geological vocabularies

Several research-grade vocabularies for describing geology have been developed in recent years, including @geoscienceaustralia2016 and @mccormick2023. Both offer a detailed set of rock and mineral definitions in a hierarchical relationship structure, but both are currently missing one or more terms essential to the New Zealand context. Additionally, the hierarchies don't always align cleanly with those used in New Zealand geology.

### Silica-content classification

Soil parent material classifications have largely focused on major geological origin (igneous, metamorphic, sedimentary), grain size, and some obvious compositional classes. [@gray2016] contend that the major differentiator for soil formation should first be silica (SiO~2~) content and propose an alternate lithology classification per @tbl-ss-sio2lith below. This is optional to record as it has not been extensively tested.

```{r}
#| label: tbl-ss-sio2lith
#| tbl-cap: "Silica-content Lithology classification"

dat_ss_sio2lith <-
  tribble(~Code, ~Name, ~ Description,
           1, 'Extremely siliceous', 'Quartz sands, chert, pure quartzite, jasper, quartz reefs, silicified rocks',
           2, 'Siliceous Upper', 'Quartz sandstone, quartz siltstone, unqualified quartzite and alluvial sands',
           3, 'Siliceous Mid', 'Granite, rhyolite and siliceous tuff, arkose sandstone, most unqualified sandstone',
           4, 'Siliceous Lower', 'Adamellite, granodiorite, dacite, monzogranite, siliceous/ intermediate tuff, most greywacke & lithic
sandstone, unqualified siltstone',
           5, 'Intermediate upper', 'Syenite, trachyte, most argillaceous rocks (mudstone, claystone, shale, slate, phyllite and schist),
alluvial loams and non-cracking clays',
           6, 'Intermediate lower', 'Monzonite, trachy-andesite, diorite, andesite, intermediate tuff, alluvial cracking clays (not black)',
           7, 'Mafic', 'Gabbro, dolerite, basalt, mafic tuff, amphibolite, alluvial black cracking clays',
           8, 'Ultramafic', 'Serpentinite, dunite, peridotite, tremolite-chlorite-talc schists',
           9, 'Calcareous', 'Limestone, dolomite, calcareous shale, calcareous sands',
          10, 'Sesquioxide', 'Laterite, bauxite, ferruginous sandstone, ironstone',
          11, 'Organic', 'Peat, coal, humified vegetative matter',
          12, 'Evaporite', 'Gypsum, halite, anhydrite')
        
tbl_ss_sio2lith <- gt(dat_ss_sio2lith) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10), Name ~pct(20))

tbl_ss_sio2lith
```

This system can be applied to the New Zealand context relatively easily, although class [10]{.ceg} may be very rare. Iron sands would map to one of classes [6]{.ceg} to [8]{.ceg}, depending on deposit purity.

-   Folk 1970, blair and mcpherson 1999
-   juilleret subsolum stuff
