# Landscape

```{r}
library(dplyr)
library(tibble)
library(gt)
library(gtExtras)
```

-   setup - going big to small
-   purpose - interpretive context
-   note: discuss when a DEM/terrain derivatives can and can't be used post-field

## Regional landscape

-   scale: 100s-1000s of meters (ref Ontario guide for this idea)
-   don't have a nationally accepted ecozone system yet
-   do have LUC (brief discussion)
-   do have SMap landform trees (brief discussion)

### Relief

### Modal slope

### RMS

-   adapt from aus

### Mesoscale topographic index

-   ref to RECCE, connect to Geomorphons concept

### Landscape description

-   shortcut terms, noting that they don't form a conceptual tesselation
-   longform descriptions

------------------------------------------------------------------------

## Local landform

-   scale: 1-10s of meters around site
-   starting with simple measurements and moving to more interpretive ones

### Slope

Slope is the inclination of the land surface relative to the horizontal plane. Slope steepness affects many soil processes, notably water runoff and infiltration behaviour.

Direct measurements of slope should be taken at the landform scale over a minimum distance of 20 meters. Stand 10m upslope from the soil site and measure downhill. Record in units of degrees, with a maximum precision of 0.1. Precision needs will vary with slope and survey purpose; whole number degrees are fine on any slope over about 5 degrees, but finer gradations can be useful for defining low-relief landforms, e.g. on alluvial plains.

Slope classes may be applied to the recorded data; some options are presented in @sec-slopeclass.

-   TODO Note about getting slope from DEM

### Aspect

Aspect is the direction a slope faces. Slope aspect helps determine exposure to sun, rain and prevailing winds, and so influences the biome growing in a particular soil. In some environments it also helps determine how much new material is received by aeolian or fluvial deposition.

Aspect should be measured at the same time and over the same transect as slope. Record aspect in whole degrees clockwise from North, using the range \[0-359\] (360 == 0). Use NA for flat ground. Use NR if unable to record.

For rapid assessment, record compass bearing on an 8 or 16 point scale (e.g. NE, SSE).

-   TODO Note: defining 'flat' in soil survey.
-   TODO Note: map vs magnetic north, and consequences for field-measured vs DEM-derived aspect measurements

### Landform element morphology

The following is adapted from [@nationalcommitteeonsoilandterrain2009, pp. 19-26].

Landform elements can be assigned to a morphological type - one of a series of raw shapes - at the scale of observation. These are:

```{r}
#| label: tbl-hr-elemmorph
#| tbl-cap: "Landform Element Morphologies"

dat_hr_elemmorph <-
  tribble(~Code, ~Name,
          'CR', 'Crest',
          'RI', 'Ridge',
          'SL', 'Slope',
          'FL', 'Flat',
          'DO', 'Open depression',
          'DC', 'Closed depression')

# TODO add some little graphics in a third column

tbl_hr_elemmorph <- gt(dat_hr_elemmorph) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(20))

tbl_hr_elemmorph

```

Note that these are paired mirror images, e.g. a closed depression is the opposite of a crest. Each morphological type extends to a break of slope or point of 0-curvature.

Slopes can be further refined by including additional information appended about their relationship to their neighbouring elements. Up and down slope, four options are available.

```{r}
#| label: tbl-hr-slopeinc
#| tbl-cap: "Slope element relative inclination"

dat_hr_slopeinc <-
  tribble(~Code, ~Name, ~Description,
          'X', 'Waxing', 'element upslope is gentler, element downslope is steeper.',
          'N', 'Waning', 'element upslope is steeper, element downslope is gentler.',
          'A', 'Maximal', 'element upslope is gentler, element downslope is gentler.',
          'I', 'Minimal', 'element upslope is steeper, element downslope is steeper.')

tbl_hr_slopeinc <- gt(dat_hr_slopeinc) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(20), Name ~pct(20))

tbl_hr_slopeinc

```

Two more are available across the slope.

```{r}
#| label: tbl-hr-slopeenc
#| tbl-cap: "Slope element relative enclosure"

# I'm stepping into new territory here D:

dat_hr_slopeenc <-
  tribble(~Code, ~Name, ~Description,
          'L', 'Enclosed', 'Adjacent slope elements face towards each other',
          'E', 'Exposed', 'Adjacent slope elements face away from each other')

tbl_hr_slopeenc <- gt(dat_hr_slopeenc) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(20), Name ~pct(20))

tbl_hr_slopeenc

```

This combination of properties gives rise to more complex shapes. For example, an open hollow could be described as an enclosed maximal slope ([**L-A-SL**]{style="color: red"}). A shoulder would be an exposed waxing slope ([**E-X-SL**]{style="color: red"}). Adding further information about slope and aspect, once classified, can define a complete element. For instance a gently sloping, northeast facing fan - GE NE E N SL.

-   TODO add graphics

### Slope length

The length of a slope element is measured downslope, and provides information on...

-   mention RUSLE connection

### Active geomorphic agents

-   eroded vs aggraded then types

### Landform description

-   shortcut terms, noting that they don't form a conceptual tesselation
-   longform descriptions
