# Parent material {#sec-pm}

```{r}
library(dplyr)
library(tibble)
library(gt)
```

Parent material can be explored in terms of geological units, lithology, and mineralogy.

## Geology {#sec-pm-geol}

Geological mapping, particularly surface and near-surface stratigraphy, is an essential resource for understanding the landscape. However, the scale and focus of the available products don't always align with soil observation needs. Some caution is needed when interpreting geological maps in a soils context. Keep the following in mind:

-   Read the accompanying reports for a given area as well as looking at the maps. They contain a wealth of contextual information.
-   When working in a depositional area, get familiar with the 'source' geologies - upstream, uphill, upwind. Those will be the main contributors to the parent material, which will often be mapped with closer reference to age and particle size than the lithology shown in the geological map unit descriptions.
-   In landscapes where layered deposits make up the uppermost materials of the land as unconsolidated cover beds (such as in central North Island), it is common for such cover beds to be missing from geological maps (although they will likely be noted in the accompanying text, e.g., see Leonard et al. 2010). This because the map units selected in geological mapping tend to reflect the underlying, large-scale deposits that have been landforming in emplacement.

Geology mapping in New Zealand can be accessed through resources provided by [Earth Sciences New Zealand](https://earthsciences.nz/). These include QMap, a unified 1:250,000 scale geological map of the country, and a large set of legacy geology maps produced at scales as detailed as 1:25,000.

## Lithology {#sec-pm-lith}

The specific mineral characteristics of the soil parent material, or its lithology, offers a more specific and soil-relevant categorisation than geology mapping generally supplies. Lithology is generally recorded when describing the soil's fine earth fraction (mineral material <2 mm), surface and in-profile rock fragments (mineral material >2 mm), subsolum rock, and profile-adjacent rock outcrops. @tbl-ss-pm provides a short list of parent material lithologies compatible with the NZSC Family [@webb2011].

```{r}
#| label: tbl-ss-pm
#| tbl-cap: "NZSC Lithology"

dat_ss_pm <-
  tribble(~Code, ~Name,
    'Cb',	'Carbonaceous Rock (Coal)',
    'Sq',	'Soft Quartzitic Sediments',
    'Hq',	'Hard Quartzitic Sediments',
    'Sm',	'Soft mudstone',
    'Ss',	'Soft sandstone',
    'Hm',	'Hard mudstone',
    'Hs',	'Hard sandstone',
    'Tm',	'Tuffaceous mudstone',
    'Ts',	'Tuffaceous sandstone',
    'Sc',	'Schist',
    'Ls',	'Soft calcareous',
    'Li',	'Hard calcareous',
    'Ma',	'Marble',
    'Rh',	'Rhyolite',
    'Ig',	'Ignimbrite',
    'An',	'Andesite',
    'Ba',	'Basalt',
    'Gr',	'Granite and Gneiss',
    'Di',	'Diorite',
    'Ga',	'Gabbro',
    'Fe',	'Iron Sands')
      
tbl_ss_pm <- gt(dat_ss_pm) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red', 
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10))

tbl_ss_pm

```

::: {.content-visible when-format="docx"}
delete me
:::

::: {#nte-softrock .callout-note appearance="minimal" collapse="true" title="What are soft rocks?"}
'Soft' rocks in this context can at most be broken readily with a light hammer blow, and can be scraped and peeled by a pocket knife. 'Hard' rocks can at best be scratched with a knife but not scraped or peeled, and require a firm hammer blow to crack [@lynn1991, p. 11].
:::

### Other lithology classifications {#sec-pm-litho}

#### Geological vocabularies {#sec-pm-vocabs}

Several research-grade vocabularies for describing geology have been developed in recent years, including @geoscienceaustralia2016 and @mccormick2023. Both offer a detailed set of rock and mineral definitions in a hierarchical relationship structure, but both are currently missing one or more terms essential to the New Zealand context. Additionally, the hierarchies don't always align cleanly with those used in New Zealand geology. The vocabularies are expected to develop over time and may be recommended in future editions of this handbook.

Other, simpler lithological vocabularies available in New Zealand and elsewhere often conflate lithology with mode of deposition and particle size when describing unconsolidated, transported materials (e.g. NZLRI rock types, @lynn1991 and WRB parent material, @iussworkinggroupwrb2022). Parent material origin and particle size distribution are described separately for soil materials in this handbook (@sec-pr-orig, @sec-tex). However, workers describing soils as part of an LUC assessment may still wish to use the NZLRI rock types list for compatibility reasons [@lynn1991].

#### Silica-content classification {#sec-pm-lithsi}

Soil parent material classifications have largely focused on major geological origin (igneous, metamorphic, sedimentary), grain size, and some obvious compositional classes. [-@gray2016] contend that the major differentiator for soil formation should first be silica (SiO~2~) content and propose an alternate lithology classification on that basis.

## Mineralogy

The mineralogy classes of the US Soil Taxonomy [@soilsurveystaff2022] may be estimated if known, but these will require laboratory confirmation and the ability to classify the profile against US Soil Taxonomy at least to Great Group level.

The New Zealand Soil Classification [@hewitt2010] references an older version of the Soil Taxonomy mineral classes, adapted to New Zealand conditions and extensively tested [@childs1990]. The classes are mentioned in relation to soil orders but are not used in the key or the diagnostic materials. Optionally, the dominant mineral class may be recorded on a per-horizon basis using the key in @tbl-ss-nzscmin. Estimates may be made using knowledge of soil parent materials and local soil-forming processes, but should be confirmed in the laboratory.

```{r}
#| label: tbl-ss-nzscmin
#| tbl-cap: "New Zealand soil mineralogy classification (reproduced from @childs1990, Table 1)"

dat_ss_nzscmin <-
  tribble(
    ~Code, ~Name, ~ Description, ~size,
    'Gy', 'Gypsic', '\\>15% by weight of gypsum', '< 20 mm',
    'Ca', 'Carbonatic', '\\>30% by weight carbonates (as CaCO~3~)',  '< 20 mm',
    'Sf', 'Superferritic', '\\> 28% citrate-dithionite extractable Fe (≥ 40% as Fe~2~O~3~)', '< 2 mm',
    'Fe', 'Ferritic', '\\> 7.0% citrate-dithionite extractable Fe (≥ 10% as Fe~2~O~3~)',  '< 2 mm',
    'Al', 'Aluminitic', '\\> 40% by weight aluminium hydroxides and oxyhydroxides',  '< 2 mm',
    'Si', 'Siliceous', '\\> 90% by weight silica (SiO~2~) minerals (quartz, chalcedony, or opal) and other durable minerals that are resistant to weathering, provided that the silica minerals are > 60% by weight',  '< 2 mm',
    'Am', 'Amorphous', '\\> 5% by weight of minerals with short-range order (principally allophane, imogolite, and ferrihydrite)',  '< 2 mm',
    'Mi', 'Micaceous', '\\> 40% by weight of mica minerals',  '< 2 mm',
    'Cl', 'Chloritic', '\\> 40% by weight of chlorite',  '< 2 mm',
    'Sp', 'Serpentinitic', '\\> 40% by weight of serpentine minerals (antigorite, chrysotile, fibrolite, and talc)',  '< 2 mm',
    'Sm', 'Smectitic', 'More smectites than any other mineral group or single mineral (if mineral group not applicable, e.g. crandallite, chlorite)', '< 0.002 mm',
    'Vm', 'Vermiculitic', 'More vermiculites than any other mineral group or single mineral', '< 0.002 mm',
    'Il', 'Illitic', 'More illites than any other mineral group or single mineral', '< 0.002 mm',
    'Ka', 'Kandic', '\\> 50% by weight Kaolin-group minerals (kaolinite, halloysite, dickite, and nacrite) *and* < 10% by weight smectites', '< 0.002 mm',
    'Cm', 'Clay-mineralic', 'Other soils with > 10% clay. The dominant mineral or mineral group in the clay fraction is named (e.g. clay-mineralic (crandallite))', '< 0.002 mm',
    'Gl', 'Glassy', '\\> 40% by weight volcanic glass', '< 2 mm',
    'Wh', 'Whole-soil-mineralic', '\\> 40% by weight of any mineral or mineral group, other than silica minerals or feldspars', '< 2 mm',
    'Ot', 'Other soils', '', 'Not applicable')
        
tbl_ss_nzscmin <- gt(dat_ss_nzscmin) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red', 
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  tab_row_group(label = 'All soils', rows = 1:10, 
                id ='all') |>
  tab_row_group(label = 'Other soils with ≥ 10% clay', rows = 11:15,
                id = 'clay') |>
  tab_row_group(label = 'Other soils with < 10% clay', rows = 16:18,
                id = 'not') |>
  row_group_order(groups = c('all', 'clay', 'not')) |>
  tab_style(style = list(cell_text(weight = 'bold')),
            locations = cells_row_groups()) |>
  fmt_markdown(columns = 3) |>
  cols_label(size = 'Determinant size fraction') |>
  tab_footnote(footnote = 'In New Zealand the following are regarded as "durable minerals": beryl, corundum, ilmenite, magnetite, tourmaline group, zircon.',
               locations = cells_body(columns = 3, rows = 6)) |>
  tab_style(style = list(cell_text(align = 'left')),
                         locations = cells_footnotes()) |>
  cols_width(Code ~pct(10), Name ~pct(20), Description ~pct(50))

tbl_ss_nzscmin
```
