# Climate and Weather

## Climate

-   note: most of this is better off not measured in-field
-   note: limitations of available climate data, use local records where possible
-   talk about getting data from LENZ vs more local sources
-   calculation from terrain for insolation

### Global-Regional

The Köppen-Geiger climate classification [@köppen1936] is the most well-known and accepted global system, and has been periodically updated (e.g. [@kottek2006; @beck2018; @beck2023]). Current maps and projections under various climate change scenarios are available from <https://www.gloh2o.org/koppen/>.

In and around New Zealand, the Köppen regions that currently occur are listed in @tbl-ss-koppenz.

```{r}
#| label: tbl-ss-koppenz
#| tbl-cap: "Köppen climate regions of New Zealand and its neighbours"

dat_ss_koppenz <-
  tribble(~Code, ~Name, ~Locations,
          'Af', 'Tropical, rainforest', 'Pacific Islands e.g. Tonga, Samoa',
          'Am' , 'Tropical, monsoon', 'Northwest coast of Fijian islands',
          'Bsk', 'Arid, steppe, cold', 'Central Otago, Melbourne, Northwest Victoria',
          'Cfa', 'Temperate, no dry season, hot summer', 'Coastal New South Wales and southern Queensland',
          'Cfb', 'Temperate, no dry season, warm summer', 'Majority of New Zealand; Chatham Islands; coastal southeastern Australia; Tasmania',
          'Cfc', 'Temperate, no dry season, cold summer', 'Small areas of North Island e.g. upper Ruahines; lower mountains in South Island e.g. Peel Range, Nelson',
          'Dfb', 'Cold, no dry season, warm summer', 'Sheltered pockets in the Southern Alps',
          'Dfc', 'Cold, no dry season, cold summer', 'Lower slopes of major volcanoes and the Southern Alps',
          'ET', 'Polar, tundra', 'Upper slopes of major volcanoes and the Southern Alps',
          'EF', 'Polar, frost', 'Antarctica')

tbl_ss_koppenz <- gt(dat_ss_koppenz) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10))

tbl_ss_koppenz
```

!!! map goes here !!!

Since most work programs will take place within one zone, regional climate rarely needs to be noted in the field.

Where a study area appears to deviate from a regional map, the extent and nature of the deviation should be discussed in project reporting.

### Local

Various classifications specific to New Zealand have been produced [@robertson1956].

NIWA have published a useful summary classification at https://niwa.co.nz/climate-and-weather/overview-new-zealands-climate, although it is not accompanied by spatial data. The classification is briefly reproduced in @tbl-ss-climnz below.

```{r}
#| label: tbl-ss-climnz
#| tbl-cap: "NIWA climate regions of New Zealand"

dat_ss_climnz <-
  tribble(~Code, ~Name, ~Locations,
          'NiN', 'Northern New Zealand', 'Auckland, Tauranga, and regions north',
          'NiC', 'Central North Island', 'Hamilton, Taupō, Rotorua',
          'NiW', 'South-West North Island', 'New Plymouth, Wanganui, Palmerston North, Wellington',
          'NiE', 'Eastern North Island', 'Gisborne, Napier, Masterton',
          'SiN', 'Northern South Island', 'Nelson, Blenheim',
          'SiW', 'Western South Island', 'Westport, Hokitika, Milford Sound',
          'SiE', 'Eastern South Island', 'Kaikoura, Christchurch, Timaru',
          'SiI', 'Inland South Island', 'Lake Tekapo, Queenstown, Alexandra, Manapouri',
          'SiM', 'Mountainous South Island', 'Mount Cook and other alpine areas',
          'SiS', 'Southern New Zealand', 'Dunedin, Invercargill',
          'Chi', 'Chatham Islands', 'Chatham Islands',
          'Ant', 'Antarctica', 'Scott Base')

tbl_ss_climnz <- gt(dat_ss_climnz) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10))

tbl_ss_climnz
```

#### Rainfall regime

-   total annual
-   rainfall seasonality (mean/sd off monthly totals)
-   summer/winter ratio (Dec+Jan+Feb)/(June+Jul+Aug)

#### Temperature regime

-   mean annual, etc
-   frostiness

#### Insolation regime

-   day length average and variability

#### Prevailing winds

-   direction, strength by season
-   rationale: aspect effects

## Paleoclimate

???

## Weather

Contemporary weather conditions have not necessarily had a strong influence on soil formation, except perhaps in young and/or rapidly developing soils. These parameters are more relevant for agricultural suitability assessment and for interpreting soil descriptions.

### Current Weather

Record the weather at the time of observation using the codes in @tbl-ss-weatherc, plus temperature in degrees Celsius. These are not always essential, but can affect colour recording and some field tests.

e.g. [**SU 23**]{style="color: red"}

```{r}
#| label: tbl-ss-weatherc
#| tbl-cap: "Current weather conditions, Schoeneberger et al. [-@schoeneberger2012], 1-1"

dat_ss_weatherc <-
  tribble(~Code, ~Name,
          'SU', 'Sunny',
          'PC', 'Partly cloudy',
          'OV', 'Overcast',
          'RA', 'Rain',
          'SL', 'Sleet',
          'SN', 'Snow')

tbl_ss_weatherc <- gt(dat_ss_weatherc) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(20))

tbl_ss_weatherc
```

### Recent rainfall {#sec-ss-recentrain}

Record recent rainfall using the codes in @tbl-ss-weatherp.

```{r}
#| label: tbl-ss-weatherp
#| tbl-cap: "Recent rainfall, adapted from FAO [-@un-fao2006], Table 2"

dat_ss_weatherp <-
  tribble(~Code, ~Name,
          'NM', 'No rain in the last month',
          'NW', 'No rain in the last week',
          'ND', 'No rain in the last 24 hours',
          'RD', 'Rain but no heavy rain in the last 24 hours',
          'RH', 'Heavy rain for some days or excessive rain in the last 24 hours',
          'RE', 'Extremely rainy or snow melting',
          'UK', 'Unknown/uncertain')

tbl_ss_weatherp <- gt(dat_ss_weatherp) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(20))

tbl_ss_weatherp
```
