# Land cover, land use, and land management {#sec-lu}

```{r}
library(dplyr)
library(tibble)
library(gt)
```

Land cover is the physical cover over the soil, both natural and human-made. Land use is the purpose(s) for which the land is used and provides context for why a particular land cover is present. Land management describes specific activities that affect land cover and are driven by land use.

Land cover, use, and management information can be captured as part of soil description when it is relevant to interpreting the landscape, landform, site, and soil profile, and especially where it is likely to have affected the soil classification. Land-use and management practice data are particularly important for long-term monitoring sites, where uses may change significantly between sampling events. Specific recording requirements for routine survey work are listed in @sec-ss-cvusmg.

## Land cover {#sec-lcov}

Land cover in New Zealand is monitored using the Land Cover Database [LCDB, @manaakiwhenualandcareresearchLCDBV60Land2025a]. The database is used to publish a semi-automated map of land cover, derived from satellite imagery, with a minimum unit size of 1 ha. The LCDB categories can also be applied on-ground. 

The categories are listed below, excluding some where it would be physically impossible to observe the soil. The first six (see @tbl-ss-lcdb) are currently mapped to over 80% of New Zealand's land area and so are the ones most likely to be encountered. The full set of category definitions is available in @manaakiwhenualandcareresearchNewZealandLand2025a. 

```{r}
#| label: tbl-ss-lcdb
#| tbl-cap: "Land Cover Database categories: the six most common categories are listed first."

dat_ss_lcdb <-
  tribble(~Code, ~Name, ~Description, 
          40, 'High Producing Exotic Grassland', 'Exotic sward grassland of good pastoral quality and vigour reflecting relatively high soil fertility and intensive grazing management.',
          41, 'Low Producing Grassland', 'Exotic sward grassland and indigenous short tussock grassland of poor pastoral quality reflecting lower soil fertility and extensive grazing management or non-agricultural use.',
          43, 'Tall Tussock Grassland', 'Indigenous snow tussocks in mainly alpine mountain-lands and red tussock in the central North Island and locally in poorly-drained valley floors, terraces and basins of both islands.',
          52, 'Mānuka and/or Kānuka', 'Scrub dominated by mānuka and/or kānuka.',
          69, 'Indigenous Forest', 'Tall forest dominated by indigenous conifer, broadleaved or beech species.',
          71, 'Exotic Forest', 'Planted or naturalised forest.')
          
tbl_ss_lcdb <- gt(dat_ss_lcdb) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color  = 'red', 
                                   weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  # heavier line after first 6 rows:
  tab_style(style = list(cell_borders(sides = 'bottom', weight = px(2))),
            locations = cells_body(rows = 6)) |>
  # italics etc for Description:
  fmt_markdown(columns = 3) |>
  cols_width(Code ~pct(10), Name ~ pct(25)) |>
  tab_source_note(source_note = md("Source: @manaakiwhenualandcareresearchNewZealandLand2025a"))
  
tbl_ss_lcdb
```

## Land use {#sec-luse}

Land use in New Zealand is recorded using the New Zealand Land Use Management classification (NZLUM, [https://nzlum.landcareresearch.co.nz](https://nzlum.landcareresearch.co.nz)), based on the framework developed in @law2024. The classification uses three nested levels of land use with increasing specificity (@fig-ss-nzlum), and allows for recording further information as 'attributes' of the land use (e.g. tenure or management practices).

Land-use classification does not have to be carried out in the field provided adequate supporting information (e.g. notes and photographs) are available for later use. Given that the NZLUM is expected to evolve over time, the observation date can be used to infer the version in use.

![Overview of the NZLUM categories at version 0.4.0 (Source: [https://nzlum.landcareresearch.co.nz](https://nzlum.landcareresearch.co.nz))](data/nzlum_v0.4.0_summary.svg){#fig-ss-nzlum .lightbox fig-alt="A diagram of the New Zealand Land Use Management (NZLUM) system, v 0.4.0" width="95%"}

## Land management {#sec-lman}

A fully developed vocabulary for land management practices does not currently exist for New Zealand. Given the multidimensional and constantly evolving nature of land management practices, it will generally be easier to record such information as free text and use text-mining techniques to extract patterns. Key soil-relevant aspects to record may include (but are not limited to):

- Tillage practices (for arable land)
- Harvest practices (for tree and vine crops)
- Planting practices (for tree and vine crops)
- Crop rotation sequences
- Grazing systems and stocking rates
- Weed and other pest control practices
- Soil amendment and fertilisation practices
- Irrigation schedule
- Periodic drainage or flooding
- Permanent drainage or land surface alterations
- Time since clearing.

In the long term some of these parameters may become formalised as land-use attributes (@sec-luse).
