---
title: "Climate and Weather"
editor: source
editor_options: 
  chunk_output_type: console
---

```{r}
library(dplyr)
library(tibble)
library(gt)
library(gtExtras)
```

## Climate {#sec-clim}

Climate parameters affect soil development directly (e.g. freeze-thaw cycles) and indirectly (e.g. by their influence on vegetation community composition and structure. 

### Global {#sec-clim-global}

The Köppen-Geiger climate classification [@köppen1936] is the most well-known and accepted global system, and has been periodically updated (e.g. [@kottek2006; @beck2018; @beck2023]). Current maps and projections under various climate change scenarios are available from <https://www.gloh2o.org/koppen/>.

In and around New Zealand, the Köppen regions that currently occur are listed in @tbl-ss-koppenz.

```{r}
#| label: tbl-ss-koppenz
#| tbl-cap: "Köppen climate regions of New Zealand and its neighbours"

dat_ss_koppenz <-
  tribble(~Code, ~Name, ~Locations,
          'Af', 'Tropical, rainforest', 'Pacific Islands e.g. Tonga, Samoa',
          'Am' , 'Tropical, monsoon', 'Northwest coast of Fijian islands',
          'Bsk', 'Arid, steppe, cold', 'Central Otago, Melbourne, Northwest Victoria',
          'Cfa', 'Temperate, no dry season, hot summer', 'Coastal New South Wales and southern Queensland',
          'Cfb', 'Temperate, no dry season, warm summer', 'Majority of New Zealand; Chatham Islands; coastal southeastern Australia; Tasmania',
          'Cfc', 'Temperate, no dry season, cold summer', 'Small areas of North Island e.g. upper Ruahines; lower mountains in South Island e.g. Peel Range, Nelson',
          'Dfb', 'Cold, no dry season, warm summer', 'Sheltered pockets in the Southern Alps',
          'Dfc', 'Cold, no dry season, cold summer', 'Lower slopes of major volcanoes and the Southern Alps',
          'ET', 'Polar, tundra', 'Upper slopes of major volcanoes and the Southern Alps',
          'EF', 'Polar, frost', 'Antarctica')

tbl_ss_koppenz <- gt(dat_ss_koppenz) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10))

tbl_ss_koppenz
```

[map goes here]{.diags}

Since most work programs will take place within one zone, regional climate rarely needs to be noted in the field.

Where a study area appears to deviate from a regional map, the extent and nature of the deviation should be discussed in project reporting.

### Regional {#sec-clim-region}

Various climate classifications specific to New Zealand have been produced [@robertson1956; @garr1991; @sturman1999].

NIWA have more recently published a useful [regional summary classification](https://niwa.co.nz/climate-and-weather/overview-new-zealands-climate), although it is not accompanied by spatial data. The classification is briefly reproduced in @tbl-ss-climnz below. As with the global classifications, these don't need to be recorded in the field.

More detailed discussions of each region's climate are published in NIWA's [Regional Climatologies](https://niwa.co.nz/climate-and-weather/regional-climatologies).

```{r}
#| label: tbl-ss-climnz
#| tbl-cap: "NIWA climate regions of New Zealand"

dat_ss_climnz <-
  tribble(~Code, ~Name, ~Locations,
          'NiN', 'Northern New Zealand', 'Auckland, Tauranga, and regions north',
          'NiC', 'Central North Island', 'Hamilton, Taupō, Rotorua',
          'NiW', 'South-West North Island', 'New Plymouth, Wanganui, Palmerston North, Wellington',
          'NiE', 'Eastern North Island', 'Gisborne, Napier, Masterton',
          'SiN', 'Northern South Island', 'Nelson, Blenheim',
          'SiW', 'Western South Island', 'Westport, Hokitika, Milford Sound',
          'SiE', 'Eastern South Island', 'Kaikoura, Christchurch, Timaru',
          'SiI', 'Inland South Island', 'Lake Tekapo, Queenstown, Alexandra, Manapouri',
          'SiM', 'Mountainous South Island', 'Mount Cook and other alpine areas',
          'SiS', 'Southern New Zealand', 'Dunedin, Invercargill',
          'Chi', 'Chatham Islands', 'Chatham Islands',
          'Ant', 'Antarctica', 'Scott Base')

tbl_ss_climnz <- gt(dat_ss_climnz) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10))

tbl_ss_climnz
```

## Specific measurements {#sec-clim-meas}

Some climate parameters are more directly related to soils than others, mainly through their influence on vegetation community and evapotranspiration. Spatial estimates of these parameters are available via the NZ Environmental Data Stack [@mccarthy2021], available from the [MWLR Datastore](https://datastore.landcareresearch.co.nz/gl/dataset/nzenvds). Estimates can also be constructed from local records, where these are available. 

### Rainfall {#sec-clim-rfall}

-   Total annual rainfall: Some soils are unlikely to form outside a particular total annual rainfall range, e.g. Allophanic soils are uncommon where total annual rainfall is less than 1000 mm [@hewitt2010].
-   Rainfall seasonality: Calculated as the mean of monthly rainfall totals, divided by their standard deviation. More pronounced seasonality implies stronger soil wetting and drying cycles, which in turn influences weathering processes.
-   An alternate measurement of seasonality involves calculating the ration of rainfall in the warmest versus the coolest three months. In New Zealand this is usually $(Dec + Jan + Feb)/(June + Jul + Aug)$.
-   Rainfall:evapotranspiration ratio: 
-   Soil moisture deficit

### Temperature {#sec-clim-temp}

-   Mean annual temperature: The mean temperature recorded year-round. 
-   Mean summer temperature: The mean temperature recorded over the warmest quarter of the year. In New Zealand, this is usually December-January-February. Mean summer temperature was a key parameter used by @singers2014 to classify NZ terrestrial ecosystems.
-   frostiness

### Insolation {#sec-clim-insol}

-   day length average
-   day length variability

### Prevailing winds {#sec-clim-wind}

-   direction, strength by season
-   rationale: aspect effects

## Weather {#sec-weath}

Contemporary weather conditions have not necessarily had a strong influence on soil formation, except perhaps in young and/or rapidly developing soils. These parameters are more relevant for agricultural suitability assessment and for interpreting soil profile descriptions.

### Current Weather {#sec-weath-now}

Record the weather at the time of observation using the codes in @tbl-ss-weatherc, plus temperature in degrees Celsius. These are not always essential, but can affect colour recording and some field tests.

e.g. [SU 21]{.ceg} for a perfect day in the field.

```{r}
#| label: tbl-ss-weatherc
#| tbl-cap: "Current weather conditions, Schoeneberger et al. [-@schoeneberger2012], 1-1"

dat_ss_weatherc <-
  tribble(~Code, ~Name,
          'SU', 'Sunny',
          'PC', 'Partly cloudy',
          'OV', 'Overcast',
          'RA', 'Rain',
          'SL', 'Sleet',
          'SN', 'Snow')

tbl_ss_weatherc <- gt(dat_ss_weatherc) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10))

tbl_ss_weatherc
```

### Recent rainfall {#sec-weath-rrain}

Record recent rainfall using the codes in @tbl-ss-weatherp.

```{r}
#| label: tbl-ss-weatherp
#| tbl-cap: "Recent rainfall, adapted from FAO [-@un-fao2006], Table 2"

dat_ss_weatherp <-
  tribble(~Code, ~Name,
          'NM', 'No rain in the last month',
          'NW', 'No rain in the last week',
          'ND', 'No rain in the last 24 hours',
          'RD', 'Rain but no heavy rain in the last 24 hours',
          'RH', 'Heavy rain for some days or excessive rain in the last 24 hours',
          'RE', 'Extremely rainy or snow melting',
          'UK', 'Unknown/uncertain')

tbl_ss_weatherp <- gt(dat_ss_weatherp) |>
   tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(color = 'red', weight = 'bold')),
            locations = cells_body(columns = 1)) |>
  cols_width(Code ~pct(10))

tbl_ss_weatherp
```
